---
title: "Imputation of Longitudinal or Multi-level Data"
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footer.html
    css: ../style.css
    pandoc_args: [
      "--output=EP16_MIlong.html"
      ]
#   learnr::tutorial:
#     progressive: true
#     allow_skip: true
# runtime: shiny_prerendered
---

```{r, include = FALSE}
runimps <- FALSE

if (names(rmarkdown::metadata$output) == "learnr::tutorial") {
  library(learnr)
  learnr::initialize_tutorial()
  static <- FALSE
}

if (names(rmarkdown::metadata$output) == "html_document") {
  knitr::opts_hooks$set(eval = function(opt) {
    if (any(opt$exercise))
      opt$eval <- opt$include <- FALSE
    
    opt
  })
  
  static <- TRUE
  
  options(width = 100)
}
```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
library(JointAI)
library(ggplot2)
library(nlme)
library(lme4)
library(splines)
```


```{r load_data, context="data", include=FALSE}
load("www/pbclong.RData")
source("www/propplot.R")
load("www/impslong.RData")

library(mice)

# pbclong <- subset(survival::pbcseq, select = c('id', 'trt', 'age', 'sex', 'day',
#                                                'ascites', 'hepato', 'bili',
#                                                'albumin'))
# 
# set.seed(2019)
# a <- factor(pbclong$ascites[match(unique(pbclong$id), pbclong$id)])
# a[sample(1:length(a), 60)] <- NA
# 
# b <- survival::pbc$copper[match(unique(pbclong$id), survival::pbc$id)]
# b[sample(1:length(b), 80)] <- NA
# 
# pbclong$ascites <- a[match(pbclong$id, unique(pbclong$id))]
# pbclong$copper <- b[match(pbclong$id, unique(pbclong$id))]
# 
# 
# save(pbclong, file = 'Practicals/MIlong/www/pbclong.RData')
```


## Preface {data-progressive=FALSE}
### R packages

```{r, eval = static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
If any of them are not installed you may be able to follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:')
```
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
    The names of the packages used (and their version numbers available on this platform) are:')
```

* `r R.version.string`
* `mice` (version: `r packageVersion("mice")`)
* `nlme` (version: `r packageVersion("nlme")`)
* `JointAI` (version: `r packageVersion("JointAI")`)
* `splines` (version: `r packageVersion("splines")`)
* `ggplot2` (version: `r packageVersion("ggplot2")`)


### Help files
You can find help files for any function by adding a `?` before the name of the 
function.
```{r, eval = !static, results = 'asis', echo = FALSE}
cat("The files might look a bit funny when they are displayed as R output.")
```

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


### Dataset 

In this practical we will work with data from a trial on primary 
biliary cirrhosis (PBC) of the liver.

```{r, eval = static, echo = FALSE}
asis_output('To get the **pbclong** data, load the file `pbclong.RData`.
            You can download it [here](https://nerler.github.io/EP16_Multiple_Imputation/practical/data/index.html).
            To load this dataset, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the downloaded file on your computer.
If you know the path to the file, you can also use `load("<path>/<pbclong.RData")`.')
```


The variables contained in the dataset `pbclong` are:

```{r pbclongvars, echo = FALSE}
pbclongvars <- rbind(
  c("`id`", "patient identifier"),
  c("`day`", "continuously measured day of follow-up time (the time variable)"),
  c("`sex`", "patients' sex (f: female, m: male)"),
  c("`trt`", "treatment group (0: D-penicillmain, 1: placebo)"),
  c("`age`", "patients' age at intake"),
  c("`ascites`", "presence of ascites at baseline (0:no, 1:yes)"),
  c("`hepato`", "presence of hepatomegaly or enlarged liver"),
  c("`bili`", "serum bilirubin level at baseline"),
  c("`copper`", "urine copper (ug/day)"),
  # c("`chol`", "serum cholesterol level at baseline"),
  c("`albumin`", "serum albumin level at follow-up (time-varying)")
) %>% as.data.frame

names(pbclongvars) <- c("", "")

pbclongvars %>% kable(format = 'html') %>%
  kable_styling()
```

The variables have the following distributions and proportions of missing values:
```{r, echo = FALSE, fig.width = 8, fig.height = 5, out.width = "100%"}
par(mgp = c(2,0.6, 0), mar = c(3,3,2.5,1))
JointAI::plot_all(pbclong, ncol = 4, use_level = TRUE, idvar = 'id')
```


The missing data pattern is:

```{r mdpatlong, echo = FALSE, fig.align = "center", fig.width = 6, fig.height = 4}
md_pattern(pbclong)
```

<br>

The longitudinal outcome `albumin` shows relatively linear trajectories over time:
```{r trajectoryplot, echo = FALSE, fig.width = 8, fig.height = 4.5}
ggplot(pbclong, aes(x = day, y = albumin, color = id, group = id)) +
  geom_line(na.rm = TRUE) +
  theme(legend.position = 'none')
```

To analyse the trajectories of `albumin` we want to use the following linear
mixed effects model with random intercept and slope
(either using `lme`  from the package **nlme** or using `lmer` from the 
package **lme4**):

```{r longmodels, eval = FALSE, echo = TRUE}
# using package nlme
lme(albumin ~ day + sex + trt + age + ascites + bili + copper, random = ~day|id)


# using package lme4
lmer(albumin ~ day + sex + trt + age + ascites + bili + copper + (day|id))
```


## Imputation using  **mice** {.tabset .tabset-fade .tabset-pills}
For imputation of longitudinal data, the **mice** package provides
special imputation methods that take into account the two-level structure of
the data.

In the `pbclong` data, missing values occur in baseline covariates `ascites` and
`copper`, and the time-varying variable `hepato`.

For imputation of baseline covariates, the imputation methods 

* `2lonly.pmm` and 
* `2lonly.norm`

are available.

Time-varying variables can be imputed with imputation methods such as `2l.norm`
or `2l.bin`. Since the imputation of `hepato` with `2l.bin` takes quite some time,
we will omit it in this part of the practical.

The `predictorMatrix` requires some extra specification to identify the 
"id" variable (set to -2) and the random effects structure (set to 2).

```{r, eval = static, echo = FALSE}
asis_output("### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Run the setup imputation and perform the necessary changes in the imputation
method and predictor matrix.

```{r, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#micelonghint">Hint</button>
<div id = "micelonghint" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
**mice** does not recognize automatically that the data are multil-level, hence
the settings chosen by default are not correct.
You need to specify imputation methods for all incomplete variables.
</div>')
```

```{r micelong0, exercise = TRUE}
...

meth_micelong <- ...
pred_micelong <- ...
```


```{r micelong0-hint-1, include = !static, echo = FALSE}
# **mice** does not recognize automatically that the data are multil-level, hence
# the settings chosen by default are not correct.
# You need to specify imputation methods for all incomplete variables.
```



```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution 1\\n")
```

```{r micelong0-solution}
micelong0 <- mice(pbclong, maxit = 0)
meth_micelong <- micelong0$method
pred_micelong <- micelong0$predictorMatrix

# don't impute hepato
meth_micelong[c("hepato")] <- ""
# exclude hepato from predictor of other models (because incomplete)
pred_micelong[, c("hepato")] <- 0

meth_micelong[c("copper", "ascites")] <- "2lonly.pmm"

pred_micelong[, "id"] <- -2
pred_micelong[, "day"] <- 2


# check the imputation method
meth_micelong

# check the predictor matrix
pred_micelong
```


```{r, eval = static, echo = FALSE}
asis_output("### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* run the imputation with the adapted imputation method `meth_micelong` 
and predictor matrix `pred_micelong`
* analyse the imputed data
* pool the results

```{r micelong-setup, include = !static, eval = !static}
micelong0 <- mice(pbclong, maxit = 0)
meth_micelong <- micelong0$method
pred_micelong <- micelong0$predictorMatrix

# don't impute hepato
meth_micelong[c("hepato")] <- ""
# exclude hepato from predictor of other models (because incomplete)
pred_micelong[, c("hepato")] <- 0

meth_micelong[c("copper", "ascites")] <- "2lonly.pmm"

pred_micelong[, "id"] <- -2
pred_micelong[, "day"] <- 2


# check the imputation method
meth_micelong

# check the predictor matrix
pred_micelong
```

```{r, eval = runimps, echo = FALSE}
savedimps_micelong <- mice(pbclong, meth = meth_micelong, pred = pred_micelong,
                 maxit = 20, seed = 2019, printFlag = FALSE)

savedimps_JointAIlong <- lme_imp(albumin ~ day + sex + trt + age + ascites + 
                         bili + copper, random = ~day|id,
                       models = c(copper = 'lognorm', bili = 'glmm_gamma'),
                       no_model = 'day', n.adapt = 100,
                       data = pbclong, n.iter = 500, seed = 2019)


library(splines)
savedimps_JointAIlong2 <- glme_imp(hepato ~ ns(day, df = 3) + sex + trt + age + ascites +
                           bili + copper, random = ~1|id, family = binomial(),
                         models = c(copper = 'lognorm', bili = 'glmm_gamma'),
                         no_model = 'day', data = pbclong, n.iter = 500,
                         seed = 2019)


save(savedimps_micelong,
     savedimps_JointAIlong,
     savedimps_JointAIlong2, file = 'www/impslong.RData')
```

```{r micelong, exercise = TRUE, exercise.timelimit = 200}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution 2\\n")
```

```{r micelong-solution, eval = FALSE}
micelong <- mice(pbclong, meth = meth_micelong, pred = pred_micelong,
                 maxit = 20, seed = 2019, printFlag = FALSE)

library(nlme)
micelong_mira <- with(micelong, lme(albumin ~ day + sex + trt + age + ascites + 
                                      bili + copper, random = ~day|id)
)

# alternative:
# library(lme4)
# micelong_mira <- with(micelong, lmer(albumin ~ day + sex + trt + age + ascites +
#                                        bili + copper + (day|id))
# )

summary(pool(micelong_mira), conf.int = TRUE)
```

```{r, echo = FALSE, eval = static, warning = FALSE}
micelong <- savedimps_micelong

library(nlme)
micelong_mira <- with(micelong, lme(albumin ~ day + sex + trt + age + ascites + 
                                      bili + copper, random = ~day|id)
)

summary(pool(micelong_mira), conf.int = TRUE)
```


## Imputation using **JointAI** {.tabset .tabset-fade .tabset-pills}

To analyse incomplete longitudinal data using a linear mixed model the R package
**JointAI** provides the function `lme_imp()`. 
The specification of the main model components is analogous to the function
`lme()` from the **nlme** package.

**Specification of longitudinal models:**<br>
When imputing variables in a longitudinal (or other multi-level) model and there
are missing values in baseline (level-2) covariates, models need to be specified
for all longitudinal covariates, even if they do not have missing values.
Specifying no model would imply that the incomplete baseline covariates are 
independent of the complete longitudinal variable (see also [here](https://nerler.github.io/JointAI/articles/ModelSpecification.html#why-do-we-need-models-for-completely-observed-longitudinal-covariates)).
Therefore, **JointAI** automatically specifies models for all longitudinal covariates in such a setting.


An exception may be the time variable: it is often reasonable to assume that
the baseline covariates are independent of the measurement times of the outcome
and longitudinal covariates.
To tell **JointAI** not to specify a model for the time variable, the argument
`no_model` can be used.


**Model types for longitudinal covariates:**<br>
For longitudinal covariates the following model types are implemented:
```{r, echo = FALSE}
tab <- rbind(lmm = c("linear mixed model", "continuous variables"),
             glmm_logit = c("logistic mixed model", "factors with two levels"),
             glmm_gamma = c("gamma mixed model (with log-link)", "skewed continuous variables"),
             glmm_poisson = c("poisson mixed model", "count variables"),
             clmm = c("cumulative logit mixed model", "ordered factors with >2 levels")
)

tab <- cbind(paste0("`", rownames(tab), "`"), tab)
colnames(tab) <- c('name', 'model', 'variable type')

knitr::kable(tab, row.names = FALSE, format = 'html') %>%
  kable_styling(full_width = FALSE)
```


**More info:**<br>
For the specification of the other arguments of `lme_imp()`, refer to 

* the [help page](https://nerler.github.io/JointAI/reference/model_imp.html),
* the vignette on 
[Model Specification](https://nerler.github.io/JointAI/articles/ModelSpecification.html#MultiLevelStructure), or
* the details given in the practical on
[Imputation with Non-linear Associations](https://nerler.github.io/EP16_Multiple_Imputation/practical/minonlin/MInonlin.html)

```{r, eval = static, echo = FALSE}
asis_output("### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Run the imputation (start with `n.iter = 500`; this will take a few seconds).

* Remember to specify appropriate models for the incomplete covariates and longitudinal variables.
* Prevent specification of a model for `age`.
* Check convergence using a `traceplot()`.
* If you are satisfied by convergence and mixing of the chains, get the model `summary()`.

```{r runJoitAIlong, exercise = TRUE, fig.width = 9, fig.height = 5}
JointAIlong <- ...
...
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution\\n")
```

```{r runJoitAIlong-solution, eval = FALSE, fig.width = 9, fig.height = 5, message = FALSE}
library(JointAI)
JointAIlong <- lme_imp(albumin ~ day + sex + trt + age + ascites + 
                         bili + copper, random = ~day|id,
                       models = c(copper = 'lognorm', bili = 'glmm_gamma'),
                       no_model = 'day', data = pbclong, n.iter = 500, seed = 2019)

traceplot(JointAIlong)
summary(JointAIlong)
```

```{r, echo = FALSE, eval = static}
JointAIlong <- savedimps_JointAIlong

traceplot(JointAIlong)
summary(JointAIlong)
```


## Additional Exercise JointAI {.tabset .tabset-fade .tabset-pills}
We want to fit a logistic mixed model for the variable `hepato` and explore
if the association is non-linear over time.

```{r, eval = static, echo = FALSE}
asis_output("### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Fit a logistic mixed model using the function `glme_imp` using the same covariates as before.
* Specify a natural cubic spline with 3 degrees of freedom for `day`.
* Check convergence using a `traceplot()`.

```{r, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#JointAIhint1">Hint 1</button>
<div id = "JointAIhint1" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
When specifying a generalized (mixed) model remember to specify the model family and link function.
</div>')

cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#JointAIhint2">Hint 2</button>
<div id = "JointAIhint2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
To use natural cubic splines use the function `ns()` from the package **splines**, 
i.e., `ns(day, df = 3)`.
</div>')
```

```{r JointAIlong2, exercise = TRUE, exercise.timelimit = 500, fig.width = 9, fig.height = 5}
JointAIlong2 <- ...
...
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution 1\\n")
```

```{r JointAIlong2-solution, eval = FALSE}
library(splines)
JointAIlong2 <- glme_imp(hepato ~ ns(day, df = 3) + sex + trt + age + ascites +
                           bili + copper, random = ~1|id, family = binomial(),
                         models = c(copper = 'lognorm', bili = 'glmm_gamma'),
                         no_model = 'day', data = pbclong, n.iter = 1000,
                         seed = 2019)

traceplot(JointAIlong2)
```


```{r, echo = FALSE, eval = static}
JointAIlong2 <- savedimps_JointAIlong2
traceplot(JointAIlong2)
```

```{r, eval = !static, echo=FALSE}
asis_output("### \\n")
```


```{r, eval = static, echo = FALSE}
asis_output("### Task 2\\n")
```
When the model has converged, we want to visualize the potentially non-linear
association of `day`. To do that, we can create a new dataset containing information
on an "average" subject, with different values for `day`.

The function `predDF()` creates such a dataset from an object of class `JointAI`.
It sets reference values (i.e., the median for continuous variables and the 
reference category for categorical variables) for all variables other than the
one specified in the argument `var`. The variable given in `var` will range 
across the range of values of that variable encountered in the data.

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Use `predDF()` to create a dataset that allows visualization of the effect of
`day`.

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution 2\\n")
```

```{r getJointAIlong2, echo = FALSE}
JointAIlong2 <- savedimps_JointAIlong2
```

```{r newdf, exercise = TRUE, exercise.setup = 'getJointAIlong2'}

```

```{r newdf-solution, eval = static}
newdf <- predDF(JointAIlong2, var = 'day')

head(newdf)

```


```{r, eval = !static, echo=FALSE}
asis_output("### \\n")
```


```{r, eval = static, echo = FALSE}
asis_output("### Task 3\\n")
```

We can now predict the outcome of our model for our "average" subject using the
function `predict()`. It takes a `JointAI` object and a `data.frame` containing
the data to predict from as arguments. The argument `quantiles` can be used to
specify which quantiles of the distribution of each fitted value are returned
(default is `2.5%` and `97.5%`).

`predict()` returns a list with the following elements

* `dat`: the `data.frame` provided by the user extended with the fitted values
  and 2.5% and 97.5% quantiles that form the credible interval for the fitted values
* `fit`: a vector containing the fitted values (the mean of the distribution of the fitted value)
* `quantiles`: a matrix containing the credible interval for each fitted value

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Use `predict()` to obtain the fitted values and corresponding intervals
* Visualize the result by plotting fitted values and quantiles (y-axis) over time (`day`; x-axis)

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution 3\\n")
```


```{r makepred-setup, echo = FALSE}
JointAIlong2 <- savedimps_JointAIlong2
newdf <- predDF(JointAIlong2, var = 'day')
```

```{r makepred, exercise = TRUE, fig.width = 7, fig.height = 5}

```

```{r makepred-solution, fig.width = 7, fig.height = 5}
pred <- predict(JointAIlong2, newdata = newdf)

ggplot(pred$dat, aes(x = day, y = fit)) +
  geom_ribbon(aes(ymin = `2.5%`, ymax = `97.5%`), alpha = 0.3) +
  geom_line()
```


```{r, eval = !static, echo=FALSE}
asis_output("### \\n")
```


**Note:**<br>
The fitted values and quantiles are on the scale of the linear predictor,
i.e., obtained by multiplying the data in `newdf` ($\mathbf x$) with the samples
of the posterior distribution of the parameters ($\boldsymbol \beta$).

For a logistic model it is more intuitive to present the fitted values on the
probability scale.

$$
\log\left(\frac{\pi}{1-\pi}\right) = \mathbf x^\top\boldsymbol\beta
\qquad \Rightarrow \pi = \frac{\exp(\mathbf x^\top\boldsymbol\beta)}{1 + \exp(\mathbf x^\top\boldsymbol\beta)}
$$

The function `plogis()` does this transformation.


```{r, echo = FALSE}
JointAIlong2 <- savedimps_JointAIlong2
newdf <- predDF(JointAIlong2, var = 'day')
pred <- predict(JointAIlong2, newdata = newdf)
```

```{r, fig.width = 7, fig.height = 5}
ggplot(pred$dat, aes(x = day, y = plogis(fit))) +
  geom_ribbon(aes(ymin = plogis(`2.5%`), ymax = plogis(`97.5%`)), alpha = 0.3) +
  geom_line() +
  ylab('probability of hepatomegaly')
```
