---
title: "Checks after Multiple Imputation"
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footer.html
    css: ../style.css
    pandoc_args: [
      "--output=EP16_MIcheck.html"
      ]
#   learnr::tutorial:
#     progressive: true
#     allow_skip: true
# runtime: shiny_prerendered
---

```{r, include = F}
if (names(rmarkdown::metadata$output) == "learnr::tutorial") {
  library(learnr)
  learnr::initialize_tutorial()
  static <- FALSE
}

if (names(rmarkdown::metadata$output) == "html_document") {
  knitr::opts_hooks$set(eval = function(opt) {
    if (any(opt$exercise))
      opt$eval <- opt$include <- FALSE
    
    opt
  })
  
  static <- TRUE
  
  options(width = 100)
}
```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
# library(JointAI)
library(reshape2) # for propplot
# library(plyr)
library(RColorBrewer) # for propplot
library(ggplot2)
```


```{r load_data, context="data", include=FALSE}
load("www/NHANES_for_practicals.RData")
NHANES$educ <- as.ordered(NHANES$educ)

source("www/propplot.R")
load("www/imps.RData")
imps <- savedimps_imp
impnaive <- savedimps_impnaive

library(mice)
```


## Preface {data-progressive=FALSE}
### R packages

```{r, eval = static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
If any of them are not installed you may be able to follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:')
```
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
    The names of the packages used (and their version numbers available on this platform) are:')
```

* `r R.version.string`
* `mice` (version: `r packageVersion("mice")`)
* `RColorBrewer` (version: `r packageVersion("RColorBrewer")`)
* `reshape2` (version: `r packageVersion("reshape2")`)
* `ggplot2` (version: `r packageVersion("ggplot2")`)


### Help files
You can find help files for any function by adding a `?` before the name of the 
function.
```{r, eval = !static, results = 'asis', echo = FALSE}
cat("The files might look a bit funny when they are displayed as R output.")
```

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


### Dataset 

For this practical, we will again use the **NHANES** dataset that we have seen in
the previous practical.
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat("In the interactive version of this practical, the data is already loaded
    and the variable `educ` correctly coded as an ordered factor.")
```
```{r, eval = static, echo = FALSE, results = 'asis'}
cat(
'To load this dataset, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the file
`NHANES_for_practicals.RData` on your computer.
If you know the path to the file, you can also use `load("<path>/NHANES_for_practicals.RData")`.

If you have not followed the first practical or if you re-loaded the NHANES data
you need to re-code the variable `educ` again:'
)
```
```{r, eval = FALSE, echo = static}
NHANES$educ <- as.ordered(NHANES$educ)
```


### Imputed data
The imputed data are stored in a `mids` object called `imp` that we created in
the previous practical.
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat("`imp` is already loaded in this interactive version of this practical.")
```
```{r, eval = static, echo = FALSE, results = 'asis'}
cat(
'You can load it into your workspace by clicking the object `imps.RData` 
if you are using RStudio.
Alternatively, you can load this workspace using `load("<path>/imps.RData")`.
You then need to run:')
```
```{r, eval = TRUE, echo = static}
imp <- savedimps_imp
```

The [help file](https://www.rdocumentation.org/packages/mice/versions/2.46.0/topics/mids-class) tells us that a `mids` object is a list with several elements:

```{r, echo = FALSE}
midsdf <- rbind(
  c("`data`:", "Original (incomplete) data set."),
  c("`imp`:",
    paste0("The imputed values: A list of `ncol(data)` components, ",
           "each list component is a matrix with `nmis[j]` rows and `m` columns.")),
  c("`m`:",
    "The number of imputations."),
  c("`where`:", "The missingness indicator matrix."),
 c("`blocks`", "The blocks argument of the `mice()` function."),  
  c("`call`:", "The call that created the `mids` object."),
  c("`nmis`:",
    "The number of missing observations per variable."),
  c("`method`:",
    "The vector imputation methods."),
  c("`predictorMatrix`:",
    "The predictor matrix."),
  c("`visitSequence`:", 
    "The sequence in which columns are visited during imputation."),
 c("`formulas`", "A named list of formulas corresponding the the imputed variables (blocks)."),
  c("`post`:",
    "A vector of strings of length `length(blocks)` with commands for post-processing."),
 # c("`blots`", ""),
  c("`seed`:",
    "The seed value of the solution."),
  c("`iteration`:", 
    "The number of iterations."),
  c("`lastSeedValue`:",
    "The most recent seed value."),
  c("`chainMean:`", 
    paste0("The mean of imputed values per variable and iteration: ",
           "a list of `m` components. Each component is a matrix with `maxit`",
           "columns and `length(visitSequence)` rows.")),
  c("`chainVar`:",
    paste0("The variances of imputed values per variable and iteration" ,
           "(same structure as `chainMean`).")),
  c("`loggedEvents`:",
    paste0("A `data.frame` with the record of automatic corrective actions and warnings; ",
           "(`NULL` if no action was made).")),
  # c("`pad`:",
  #   paste0("A list containing the internally used version of the `data`,",
  #          "`method`, `predictorMatrix`, `visitSequence`, `post` and dummy coding.")
 c("`version`", "Version number of the **mice** package that created the object."),
 c("`date`", "Date at which the object was created.")
) %>% as.data.frame

names(midsdf) <- c(" ", " ")

midsdf %>% kable(format = 'html') %>%
  kable_styling()
  # row_spec(c(4:10, 12:16), background = "#f2f2f2")
```



**Details of the `loggedEvents`:**

`mice()` does some pre-processing of the data:

  * variables containing missing values, that are not imputed but used as
    predictor are removed
  * constant variables are removed
  * collinear variables are removed
  
  
Furthermore, during each iteration

  * variables (or dummy variables) that are linearly dependent are removed
  * `polr` imputation that does not converge is replaced by `polyreg`. 
  
The `data.frame` in `loggedEvents` has the following columns:
```{r, echo = FALSE}
rbind(
  c("`it`", "iteration number"), 
  c("`im`", "imputation number"),
  c("`dep`", "name of the name of the variable being imputed"),
  c("`meth`", "imputation method used"),
  c("`out`", "character vector with names of altered/removed predictors")
) %>%
  kable(format = "html") %>%
  kable_styling(full_width = FALSE)

```

## Evaluate the imputation
### Checking the settings {.tabset .tabset-fade .tabset-pills}

It is good practice to make sure that `mice()` has not done 
any processing of the data that was not planned or that you are not aware of.
This means checking that the correct `method`, `predictorMatrix` and
`visitSequence` were used.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Do these checks for `imp`.

```{r checkmids, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r checkmids-solution, solution = TRUE}
imp$method
imp$predictorMatrix
imp$visitSequence

# you can also try
# identical(imp$method, meth)
# identical(imp$predictorMatrix, pred)
# identical(imp$visitSequence, visSeq)
```

### Logged events {.tabset .tabset-fade .tabset-pills}
Checking the `loggedEvent` shows us if `mice()` detected any problems during
the imputation.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Check the `loggedEvents` for `imp`.

```{r imploggedEvents, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r imploggedEvents-solution, solution = TRUE}
imp$loggedEvents
```

There are no logged events, great!

```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, echo = FALSE}
impnaive <- savedimps_impnaive
```

Let's see what would have happened if we had not prepared the
`predictorMatrix`, `method` and `visitSequence` before imputation.
```{r, eval = !static, echo = FALSE}
asis_output("The object `impnaive` contains the result of")
```
```{r, eval = FALSE, include = !static}
impnaive <- mice(NHANES, m = 5, maxit = 30)
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')

asis_output('Run the imputation without setting any additional arguments:<br>
            `impnaive <- mice(NHANES, m = 5, maxit = 30)`')
```

Take a look at the `loggedEvents` of `impnaive`. 

```{r impnaiveloggedEvents, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r impnaiveloggedEvents-solution, solution = TRUE}
impnaive$loggedEvents
```


```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```


The `loggedEvents` of the "naive" imputation show that the constant variable
`cohort` was excluded before the imputation (as it should be).
Furthermore, in the imputation model for `HyperMed`, the variable `hyptenyes`
was excluded (`hyptenyes` is the dummy variable belonging to `hypten`).

<br>

```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

We did not change the `visitSequence` in `impnaive`. Find out how that
affected the imputed values of `BMI`.

```{r impnaiveVisSeq, exercise = TRUE, exercise.setup = "allimp"}

```

```{r impnaiveVisSeq-hint-1, include = !static, echo = FALSE}
# You can get an imputed datasets from a mids object using the function complete().
# Use ?complete for more information.
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r, impnaiveVisSeq-solution, solution = TRUE}
naiveDF1 <- complete(impnaive, 1)
naivecalcBMI <- with(naiveDF1, wgt/hgt^2)

impDF1 <- complete(imp, 1)
impcalcBMI <- with(impDF1, wgt/hgt^2)

cbind(naiveBMI = naiveDF1$BMI, naivecalcBMI,
      impBMI = impDF1$BMI, impcalcBMI)[which(is.na(NHANES$BMI)), ]

```

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

When we compare the imputed and calculated values of `BMI` from `impnaive` we
can see that the imputed `hgt` and `wgt` give a different `BMI` than is
imputed. This is because `BMI` is imputed before `wgt`, which means that the
most recent imputed value of `wgt` is from the previous iteration.

Changing the `visitSequence` in `imp` prevented this inconsistency.


### Convergence {.tabset .tabset-fade .tabset-pills}
In order to obtain correct results, the MICE algorithm needs to have converged.
This can be checked visually by plotting summaries of the imputed values
accross the iterations.

The mean and variance of the imputed values per iteration and variable are
stored in the elements `chainMean` and `chainVar` of the `mids` object.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Plot them to see if our imputation `imp` has converged.

```{r convergence, exercise = TRUE, exercise.setup = "allimp", fig.width = 9, fig.height = 7}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r convergence-solution, solution = TRUE, fig.width = 8.5, fig.height = 7}
# implemented plotting function (use layout to change the number of rows and columns)
plot(imp, layout = c(6, 6))
```

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

The chains in `imp` seem to have converged, however it is difficult to judge this
based on only 10 iterations. In practice, more iterations should be done.


To save you some time, I ran the imputation again with 30 iterations and the
traceplots confirm convergence:
```{r, echo = FALSE}
imp30 <- savedimps_imp30
impnaive <- savedimps_impnaive
```

```{r, fig.width = 9, fig.height = 7}
plot(imp30, layout = c(6, 6))
```



```{r, eval = static, echo = FALSE}
asis_output("#### Continue\\n")
```

In comparison, `impnaive` had some convergence problems:
```{r, fig.width = 9, fig.height = 7}
plot(impnaive, layout = c(6, 6))
```


`hgt`, and `wgt` show a clear trend and the chains do not mix well, i.e., there
is more variation between the chains than within each chain.
(the same is the case for `BMI`).

These are clear signs that there is correlation or identification
problems between these variables and some other variables (which is why we made
adjustments to the `predictorMatrix` for `imp`).

### Imputed values {.tabset .tabset-fade .tabset-pills}
Now that we know that `imp` has converged, we can compare the distribution of
the imputed values against the distribution of the observed values.
When our imputation models fit the data well, they should have similar
distributions (conditional on the covariates used in the imputation model).

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1 \\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Plot the distributions of the imputed variables (continuous and categorical).
* Make sure the imputed values are realistic (e.g., height of 2.50m or weight of 10kg for adults).

```{r, echo = FALSE, eval = !static}
asis_output("**Note:** The function `propplot()` is already loaded. To use this
            function on your own computer you can download it from
            [https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab](https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab)")
```

```{r distrplot, exercise = TRUE, exercise.setup = "allimp", fig.width = 9, fig.height = 7}

```

```{r impdistrhint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#impdistrhint">Hint 1</button>
<div id = "impdistrhint" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
You can use `densityplot()` and `propplot()` to get plots for all continuous and
categorical variables.

`propplot()` is not part of any package. Copy the following syntax that defines
this function:

<script src="https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab.js"></script>

</div>')
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#impdistrhint2">Hint 2</button>
<div id = "impdistrhint2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
To check all imputed values you can either get a summary of the `imp` element
of the `mids` object or create a complete dataset containing all imputations
using the function `complete()` and get the summary of that.
</div>')
```

```{r distrplot-hint-1, include = !static, echo = FALSE}
# You can use "densityplot()" and "propplot()" to get plots for all continuous and
# categorical variables.
# 
# The function "propplot() is not part of any package but can be found online:
# https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab
# 
# In this practical it is available automatically.
# 
# To check all imputed values you can either get a summary of the "imp" element
# of the mids object or create a complete dataset containing all imputations
# using the function "complete()" and get the summary of that.
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```


```{r distrplot-solution, solution = TRUE, fig.width = 9, fig.height = 7, cache = TRUE}
# plot densities of continuous variables
densityplot(imp)


# plot for all categorical variables
propplot(imp)

# get the summary of the "the" imputed values
sapply(Filter(function(x) nrow(x) > 0, imp$imp),
       function(x) summary(unlist(x))
)
```

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

Unfortunately, we have some negative imputed values for `bili`. Often, this
would not result in bias in the analysis, but may be difficult to explain
when providing a summary of the imputed data in a publication.
In the present example we can see that the observed values have a slightly
right-skewed distribution compared to the imputed values. Re-doing the
imputation with `pmm` instead of `norm` for `bili` should fix this.
(However, since the imputations seem fine overall, and there is little knowledge
gain in re-doing the previous steps, we will skip this repetition 
in this practical.)

The distributions of the imputed values for `hgt` and `SBP` differ a bit
from the distributions of the observed data.

We also imputed a larger proportion than might have been expected in the 
highest category of `alc`, and the distribution of values for `smoke`
looks a bit weird (but `smoke` only has one missing value, which makes it difficult
to judge the distribution of the imputed values).

```{r, eval = static, echo = FALSE}
asis_output("#### Task 2 \\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Investigate if differences in the distributions of observed and imputed values,
can be explained by other variables. Check this for

* `SBP` conditional on `gender` and `hypten`
* `hgt` conditional on `gender`
* `alc` conditional on `gender` or `smoke`


```{r distrplot2, exercise = TRUE, exercise.setup = "allimp", fig.width = 9, fig.height = 7}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```


```{r distrplot2-solution, solution = TRUE, fig.width = 8, fig.height = 6, cache = TRUE}
densityplot(imp, ~SBP|hypten + gender)
densityplot(imp, ~hgt|gender)

propplot(imp, alc ~ gender + smoke)
```


