---
title: "Getting to Know Your Incomplete Data"
output:
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footer.html
    css: ../style.css
    pandoc_args: [
      "--output=EP16_IncompleteData.html"
      ]
#   learnr::tutorial:
#     progressive: true
#     allow_skip: true
# runtime: shiny_prerendered
---

```{r, include = F}
if (names(rmarkdown::metadata$output) == "learnr::tutorial") {
  library(learnr)
  learnr::initialize_tutorial()
  static <- FALSE
}

if (names(rmarkdown::metadata$output) == "html_document") {
  knitr::opts_hooks$set(eval = function(opt) {
    if (any(opt$exercise))
      opt$eval <- opt$include <- FALSE
    
    opt
  })
  
  static <- TRUE
  
  options(width = 100)
}

library(knitr)
library(VIM)
library(visdat)
library(mice)
library(JointAI)
library(plyr)
library(corrplot)
```


```{r load_data, context="data", include=FALSE}
load("www/NHANES_for_practicals.RData")
```

## Preface {data-progressive=FALSE}
### R packages

```{r, eval = static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
If any of them are not installed you may be able to follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:')
```
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
    The names of the packages used (and their version numbers available on this platform) are:')
```

* `r R.version.string`
* `mice` (version: `r packageVersion("mice")`)
* `visdat` (version: `r packageVersion("visdat")`) 
* `JointAI` (version: `r packageVersion("JointAI")`)
* `VIM` (version: `r packageVersion("VIM")`)
* `plyr` (version: `r packageVersion("plyr")`)
* `corrplot` (version: `r packageVersion("corrplot")`)
* `ggplot2` (version: `r packageVersion("ggplot2")`)


### Help files
You can find help files for any function by adding a `?` before the name of the 
function.
```{r, eval = !static, results = 'asis', echo = FALSE}
cat("The files might look a bit funny when they are displayed as R output.")
```

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


## Dataset 
### Overview {.tabset .tabset-fade .tabset-pills}

For this example, we will use the **NHANES** dataset.
```{r, eval = static, echo = FALSE, results = 'asis'}
cat(
'To get the **NHANES** data, load the file `NHANES_for_practicals.RData`.
You can download it [here](https://nerler.github.io/EP16_Multiple_Imputation/practical/data/index.html). To load this dataset into R, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the file on your computer.
If you know the path to the file, you can also use `load("<path>/NHANES_for_practicals.RData")`.'
)
```
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat("In the interactive version
of this practical, the data is already loaded. ")
```


```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Let's take a first look at the data. Useful functions are
`dim()`, `head()`, `str()` and `summary()`.

```{r load-mice, exercise = TRUE}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r load-mice-solution, solution = TRUE}
dim(NHANES)
head(NHANES)
str(NHANES)
summary(NHANES)
```


### Variable coding {.tabset .tabset-fade .tabset-pills}
It is important to check that all variables are coded correctly, i.e., have
the correct `class`. Imputation software (e.g., the **mice** package) uses the
`class` to automatically select imputation methods.
When importing data from other software it can happen that factors become
continuous variables or that ordered factors lose their ordering.

`str()` showed that in the NHANES data `smoke` and `alc` are correctly specified
as ordinal variables, but `educ` is an unordered factor.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Using `levels(NHANES$educ)` we can print the names of the categories of `educ`.
Convert the unordered factor to an ordered factor, for example using `as.ordered()`.
Afterwards, check if the conversion was successful.
```{r ordered_educ, exercise = TRUE}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r ordered_educ-solution, solution = TRUE}
levels(NHANES$educ)

NHANES$educ <- as.ordered(NHANES$educ)

str(NHANES$educ)
```

```{r educrecode, echo = F}
NHANES$educ <- as.ordered(NHANES$educ)
```

## Distribution of missing values
### Missing data pattern {.tabset .tabset-fade .tabset-pills}
In the `summary()` we could already see that there are missing values in several
variables. The missing data pattern can be obtained and visualized by several
functions from different packages.
Examples are 

* `md.pattern()` from package **mice**
* `md_pattern()` from package **JointAI** (with argument `patter = TRUE`)
* `aggr()` from package **VIM**
* `vis_dat()` and `vis_miss()` from package **visdat**


```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

```{r, eval = static, echo = FALSE}
asis_output("Explore the missing data pattern of the NHANES data.")
```


```{r md-pattern, exercise = TRUE, exercise.setup = "educrecode", fig.height = 10, fig.width = 7}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution (i)\\n")
```

```{r md-pattern-solution, solution = TRUE, fig.width = 7, fig.height = 10, include = !static}
mdp <- mice::md.pattern(NHANES)
mdp

# or
MDP <- JointAI::md_pattern(NHANES, pattern = TRUE)
MDP

# or
VIM::aggr(NHANES)
```

```{r md-pattern-solution-mice, solution = TRUE, fig.width = 7, fig.height = 10, include = static}
mdp <- mice::md.pattern(NHANES)
mdp
```


```{r, eval = !static, echo=FALSE}
asis_output("### \\n")
```

`md.pattern()` from package **mice** gives us a matrix where each row represents
one missing data pattern (1 = observed, 0 = missing).
The rownames show how many rows of the dataset have the given pattern.
The last column shows the number of missing values in each
pattern, the last row the number of missing values per variable.

`md.pattern()` also plots the missing data pattern automatically.

```{r, eval = static, echo = FALSE}
asis_output("#### Solution (ii)\\n")
```

```{r md-pattern-solution-JointAI, fig.width = 7, fig.height = 10, include = static}
MDP <- JointAI::md_pattern(NHANES, pattern = TRUE)
MDP
```

The function `md_pattern()` from package **JointAI** gives a matrix very
similar to the one obtained from `mice::md.pattern()`.
However, here, the number of rows in the data that have a particular missing
data pattern are given in the last column.

For more information on how to customize the visualization by `md_pattern()`
see the vignette [Visualizing Incomplete Data](https://nerler.github.io/JointAI/articles/VisualizingIncompleteData.html)
on the [webpage of **JointAI**](https://nerler.github.io/JointAI/index.html).

```{r, eval = static, echo = FALSE}
asis_output("#### Solution (iii)\\n")
```

```{r md-pattern-solution-VIM, fig.width = 8, fig.height = 5, include = static}
VIM::aggr(NHANES)
```

`aggr()` from **VIM** plots a histogram of the proportion of missing values
per column as well as a visualization of the missing data pattern. Here,
a small histogram on the right edge of the plot shows the proportion of cases
in each pattern.

For more options how to customize the plot, see the 
[**VIM** documentation](https://CRAN.R-project.org/package=VIM).

### Proportion of missing values {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Now get an overview of

* how much missingness there is in each variable, and
* the proportion of (in)complete cases.

```{r misprop, exercise = TRUE}

```

```{r misprop-hint-1, eval = !static, include = !static}
# The function complete.cases() will give you a vector that is TRUE if the
# the case is complete and FALSE if there are missing values.

# is.na() returns TRUE if the value is missing, FALSE if the value is observed
# colSums() calculates the sum of values in each column of a data.frame or matrix
# colMeans() calculates the mean of values in each column of a data.frame or matrix
```

```{r mispropr_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#misprop">Hint</button>
<div id = "misprop" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
The function `complete.cases()` will give you a vector that is TRUE if the
the case is complete and FALSE if there are missing values.

`is.na()` returns `TRUE` if the value is missing, `FALSE` if the value is observed

`colSums()` calculates the sum of values in each column of a data.frame or matrix

`colMeans()` calculates the mean of values in each column of a data.frame or matrix
</div>')
```


```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r misprop-hint-2, solution = TRUE}
# number and proportion of complete cases
cbind(
  "#" = table(ifelse(complete.cases(NHANES), 'complete', 'incomplete')),
  "%" = round(100 * table(!complete.cases(NHANES))/nrow(NHANES), 2)
)

# number and proportion of missing values per variable
cbind("# NA" = sort(colSums(is.na(NHANES))),
      "% NA" = round(sort(colMeans(is.na(NHANES))) * 100, 2))

```

### Relationship between variables {.tabset .tabset-fade .tabset-pills}
In the missing data pattern we could already see that some variables tend to 
be missing together. But there may be different types of relationships between
variables that are of interest.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


Our data contains `hgt` (height), `wgt` (weight) and `BMI`. 
Check the missing data pattern specifically for those three variables.

```{r heightweight, exercise = TRUE}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r heightweight-solution, solution = TRUE}
# Three solutions to choose from:
JointAI::md_pattern(NHANES[, c("hgt", "wgt", "BMI")], pattern = TRUE, plot = FALSE)

with(NHANES, 
     table(ifelse(is.na(hgt), 'height mis.', 'height obs.'),
           ifelse(is.na(wgt), 'weight mis.', 'weight obs.'),
           ifelse(is.na(BMI), 'BMI mis.', 'BMI obs.'))
)

plyr::ddply(NHANES, c(height = "ifelse(is.na(hgt), 'missing', 'observed')",
                      weight = "ifelse(is.na(wgt), 'missing', 'observed')",
                      BMI = "ifelse(is.na(BMI), 'missing', 'observed')"),
            plyr::summarize,
            N = length(hgt)
)
```

As we have already seen in the lecture, there are some cases where
only either `hgt` or `wgt` is missing. `BMI` is only observed when both
components are observed. To use all available information, we want to 
impute `hgt` and `wgt` separately and calculate `BMI` from the imputed values.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

The data contains variables on hypertension (`hypten`) and the use of medication
to treat hypertension (`HyperMed`).
We might expect that medication is only prescribed for patients with hypertension, hence, we need to investigate the relationship between `HyperMed` and `hypten`.

* Make a table of the two variables to confirm our expectation.
* Make sure that missing values are also displayed in that table!

```{r hyptentab, exercise = TRUE}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r hyptentab-solution, solution = TRUE}
with(NHANES, table(HyperMed, hypten, exclude = NULL))
```

```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Furthermore, we can expect a systematic relationship between the variables
`hypchol` (hypercholesterolemia) and `chol` (serum cholesterol).
Find out how these two variables are related.

```{r hypchol, exercise = TRUE}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```


```{r hypchol-solution, solution = TRUE}
with(NHANES, plot(chol ~ hypchol))

with(NHANES, summary(chol[hypchol == "no"]))
with(NHANES, summary(chol[hypchol == "yes"]))
```

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

It seems that `hypchol` is defined as `chol > 6.2`, which makes `hypchol`
dependent on `chol`. 


```{r, eval = static, echo = FALSE}
asis_output("#### Task 4\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Visualize the correlations between variables to detect associations that you
may have missed.

```{r corrplot, exercise = TRUE}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 4\\n")
```


```{r corrplot-solution, solution = TRUE, fig.width = 9, fig.height = 9, warning=FALSE}
# convert all variables to numeric
dat_num <- sapply(NHANES, as.numeric)
cormat <- cor(dat_num, use = 'pair', method = 'spearman')
corrplot::corrplot(cormat, method = 'square', type = 'lower')
```

**Note:** Correlations involving categorical variables are not usually done!
We only use this as a quick-and-dirty method for visualization of relationships.

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

The question marks in the plot indicate that no correlation
could be calculated for `cohort` (because `cohort` is the same for all subjects)
and neither between `HyperMed` and `hypten` (but we already knew that).

The plot shows that besides the relations we were already aware of, `wgt` is
strongly correlated with `WC`.
Comparing the number of cases where only either `wgt` or `WC` is missing
shows that there are 14 cases where `wgt` is missing but `WC` is observed
and 58 cases where `WC` is missing and `wgt` is observed.

```{r}
with(NHANES, table(ifelse(is.na(WC), 'WC mis.', 'WC obs.'),
                   ifelse(is.na(wgt), 'wgt mis.', 'wgt obs.')
))
```


## Distribution of the data {.tabset .tabset-fade .tabset-pills}
Before imputing missing values it is important to take a look at how the 
observed parts of incomplete variables are distributed, so that we can choose
appropriate imputation models.

```{r, eval = static, echo = FALSE}
asis_output("### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Visualize the distributions of the incomplete continuous and categorical 
variables. The package **JointAI** provides the convenient function `plot_all()`,


Pay attention to

* whether continuous distributions deviate considerably from the normal distribution,
* if variables have values close to the limits of the range they are defined in,
* whether categorical variables are very unbalanced (i.e., some category very small).


```{r distr, exercise = TRUE, fig.width = 9, fig.height = 8, exercise.setup = "educrecode"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution\\n")
```

```{r distr-solution, solution = TRUE, fig.width = 9, fig.height = 10.5, dpi=600}
par(mar = c(3, 3, 2, 1), mgp = c(2, 0.6, 0))
JointAI::plot_all(NHANES, nclass = 30, ncol = 4)
```

To learn more about additional options of `plot_all()` check out the vignette [Visualizing Incomplete Data](https://nerler.github.io/JointAI/articles/VisualizingIncompleteData.html)
on the [webpage of **JointAI**](https://nerler.github.io/JointAI/index.html).
