---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "5. Know Your Data"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: aspectratio=169
---

```{r setup, include = FALSE}
projdir <- gsub("/Slides", "", getwd())


knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    # } else {
    #   knitr::asis_output("\\vspace*{-1.22222ex}")
    }
  }
)


knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%")

options(width = 120)

load(file.path(projdir, "Slides/data", "NHANES.RData"))
NHANES <- subset(NHANES, select = -DBP)
NHANES <- NHANES[, c(which(names(NHANES) != 'BMI')[1:7],
                     which(names(NHANES) == 'BMI'),
                     which(names(NHANES) != 'BMI')[-c(1:7)])]
```



## Missing Data Pattern
To demonstrate the \blue{work flow} when performing \blue{multiple imputation}
with the \textbf{mice} package, we use data from the
\href{https://www.cdc.gov/nchs/nhanes/index.htm}{National Health and Nutrition
Examination Survey (NHANES)}.


\vfill

There are several packages in \includegraphics[width = 1em]{TemplateGraphics/Rlogo.png}
that provide functions to investigate
\blue{the missing data pattern}.

\bigskip

Examples are:\newline
\textbf{mice}, \textbf{JointAI}, \textbf{VIM}, \textbf{Amelia}, \textbf{visdat}, \textbf{naniar}, \ldots


## Missing Data Pattern
\scriptsize
```{r}
mdp <- mice::md.pattern(NHANES, plot = FALSE)
head(mdp[, -c(7:14)]) # omit some columns to fit it on the slide
tail(mdp[, -c(7:14)])
```

## Missing Data Pattern
\scriptsize
```{r JointAI_md_pattern, fig.width = 9, fig.height = 8, fig.show = 'hide'}
JointAI::md_pattern(NHANES)
```

\begin{center}
\vspace*{-1.8ex}
\includegraphics[width=0.6\linewidth]{05_Know_Your_Data_files/figure-beamer/JointAI_md_pattern-1}
\vspace*{-1.5ex}
\end{center}

## Missing Data Pattern
\scriptsize
```{r VIM_aggr, fig.width = 10, fig.height = 5, out.width = "95%"}
VIM::aggr(NHANES, prop = TRUE)
```


## Missing Data Pattern
We are also interested in the number and proportion of (in)complete cases ...
\small
```{r NCC_calc, size = 'small'}
cctab <- table(complete.cases(NHANES))
cbind(
  "#" = setNames(cctab, c('incomplete', 'complete')),
  "%" = round(100 * cctab/nrow(NHANES), 2)
)
```


## Missing Data Pattern
... and the proportion of missing values per variable:
\small
```{r nmis, eval = FALSE}
cbind("# NA" = sort(colSums(is.na(NHANES))),
      "% NA" = round(sort(colMeans(is.na(NHANES))) * 100, 2))
```

\begincols[t, onlytextwidth]
\begincol{0.29\linewidth}
\small
```{r nmis_print, echo = FALSE}
nmis <- cbind("# NA" = sort(colSums(is.na(NHANES))),
              "% NA" = round(sort(colMeans(is.na(NHANES))) * 100, 2))
nmis[1:ceiling(nrow(nmis)/3), ]
```
\endcol
\begincol{0.3\linewidth}
```{r nmis_print2, echo = FALSE}
nmis[(1 + ceiling(nrow(nmis)/3)):ceiling(2 * nrow(nmis)/3), ]
```
\endcol
\begincol{0.35\linewidth}
```{r nmis_print3, echo = FALSE}
nmis[(1 + ceiling(2 * nrow(nmis)/3)):nrow(nmis), ]
```
\endcol
\endcols

## Missing Data Pattern
See also:

\begincols[t, onlytextwidth]
\begincol{0.48\linewidth}

* `mice::md.pattern()`
* `visdat::vis_miss()`
* `visdat::vis_dat()`
* \ldots

\endcol
\begincol{0.48\linewidth}

* `naniar::prop_miss_case()`,\newline
  `naniar::pct_miss_case()`
* `naniar::prop_complete_case()`,\newline`naniar::pct_complete_case()`
* `naniar::miss_var_summary()`
* `mice::md.pairs()`
* \ldots

\endcol
\endcols

## Data Distribution
\small
```{r eval = FALSE}
JointAI::plot_all(NHANES[, -1])  # exclude 1st column to fit on slide
```

\vspace*{-2.5ex}
```{r NHANESplotall, fig.width = 9, fig.height = 4.4, out.width = "100%", echo = FALSE}
par(mgp = c(2, 0.6, 0), mar = c(2, 3, 2.5, 0.5))
JointAI::plot_all(NHANES[, -1], breaks = 30, ncol = 5)
```
\vspace*{-1.5ex}


```{r, echo = FALSE}
options(width = 70)
```


## Correlations & Patterns

A quick (and dirty) way to check for strong correlations between variables is:
\small
```{r corr}
# re-code all variables as numeric and calculate spearman correlation
Corr <- cor(sapply(NHANES, as.numeric),
           use = "pairwise.complete.obs", method = "spearman")
```

```{r, echo = FALSE}
options(width = 120)
```

```{r corrplot syntax, eval = FALSE}
corrplot::corrplot(Corr, method = "square", type = "upper",
                   tl.col = "black")
```
\normalsize
\blue{Note:} \parbox[t]{0.9\linewidth}{
We only use the correlation coefficient for categorical variables for\\
visualization, not as a statistical result!
}

## Correlations & Patterns
```{r corrplotplot, echo = FALSE, out.width = "78%", crop = TRUE}
# corrplot::corrplot.mixed(Corr, upper = 'square', lower = 'number',
#                          lower.col = 'black', tl.cex = 0.6,
#                          number.cex = 0.5, bg = 'transparent')
corrplot::corrplot(Corr, method = "square", type = "upper", tl.col = "black")
```
\vspace*{-2ex}

## Correlations & Patterns
Check out what the problem is with \Rstring{hypertension} and \Rstring{HyperMed}:
\small
```{r tab_hyper}
table(hypertension = NHANES$hypten,
      HyperMed = NHANES$HyperMed, exclude = NULL)
```


## Why are values missing?
Knowing your data also means being able to answer these questions:
\begin{itemize}
\item Do missing values in multiple variables always \blue{occur together}?\\
      (e.g. blood measurements)
\item Are there \blue{structural missing values}? (e.g. pregnancy status in men)
\item Are there \blue{patterns} in the missing values?\\
      (e.g. only patients with hypertension have observations of \Rstring{HyperMed})
\item Are values \blue{missing by design}?
\item Is the \blue{assumption of ignorable missingness} (MAR or MCAR) justifiable?
\end{itemize}


## Auxiliary Variables
\blue{Auxiliary variables} are variables that are not part of the analysis but
\blue{can help during imputation}.

\bigskip

Good auxiliary variables
\begin{itemize}
\item are \blue{related to the probability of missingness} in a variable, or
\item are \blue{related to the incomplete variable} itself,
\item do \blue{not have many missing values} themselves and
\item are (mostly) \blue{observed} when the incomplete variable of interest is missing.
\end{itemize}
