---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "6. Imputation with mice"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: [aspectratio=169]
bibliography: references.bib
---

```{r setup, include = FALSE}
projdir <- gsub("/Slides", "", getwd())

runimps <- FALSE
  
knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)


knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%")

options(width = 120)

suppressPackageStartupMessages(library("mice"))

load(file.path(projdir, "Slides/data", "NHANES.RData"))
NHANES <- subset(NHANES, select = -DBP)
NHANES <- NHANES[, c(which(names(NHANES) != 'BMI')[1:7],
                     which(names(NHANES) == 'BMI'),
                     which(names(NHANES) != 'BMI')[-c(1:7)])]
```


## Main Function Arguments

The main function for imputation with the R package **mice** is `mice()`{.R}.\
Its most important arguments are:

>- \Rarg{data}: the dataset
>- \Rarg{m}: number of imputed datasets
>- \Rarg{maxit}: number of iterations
>- \Rarg{method}: vector of imputation methods
>- \Rarg{defaultMethod}: vector of default imputation methods
>- \Rarg{predictorMatrix}: matrix specifying roles of variables


## Iterations and Imputations

By default, **mice** will use

* \Rarg{maxit} ` = 5` (i.e., the algorithm is run for 5 iterations)
* \Rarg{m} ` = 5` (i.e., 5 imputed datasets will be created)

Often, these values need to be larger, but we will come back to this
later in the course.

## Imputation Methods

\textbf{mice} has implemented many \blue{imputation methods}, the
most commonly used ones are:

>- \Rstring{pmm}: predictive mean matching (any)
>- \Rstring{norm}: Bayesian linear regression (numeric)
>- \Rstring{logreg}: binary logistic regression (binary)
>- \Rstring{polr}: proportional odds model (ordered factors)
>- \Rstring{polyreg}: polytomous logistic regression (unordered factors)

\bigskip

\onslide<6->{
Imputation methods are chosen \blue{automatically based on the type of variable}.


\blue{\ding{225} }\parbox[t]{0.80\linewidth}{
Make sure all \blue{variables are coded correctly}, so that the
automatically chosen imputation methods are appropriate.
}
}

## Imputation Methods: \Rarg{defaultMethod}

The default setting for the argument \Rarg{defaultMethod} is:

`defaultMethod = c("pmm", "logreg", "polyreg", "polr")`{.R}

i.e.,\vspace*{-1.5ex}

 * \Rstring{pmm} for numerical variables
 * \Rstring{logreg} for binary variables
 * \Rstring{polr} for unordered factors with $>2$ levels
 * \Rstring{polyreg} for ordered factors

\pause
\bigskip

**Example:**\
To use \Rstring{norm} instead of \Rstring{pmm} for all
continuous incomplete variables:
\small
```{r mice example, eval = FALSE, size = 'small'}
mice(NHANES, defaultMethod = c("norm", "logreg", "polyreg", "polr"))
```
\normalsize


## Tip: Use a Setup Run

To obtain a default version for the arguments like \Rarg{method} and
\Rarg{predictorMatrix} it is convenient to \blue{do
a setup run} of `mice()`{.R} without iterations:

\small
```{r imp0}
library("mice")
imp0 <- mice(NHANES, maxit = 0)
```
\normalsize

\pause

The function `mice()`{.R} returns an object of class \blue{mids}
(multiply imputed data set).

\pause\bigskip

The default versions can then be extracted and modified, for example:

\small
```{r}
meth <- imp0$method
```
\normalsize

## Imputation Methods: \Rarg{method}
```{r echo = FALSE}
op <- getOption("width")
options(width = 70)
```

\small
```{r meth, eval = FALSE}
meth
```
\vspace*{-3ex}
```{r, ref.label="meth", eval = TRUE, echo = FALSE}
```
\normalsize
\pause

Variables that do \blue{not need to be imputed} (= do not have any missing values)
are set to \blue{""}.

To change the imputation method for single variables, 
the vector `meth` can then be adapted:
\small
```{r imp0orig1}
meth["albu"] <- "norm"
```
\normalsize
```{r echo = FALSE}
options(width = op)
```

## Imputation Methods: \Rarg{method}
For variables that \blue{should not be imputed} the imputation method can be
set to \blue{""}:
\small
```{r imp0orig2}
meth["HyperMed"] <- ""
```
\normalsize

For the actual imputation, `mice()`{.R} is called using the adapted argument(s):
```{r, eval = FALSE}
imp1 <- mice(NHANES, method = meth)
```


## Predictor Matrix
The \Rarg{predictorMatrix} is a matrix that specifies \blue{which variables are used as
predictors} in which imputation model.\newline

Each row represents the model for the variable given in the row name.

\small
```{r predmat}
head(imp0$predictorMatrix)[, 1:11]  # subset, to fit on the slide
```
\normalsize

Variables \blue{not used as predictor} are (or have to be) set to \blue{zero}.\
By default, \blue{all variables} (except the variable itself) \blue{are used}
as predictors.


## Predictor Matrix
\blue{Important:}\
A variable that has \blue{missing values needs to be imputed} in order to be used as a
predictor for other imputation models!!!
\pause

**Example:**\

* We set `meth["HyperMed"] = ""`{.R}\
  \ding{225} \Rstring{HyperMed} will not be imputed
* \Rstring{HyperMed} is still used as predictor variable (default setting of \Rarg{predicorMatrix})
* for cases with missing \Rstring{HyperMed} none of the other variables can be imputed

\bigskip
\pause

\blue{\ding{225}} We also have to set
```{r}
pred <- imp0$predictorMatrix
pred[, 'HyperMed'] <- 0
```

<!-- \blue{\ding{225}} Make sure to adjust the \Rarg{predictorMatrix} and \Rarg{method} -->
<!-- to avoid using ID variables or other columns of the data that should not be part -->
<!-- of the imputation. -->


## A Quick Summary
\small
```{r impset}
library("mice")
# setup-run
imp0 <- mice(NHANES, maxit = 0,
             defaultMethod = c("norm", "logreg", "polyreg", "polr"))

# adjust imputation methods
meth <- imp0$method
meth["HyperMed"] <- ""

# adjust predictor matrix
pred <- imp0$predictorMatrix
pred[, "HyperMed"] <- 0
```
```{r imp, eval = FALSE}
# run imputation with adjusted settings
imp1 <- mice(NHANES, method = meth, predictorMatrix = pred)
```

```{r imprun, eval = runimps, ref.label = 'imp', include = FALSE}
```
```{r, echo = FALSE, eval = runimps}
save(imp1, file = file.path(projdir, "Slides/workspaces/imp1.RData"))
```


```{r impget, echo = FALSE}
load(file.path(projdir, "Slides/workspaces/imp1.RData"))
```
\normalsize


## Passive Imputation
In some cases, variables are \blue{functions of other variables}:
$BMI = \frac{weight}{height^2}$\bigskip

If we impute \Rstring{BMI} directly, its values may be \blue{inconsistent} with the
(imputed) values of \Rstring{height} and \Rstring{weight}.

\vspace*{1.5ex}\pause

\small
```{r demo_passive1, results='hide'}
DF1 <- complete(imp1, 1) # select the first imputed dataset

round(cbind(BMI = DF1$BMI,                         # imputed
            "wgt/hgt^2" = DF1$weight/DF1$height^2  # calculated
), 2)[which(is.na(NHANES$BMI))[1:3], ]
```
\normalsize \pause
\vspace*{-1.5ex}
\begincols[t, onlytextwidth]
\begincol{0.35\linewidth}
\small
```{r demo_pasive1_print, ref.label = "demo_passive1", echo = FALSE, results = 'markup'}
```
\normalsize
\endcol
\begincol{0.6\linewidth}
The imputed values of \Rstring{BMI} are impossible given the corresponding
values of \Rstring{height} and \Rstring{weight}.
\endcol
\endcols


## Passive Imputation
Moreover, if some components of a variable are observed we want to use that
\blue{information to reduce uncertainty}.

\small\vspace*{1.5ex}
```{r demo_passive2, eval = FALSE}
table(weight_missing = is.na(NHANES$weight),
      height_missing = is.na(NHANES$height))
```
\vspace*{-2ex}
```{r, eval = TRUE, ref.label = "demo_passive2", echo = FALSE}
```
\normalsize
Here we have
`r sum(is.na(NHANES$height) & !is.na(NHANES$weight))` +
`r sum(is.na(NHANES$weight) & !is.na(NHANES$height))` =
`r sum(xor(is.na(NHANES$height), is.na(NHANES$weight)))` cases in which
either \Rstring{height} or \Rstring{weight} is observed.\bigskip

\pause

\blue{\ding{225} }\parbox[t]{0.80\linewidth}{
We would like to impute \Rstring{height} and \Rstring{weight} separately and
calculate \Rstring{BMI} from the (imputed) values of the two variables.
}



## Passive Imputation
If \Rstring{BMI} is not a relevant predictor in any of the other imputation models,
we could just exclude BMI from the imputation and \blue{re-calculate it afterwards}.

\smallskip

To use \Rstring{BMI} as predictor in the imputation, it has to be \blue{calculated
in each iteration} of the algorithm.\
In \textbf{mice} this is possible with \blue{\textit{passive imputation}}:

\smallskip

\pause
Instead of using a standard imputation \Rarg{method}, we can specify a 
\blue{formula to calculate \Rstring{BMI}}:
\small
```{r passive_BMI1}
meth["BMI"] <- "~I(weight/height^2)"
```
\normalsize\pause
To \blue{prevent feedback} from \Rstring{BMI} in the imputation of \Rstring{height}
and \Rstring{weight} the \Rarg{predictorMatrix} needs to be modified:
\small
```{r passive_BMI2}
pred[c("weight", "height"), "BMI"] <- 0
```
\normalsize


## Passive Imputation
Since \Rstring{BMI} depends on \Rstring{weight}, and the two variables are highly correlated
(
$\rho = `r round(cor(NHANES$weight, NHANES$BMI, use = "pair"), 2)`$
)
it \blue{\ul{may}} be beneficial \blue{not to use them simultaneously} as predictors in the
other imputation models.

Which one to use may differ between imputation models.

\bigskip\pause

\blue{Passive imputation} can also be useful in settings where

>- imputation models include \blue{interaction terms} between incomplete variables
  (see @Buuren2012, p. 133 for an example), or when
>- a number of covariates is used to form a \blue{sum score}.\newline
  The sum score, instead of all single elements, can then be used as predictor
  in other imputation models.


## Post Processing
`mice()`{.R} has an argument \Rarg{post} that can be used to specify functions
that modify imputed values.\bigskip

\vfill
\pause

\textbf{Example:}\
When inspecting the imputed values from \Robj{imp}, we find that some imputed
values in \Rstring{creat} are negative.

\small
```{r demo_squeeze}
# DF1 is the first imputed dataset we extracted earlier
summary(DF1$creat)
```
\normalsize

```{r interactiveHistogram, eval = FALSE, fig.width = 7, fig.height = 3.8, echo = FALSE}
library(ggplot2)
plotDF <- reshape2::melt(cbind(orig = NHANES$creat,
                     imp = DF1$creat))

ggplot(plotDF, aes(x = value, fill = Var2, color = Var2)) +
  geom_histogram(alpha = 0.3, bins = 200, position = "identity", na.rm = TRUE) +
  theme_light() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 10)) +
  scale_fill_brewer(name = "", palette = "Dark2") +
  scale_color_brewer(name = "", palette = "Dark2") +
  scale_x_continuous(name = "Creatinine level", seq(0, 10, 1))
```


## Post Processing
A helpful function:

* `mice::squeeze()`{.R} to censor variables at given boundaries

With the following syntax all imputed values of \Rstring{creat} that are
outside the interval \texttt{c(0, 100)} will be \blue{set to those limiting values}.
\small
```{r}
post <- imp1$post
post["creat"] <- "imp[[j]][,i] <- squeeze(imp[[j]][,i], c(0, 100))"
```
\pause
```{r imp2, eval = FALSE}
imp2 <- update(imp1, post = post)
```
```{r, eval = runimps, include = FALSE}
imp2 <- update(imp1, post = post, seed = 2020)
save(imp2, file = file.path(projdir, "Slides/workspaces/imp2.RData"))
```
```{r imp2get, echo = FALSE}
load(file.path(projdir, "Slides/workspaces/imp2.RData"))
```
\normalsize

\vfill\pause

\blue{Note:}\
When many observations are outside the limits it may be better to \blue{change the
imputation model} since the implied \blue{assumption of the imputation model}
apparently \blue{does not fit the complete data distribution}.


## Post Processing
The \blue{post-processing} functionality allows for many
\blue{more data manipulations} and is not restricted to \Rfct{squeeze}.\bigskip

Any string of \textsf{R} commands provided will be evaluated after the
corresponding variable is imputed, within each iteration.\bigskip

\pause

**For example,** if subjects with SBP $>140$ should be classified as
hypertensive:
\small
```{r eval = FALSE}
post["hypten"] <- "imp[[j]][data[where[, j], 'SBP'] > 140, i] <- 'yes'"
```
\normalsize
\pause

This also allows for (some) \blue{MNAR scenarios}, for example, by multiplying
or adding a constant to the imputed values, or to re-impute values depending
on their current value.


## Visit Sequence
When the \blue{post-processed or passively imputed values} of a variable depend on
other variables, the \blue{sequence in which the variables are imputed} may be
important to obtain \blue{consistent values}.

In \Rfct{mice} the argument \Rarg{visitSequence} specifies in which order the
columns of the \Robj{data} are imputed.

\bigskip\pause

\blue{Example:}\
If \Rstring{BMI} is passively imputed (calculated) before the new
imputations for \Rstring{height} and \Rstring{weight} are drawn, the resulting
values of \Rstring{BMI}, will match \Rstring{height} and \Rstring{weight} from
the \blue{previous iteration}, but not the iteration given in the imputed dataset.

## Visit Sequence
```{r echo = FALSE}
options(width = 70)
```

**Example:**\
Currently, \Rstring{BMI} is imputed before \Rstring{height} and \Rstring{weight}:

\small
```{r visitSeq1, size = 'small'}
visitSeq <- imp2$visitSequence
visitSeq
```
\normalsize

To get consistent values, we need to change the \Rarg{visitSequence}:

\small
```{r visitSeq2}
visitSeq <- c(visitSeq[-which(visitSeq == "BMI")], # everything else
              "BMI")                               # BMI 
```
\normalsize


## Visit Sequence

* By default \Rfct{mice} imputes in the \blue{order of the columns in the \Robj{data}}.

\medskip


* The \Rarg{visitSequence} may specify that a column is \blue{visited multiple times}
  during one iteration.

\medskip

* All incomplete variables must be \blue{visited at least once}.



## Automated Changes \label{subsec:goodtoknow}

\Rfct{mice} automatically performs some \blue{pre-processing} and \blue{removes}
\begin{itemize}
\item incomplete variables that are not imputed but are specified as predictors,
\item constant variables, and
\item collinear variables.
\end{itemize}

\bigskip\pause

In each iteration
\begin{itemize}
\item linearly dependent variables are removed and
\item \Rstring{polr} imputation models that do not converge are replaced by
      \Rstring{polyreg}.
\end{itemize}

\bigskip

\textbf{Why?}\
To avoid problems in the imputation models.


## Automated Changes
As a \blue{consequence}
\begin{itemize}
\item imputation models may differ from what the user has specified or assumes
      is happening, or
\item variables that should be imputed are not.
\end{itemize}

\bigskip\pause

\begin{block}{}
\begin{itemize}
\item[\blue{\ding{225}}] Know your data
\item[\blue{\ding{225}}] Make sure \Rarg{method} and \Rarg{predictorMatrix} are
      specified appropriately
\item[\blue{\ding{225}}] Check the output and log of these automatic actions carefully
\end{itemize}
\end{block}


## A note

>``Please realize that these choices are always needed. Imputation software
needs to make default choices. These choices are intended to be useful across
a wide range of applications. However, the \blue{default choices are not necessarily
the best for the data at hand. There is simply no magical setting that always
works}, so often some tailoring is needed.'' [@Buuren2012, \ p.124]

## References
