---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "7. Convergence & Diagnostics"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: [aspectratio=169]
bibliography: references.bib
---

```{r setup, include = FALSE, purl = FALSE}
projdir <- gsub("/Slides", "", getwd())

runimps <- FALSE
  
knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)

knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%",
                      opts.label = 'nopurl', purl = FALSE)

knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_template$set(nopurl = list(purl = FALSE))
knitr::opts_template$set(dopurl = list(purl = TRUE))


options(width = 120)

suppressPackageStartupMessages(library(mice))

load(file.path(projdir, "Slides/data", "NHANES.RData"))
NHANES <- subset(NHANES, select = -DBP)
NHANES <- NHANES[, c(which(names(NHANES) != 'BMI')[1:7],
                     which(names(NHANES) == 'BMI'),
                     which(names(NHANES) != 'BMI')[-c(1:7)])]


load(file.path(projdir, "Slides/workspaces/imp3.RData"))
load(file.path(projdir, "Slides/workspaces/imp4.RData"))
```


## Setting
In this section, we use imputed data from the following set-up:
\small
```{r, opts.label = "dopurl"}
library("mice")
imp0 <- mice(NHANES, maxit = 0,
             defaultMethod = c("norm", "logreg", "polyreg", "polr"))

meth <- imp0$method
meth["HyperMed"] <- ""
meth["BMI"] <- "~I(weight/height^2)"

pred <- imp0$predictorMatrix
pred[, "HyperMed"] <- 0

post <- imp0$post
post["creat"] <- "imp[[j]][,i] <- squeeze(imp[[j]][,i], c(0, 100))"
```

## Setting
Knowing that we "forgot" to change the predictor matrix to prevent feedback from 
\Rstring{BMI} to \Rstring{height} and \Rstring{weight}, we use the resulting
`mids` object \Robj{imp3} for demonstratin purposes:
\small
```{r, eval = FALSE}
imp3 <- mice(NHANES, method = meth, predictorMatrix = pred, post = post)
```
```{r, eval = runimps, ref.label="imp3", echo = FALSE}
```
```{r, echo = FALSE, eval = runimps}
save(imp3, file = file.path(projdir, "Slides/workspaces/imp3.RData"))
```
\normalsize

\pause\bigskip

Additionally, we work with the improved imputation using the following additional
settings:
\small
```{r, opts.label = "dopurl"}
pred[c("weight", "height"), "BMI"] <- 0
```
```{r imp4, eval = FALSE}
imp4 <- mice(NHANES, method = meth, predictorMatrix = pred, post = post,
             maxit = 30)
```
```{r, eval = runimps, ref.label="imp4", echo = FALSE}
```
```{r, echo = FALSE, eval = runimps}
save(imp4, file = file.path(projdir, "Slides/workspaces/imp4.RData"))
```
\normalsize


## Logged Events

Information on the automatic changes that were done by **mice** is returned as
`loggedEvents`, which is part of the `mids` object.

`loggedEvents` is a `data.frame` and has the following columns:

\begin{tabular}{lp{20em}}
\Rstring{it} & iteration number\\
\Rstring{im} & imputation number\\
\Rstring{dep} & dependent variable\\
\Rstring{meth} & imputation method used\\
\Rstring{out} & names of altered or removed predictors
\end{tabular}

\bigskip

It can be obtained as
\small
```{r, eval = FALSE}
imp3$loggedEvents
```
\normalsize

## Logged Events
Neither `imp3` nor `imp4` had any logged events.

\medskip

To demonstrate `loggedEvents` we create a small dataset with some
"mistakes" in it:
\small
```{r demoimp}
demo <- NHANES[, 1:5]                  # first 5 variables from NHANES
demo$dupl <- demo[, 4]                 # create a duplicate variable
demo$const <- 1                        # create a constant variable
demo$age[demo$gender == 'male'] <- NA  # set age missing for all males
```
```{r, eval = FALSE}
demoimp <- mice(demo)
```
\vspace*{-2ex}
```{r, echo = FALSE}
demoimp <- mice(demo, printFlag = FALSE, maxit = 2, m = 3, seed = 123)
```


## Logged Events
\small
```{r, eval = FALSE}
head(demoimp$loggedEvents)
```
\normalsize
\begincols[onlytextwidth, T]
\begincol{0.5\linewidth}
\small
```{r, echo = FALSE}
head(demoimp$loggedEvents)
```
\endcol
\begincol{0.48\linewidth}
Before imputation (iteration 0):

* the constant variable was removed
* the duplicate variable was identified as collinear and removed.

During imputation:

* gender was removed from the model for age

\endcol
\endcols

## Convergence

From a previous section of this course we know that
\textbf{mice} uses an \blue{iterative algorithm} and imputations from the first few
iterations may not be samples from the ``correct'' distributions.

\bigskip\pause

\blue{Traceplots} can be used to visually assess \blue{convergence}.\bigskip


In \textbf{mice}, the
function \Rfct{plot} produces traceplots of the  mean and standard deviation
(across subjects) per incomplete variable.


## Convergence
\small
```{r convergence_imp3, eval = FALSE}
plot(imp3, layout = c(8, 4))
```
\vspace*{-5ex}
```{r, fig.width = 12, fig.height = 6, fig.keep = "first", eval = TRUE, ref.label="convergence_imp3", echo = FALSE, out.width = "103%"}
```
\vspace*{-2ex}
\normalsize


## Convergence
\small
```{r convergence_imp4, eval = FALSE}
plot(imp4, layout = c(8, 4))
```
\vspace*{-5ex}
```{r, fig.width = 12, fig.height = 6, fig.keep = "first", eval = TRUE, ref.label="convergence_imp4", echo = FALSE, out.width = "103%"}
```
\vspace*{-2ex}
\normalsize

## Convergence

\blue{Strong trends} and traces that show \blue{correlation} between variables
indicate \blue{problems of feedback}.
This needs to be investigated and resolved in the specification of
the \Rarg{predictorMatrix}.\bigskip

\blue{Weak trends} may be artefacts that often disappear when the imputation is
performed with more iterations.


## Diagnostics
When MCMC chains have converged, the \blue{distributions of the imputed and observed
values} can be compared to investigate differences between observed and imputed
data.

\bigskip

\blue{Note:}\
Plots usually show the \blue{marginal} distributions of observed and imputed values,
which do not have do be identical under MAR.

\bigskip

\blue{But:}\
The \blue{conditional} distributions
(given all the other variables in the imputation model) of the imputed values
are assumed to be the same as the conditional distributions of the observed data.



## Diagnostics
\textbf{mice} provides several functions for visual diagnosis of imputed values:
\begin{itemize}
\item \Rfct{densityplot} (for large datasets and variables with many NAs)
\item \Rfct{stripplot} (for smaller datasets and/or variables with few NAs)
\item \Rfct{bwplot}
\item \Rfct{xyplot}
\end{itemize}

\bigskip

These functions create \href{https://stat.ethz.ch/R-manual/R-devel/library/lattice/html/Lattice.html}{lattice graphics},
which can be modified analogously to their parent functions from the
\textbf{lattice} package.


## Diagnostics
\small
```{r densityplot1, eval = FALSE}
densityplot(imp4)
```
\normalsize\vspace*{-2ex}
```{r, fig.width = 9, fig.height = 4.5, echo = FALSE, eval = TRUE, ref.label='densityplot1'}
```
\vspace*{-2ex}

## Diagnostics
\begincols[onlytextwidth]
\begincol{0.5\linewidth}
```{r denszoom, eval = FALSE}
densityplot(imp4, ~ height + creat,
            layout = c(1, 2))
```

\bigskip

The `densityplot()`{.R} shows that 

* imputed values of \Rstring{height} are smaller than the observed values
* the distribution of the imputed values of \Rstring{creat} is wider than the 
  distribution of the observed values

\endcol
\begincol{0.5\linewidth}
```{r, ref.label="denszoom", eval = TRUE, fig.width = 5, fig.height = 5.5, echo = FALSE}
```
\endcol
\endcols


## Diagnostics
\begincols[onlytextwidth]
\begincol{0.5\linewidth}
In some cases \blue{differences} in distributions \blue{can be explained by 
strata} in the data, however, here, \Rstring{gender} does not explain the
difference in observed and imputed values.
```{r densityplot2, eval = FALSE}
densityplot(imp4, ~height|gender,
            layout = c(1, 2))
```
\endcol
\begincol{0.5\linewidth}
\small\vspace*{-2ex}
```{r, fig.width = 5, fig.height = 5.5, eval = TRUE, echo = FALSE, out.width = "102%", ref.label = 'densityplot2'}
```
\endcol
\endcols

```{r, echo = FALSE}
op <- getOption("width")
options(width = 60)

wrap.simpleError <- function(x, options) {
  # x is an error object, with components "call" and "message".  Ignore
  # the call, but wrap the result like code:
  paste0("```\n## Error: ", x$message, "\n```")
}
```

## Diagnostics
As an alternative, we might consider \Rstring{race} to explain the differences
\small
```{r densityplot fail, fig.width = 9, fig.height = 5, eval = FALSE}
densityplot(imp4, ~height|race)
```
\vspace*{-3ex}
```{r, ref.label="densityplot fail", error = TRUE, eval = TRUE, echo = FALSE}
```
\normalsize
\pause
```{r densityplot fail2, eval = FALSE}
with(NHANES, table(race = race, "height missing" = is.na(height)))
```
\begincols[onlytextwidth]
\begincol{0.6\linewidth}
```{r, ref.label="densityplot fail2", echo = FALSE, eval = TRUE}
```
\endcol
\begincol{0.4\linewidth}
There are not enough missing values of \Rstring{height} per
categories of \Rstring{race} to estimate densities.
\endcol
\endcols

## Diagnostics
\begincols[onlytextwidth]
\begincol{0.58\linewidth}
In that case, a \Rfct{stripplot} may be better suited.

\medskip

\small
```{r stripplot1, eval = FALSE}
stripplot(imp4, height ~ race|gender,
          pch = c(1, 20), layout = c(1, 2),
          scales = list(x = list(rot = 45)))
```
\endcol
\begincol{0.4\linewidth}
\vspace*{-3ex}
```{r, ref.label='stripplot1', crop = TRUE, eval = TRUE, echo = FALSE, fig.width = 4.5, fig.height = 6, out.width = "108%"}

```
\endcol
\endcols

## Diagnostics
Alternatively, observed and imputed data can be represented by box-and-whisker plots:
\vspace*{2ex}
\small
```{r bwplot, eval = FALSE}
bwplot(imp4, height + weight + bili + chol ~.imp)
```
\vspace*{-3ex}
```{r, ref.label="bwplot", eval = TRUE, fig.width = 10, fig.height = 4, echo = FALSE}
```


## Diagnostics
The function \Rfct{xyplot} allows multivariate investigation of the imputed
versus observed values.
\vspace*{2ex}
\small
```{r xyplot, eval = FALSE}
xyplot(imp4, height ~ weight | gender, pch = c(1, 20))
```
\vspace*{-3ex}
```{r, fig.width = 10, fig.height = 4.8, out.width = "95%", eval = TRUE, echo = FALSE, ref.label="xyplot"}
```

## Diagnostics
All of the above graphs displayed only continuous imputed variables.


For categorical variables we can compare the proportion of values in each
category.

\bigskip

\textbf{mice} does not provide a function to do this, but we can write one
ourselves, as for instance the function \Rfct{propplot}, for which the syntax
can be found here: \href{https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab}{https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab}

```{r load propplot, message = F, echo = F}
source(file.path(projdir, "Slides", "Rfcts", "propplot.R"))
```


## Diagnostics
\small
```{r propplot, eval = FALSE}
propplot(imp4)
```
\vspace*{-2ex}
```{r, echo = FALSE, warning = FALSE, fig.width = 10, fig.height = 4.5}
propplot(imp4, strip.text = element_text(size = 14),
         legend.position = c(0.83, 0.2))

```

## Diagnostics
\Rstring{smoke} and \Rstring{educ} have \blue{very few missing values}
(`r sum(is.na(NHANES$smoke))` and  `r sum(is.na(NHANES$educ))`)\
\blue{\ding{225}} no need to worry about differences

\bigskip\pause

\begincols[onlytextwidth]
\begincol{0.5\linewidth}
\Rstring{alc}: missing values are imputed in the category "\<=1"
more often than we would expect from the observed data

\bigskip

\onslide<3->{
If we expect that \Rstring{gender} and \Rstring{race} might explain the differences
for \Rstring{alc}, we can include those factors into the plot.
}
\endcol
\begincol{0.5\linewidth}
```{r, fig.width = 5, fig.height = 4.5, echo = FALSE}
propplot(imp4, alc ~ 1, legend.position = 'none', strip.text = element_text(size = 14))
```
\endcol
\endcols


## Diagnostics
\small
```{r propplot_alc_print, eval = FALSE}
propplot(imp4, formula = alc ~ race + gender)
```
\vspace*{-2ex}
```{r propplot_alc, fig.width = 10, fig.height = 5.2, echo = FALSE}
propplot(imp4, formula = alc ~ race + gender, nrow = 2, ncol = 5,
         strip.text = element_text(size = 12), legend.position = 'bottom')
``` 
\vspace*{-2ex}

## Diagnostics
\begincols[onlytextwidth]
\begincol{0.5\linewidth}
We also see that

* \Rstring{hypten} is less frequent and
* \Rstring{hypchol} a bit more frequent, in the imputed data compared to the observed.

\endcol
\begincol{0.5\linewidth}
```{r, fig.width = 5, fig.height = 5.2, echo = FALSE}
propplot(imp4, hypten + hypchol ~ 1, ncol = 1, nrow = 2,
         legend.position = 'none', strip.text = element_text(size = 14))
```
\endcol
\endcols


## Diagnostics
Since hypertension is more common in older individuals, we may want to
investigate if \Rstring{age} can explain the differences in imputed values of
\Rstring{hypten}.
\small
```{r hypten_age_table, eval = FALSE}
round(sapply(split(NHANES[, "age"], addNA(NHANES$hypten)), summary), 1)
```

\begincols[onlytextwidth]
\begincol{0.5\linewidth}
\small
```{r, eval = TRUE, ref.label= "hypten_age_table", echo = FALSE}
```
\normalsize
\endcol
\begincol{0.5\linewidth}
The distribution of \Rstring{age} in participants with missing
\Rstring{hypten} is very similar to the distribution of \Rstring{age} 
in participants without \Rstring{hypten}.
\endcol
\endcols


## Diagnostics
Plotting the proportions of observed and imputed \Rstring{hypten} separately
per quartile of \Rstring{age}:
\small
```{r propplot_hypten_print, fig.width = 10, fig.height = 4.2, eval = FALSE}
propplot(imp4, formula = hypten ~ cut(age, quantile(age), include.lowest = T))
```

```{r propplot_hypten, fig.width = 10, fig.height = 4.2, echo = FALSE}
propplot(imp4, formula = hypten ~ cut(age, quantile(age), include.lowest = T),
         strip.text = element_text(size = 12), nrow = 1, legend.position = 'bottom')
```
\normalsize


