---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "8. Analysis of Imputed Data, Pooling & More"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: [aspectratio=169]
bibliography: references.bib
---

```{r setup, include = FALSE}
projdir <- gsub("/Slides", "", getwd())

runimps <- FALSE
  
knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)


knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%")

options(width = 120)

suppressPackageStartupMessages(library(mice))

load(file.path(projdir, "Slides/data", "NHANES.RData"))
NHANES <- subset(NHANES, select = -DBP)
NHANES <- NHANES[, c(which(names(NHANES) != 'BMI')[1:7],
                     which(names(NHANES) == 'BMI'),
                     which(names(NHANES) != 'BMI')[-c(1:7)])]

source(file.path(projdir, 'Slides/07_Convergence_and_Diagnostics.R'))

load(file.path(projdir, "Slides/workspaces/imp3.RData"))
load(file.path(projdir, "Slides/workspaces/imp4.RData"))
load(file.path(projdir, "Slides/workspaces/imp4b.RData"))
```



## Analysing Imputed Data
Once we have confirmed that our imputation was successful, we can move on
to the \blue{analysis of the imputed data}.\bigskip

For example a logistic regression model:
\small
```{r fitglm, eval = FALSE}
glm(DM ~ age + gender + hypchol + BMI + smoke + alc, family = binomial())
```
\normalsize
\pause

To fit the model on each of the imputed datasets:

* no need to extract the data from the \Robj{mids} object
* instead: can use \Rfct{with}

\small
```{r glm}
mod1 <- with(imp4, glm(DM ~ age + gender + hypchol + BMI + smoke + alc,
                       family = binomial()))
```

## Analysing Imputed Data
\Robj{mod1} is an object of class \Robj{mira} (Multiply Imputed Repeated Analyses).

\pause\bigskip

The `mira` object has elements

* \blue{call:} the current call
* \blue{call1:} call that created the `mids` object
* \blue{nmis:} vector with number of missing values per variable
* \blue{analyses:} a list of the separate analyses


## Pooling the Results
Pooled results can be obtained using \Rfct{pool} and its summary.

\scriptsize
```{r, eval = FALSE}
pool(mod1)$pooled
```
```{r, echo = FALSE}
print(pool(mod1)$pooled, digits = 3)
```

<!-- ## Pooling the Results -->

<!-- The function `pool()`{.R} returns the elements necessary for the pooling of the results: -->

<!-- * `estimate`: pooled estimate ($\bar Q$) -->
<!-- * `ubar`: within-imputation variance ($\bar U$) -->
<!-- * `b`: between-imputation variance ($B$) -->
<!-- * `t`: total variance ($T$) -->
<!-- * `dfcom`: degrees of freedom in the complete data ($N - \#params$) -->
<!-- * `df`: degrees of freedom of the $t$-statistic ($\nu$) -->
<!-- * `riv`: Relative increase in variance ($r_m$) -->
<!-- * `lambda`: proportion attributable to the missingness $\left(\frac{B + B/m}{T}\right)$ -->
<!-- * `fmi`: fraction of missing information -->


## Pooling the Results
\footnotesize
```{r pooled, eval = FALSE}
summary(pool(mod1), conf.int = TRUE)
```
```{r, echo = FALSE}
print(summary(pool(mod1), conf.int = TRUE), digits = 3)
```



## Pooling the Results

\blue{Pooling} with \Rfct{mice::pool} is available for most types of models.\bigskip

It extracts the model coefficients and variance-covariance matrices using \Rfct{tidy}
from the package \textbf{broom}.
Hence, pooling using the \Rfct{pool} function from \textbf{mice} only works for
models of classes for which a method \Rfct{tidy} exists.

\bigskip

An alternative is offered by the package \textbf{mitools} and the function
\Rfct{MIcombine}.



## Functions for Pooled Results
\textbf{mice} provides some functions for evaluating model fit or model comparison.

\bigskip \pause

`pool.r.squared()`{.R} calculates the pooled (adjusted) $R^2$:
\small
```{r poolrsquared, size = 'small'}
mod2 <- with(imp4, lm(SBP ~ DM + age + hypten))
pool.r.squared(mod2, adjusted = TRUE)
```
\normalsize

The argument \Rarg{adjusted} specifies whether the adjusted $R^2$ or the
standard $R^2$ is returned.


## Functions for Pooled Results

To compare nested models:

* `D1()`{.R}: multivariate Wald test
* `D3()`{.R}: likelihood-ratio test statistic

\bigskip

To pool test statistics when no variance-covariance matrix is available:

* `D2()`{.R}: Combining test statistics

\bigskip\vfill

For details, see @Buuren2012, [Section 5.3](https://stefvanbuuren.name/fimd/sec-multiparameter.html#sec:wald)
and @Schafer1997.



## Functions for Pooled Results
\textbf{Example:}
To test if \Rstring{smoke} has a relevant contribution to the model for
\Rstring{DM} from above we re-fit the model without \Rstring{smoke} and
compare the two models:
\small
```{r poolcompare, size = 'small'}
mod3 <- with(imp4, glm(DM ~ age + gender + hypchol + BMI + alc,
                       family = "binomial"))
# likelihood ratio test
D3(mod1, mod3)
```
\normalsize

\Rfct{anova} allows comparison of multiple nested models



## Functions for Pooled Results
The package \textbf{miceadds} extends \textbf{mice}, for example with the
following functionality:

\bigskip

\blue{Combine $\chi^2$ or F statistics from multiply imputed data:}
\small
```{r miceaddsfcts, eval = FALSE, size = 'small'}
miceadds::micombine.chisquare(dk, df, ...)
miceadds::micombine.F(values, df1, ...)
```
\normalsize

\bigskip\pause

\blue{Calculate correlation or covariance of imputed data:}
\small
```{r miceaddsfcts2, eval = FALSE}
miceadds::micombine.cor(mi.res, ...)
miceadds::micombine.cov(mi.res, ...)
```


## Functions for Pooled Results
\small
```{r}
# fit chi-square test on each imputed dataset
chisq_mira <- with(imp4, chisq.test(table(educ, smoke)))
```
\pause
```{r}
# extract degrees of freedom
dfs <- chisq_mira$analyses[[1]]$parameter

# extract test statistic
stat <- sapply(chisq_mira$analyses, "[[",  'statistic')
```
\pause
```{r}
# pool the tests
miceadds::micombine.chisquare(dk = stat, df = dfs)
```


## Functions for Pooled Results
\small
```{r, eval = FALSE}
miceadds::micombine.cor(imp4, variables = c('weight', 'BMI', 'creat'))
```
\scriptsize
```{r, echo = FALSE}
print(miceadds::micombine.cor(imp4, variables = c('weight', 'BMI', 'creat')), digits = 2)
```



## Extract Imputed Data
The function \Rfct{complete} allows us to \blue{extract the imputed data} from a \Robj{mids} object:
\small
```{r completedata, eval = FALSE}
mice::complete(data, action = 1, include = FALSE)
```
\normalsize

\begin{itemize}
\item \Robj{data}: the \Robj{mids} object
\item \Rarg{action}:
      \begin{itemize}
      \item \Rval{1}, \ldots, \Rval{m} (single imputed dataset)
      \item \Rstring{"long"}: long format (imputed data stacked vertically)
      \item \Rstring{"broad"}: \parbox[t]{0.75\linewidth}{
                                  wide format (imputed data combined horizontally;\\
                                  ordered by imputation)}
      \item \Rstring{"repeated"}: (like \Rstring{"broad"}, but ordered by variable)
      \end{itemize}
\item \Rarg{include}: include the original data?\\
      (if \Rarg{action} is \Rstring{"long"}, \Rstring{"broad"} or \Rstring{"repeated"})
\end{itemize}



## Combining \texttt{mids} objects
To \blue{increase the number of imputed datasets} without re-doing the initial $m$ imputations,
a second set of imputations can be done and the two \Robj{mids} objects combined
using \Rfct{ibind}.

\small
```{r ibind, eval = runimps}
# same syntax as before, but different seed
imp4b <- update(imp4, seed = 456)
```

```{r ibindrun, eval = runimps, echo = FALSE}
save(imp4b, file = file.path(projdir, "Slides", "workspaces/imp4b.RData"))
```

```{r}
imp4combi <- ibind(imp4, imp4b) # combine
```

```{r}
# check the new number of imputed datasets:
imp4combi$m
```


## Adding variables to \texttt{mids} objects

The function \Rfct{cbind.mids} allows us to \blue{add columns} to a \Robj{mids} object.
The extra columns can either be a \Robj{data.frame}, \Robj{matrix}, \Robj{vector}
or \Robj{factor} or another \Robj{mids} object.\bigskip


For example data columns that should be part of the imputed data for completeness,
but are not needed in the imputation.

\small
```{r cibindmids, eval = FALSE}
# "otherdata" is a data.frame
impextra <- mice:::cbind.mids(x = imp4, y = otherdata)
```

\blue{Note:}
\Rfct{cbind} just adds columns to the data, you need to make sure they are
\blue{sorted correctly} so that the rows of the new data are from the same subjects
as the corresponding rows in the imputed data.



## References

