---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "9. Imputation in Complex Settings"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: [aspectratio=169]
bibliography: references.bib
csl: taylor-and-francis-apa.csl
---

```{r setup, include = FALSE}
projdir <- gsub("/Slides", "", getwd())

library(knitr)
library(kableExtra)
library(plyr)

source(file.path(projdir, 'Slides/Rfcts/Section09_plots.R'))
```

## Quadratic Effect
Consider the case where the \blue{analysis model} (which we assume to be true)
is
$$y = \beta_0 + \beta_1 x + {\color{EMCdark}\mathbf{\boldsymbol{\beta_2 x^2}}} + \ldots,$$
i.e., $y$ has a \blue{quadratic relationship} with $x$, and $x$ is incomplete.

\begin{columns}[onlytextwidth]
\begin{column}{0.6\linewidth}
\begin{center}
\includegraphics[width = \linewidth]{graphics/Section09/p_qdr1.pdf}
\end{center}
\end{column}
\begin{column}{0.4\linewidth}
The original data show a curved pattern.
\end{column}
\end{columns}

## Quadratic Effect
The model used to \blue{impute $x$} when using MICE (naively) is
$$x = \theta_{10} + \theta_{11} y + \ldots,$$
i.e., a \blue{linear relation} between $x$ and $y$ is assumed.
\begin{columns}[onlytextwidth]
\begin{column}{0.6\linewidth}
\begin{center}
\includegraphics[width = \linewidth]{graphics/Section09/p_qdr2.pdf}
\end{center}
\end{column}
\begin{column}{0.4\linewidth}
The imputed values \blue{distort the curved pattern} of the original data.
\end{column}
\end{columns}


## Quadratic Effect
The model fitted on the imputed data gives \blue{severely biased results}; the
non-linear shape of the curve has almost completely disappeared.

\begin{center}
\includegraphics[width = 0.6\linewidth]{graphics/Section09/p_qdr3.pdf}
\end{center}




## Interaction effect
Another example: consider the analysis model (again, assumed to be true)
$$y = \beta_0 + \beta_x x + \beta_z z + {\color{EMCdark}\boldsymbol{\mathbf{\beta_{xz} xz}}} + \ldots,$$
i.e., $y$ has a \blue{non-linear relationship} with $x$ due to the
\blue{interaction term}.


\begin{columns}[onlytextwidth]
\begin{column}{0.6\linewidth}
\begin{center}
\includegraphics[width = \linewidth]{graphics/Section09/p_int1.pdf}
\end{center}
\end{column}
\begin{column}{0.4\linewidth}
The original data shows a ``$<$'' shaped pattern.
\end{column}
\end{columns}


## Interaction effect
The model used to impute $x$ when using MICE (naively) is
$$x = \theta_{10} + \theta_{11} y + \theta_{12} z + \ldots,$$
i.e., a linear relation between $x$ and $y$ is assumed.
\begin{columns}[onlytextwidth]
\begin{column}{0.6\linewidth}
\begin{center}
\includegraphics[width = \linewidth]{graphics/Section09/p_int2.pdf}
\end{center}
\end{column}
\begin{column}{0.4\linewidth}
The ``$<$'' shaped pattern of the true data is \blue{distorted by the imputed values}.
\end{column}
\end{columns}


## Interaction effect
And the analysis on these naively imputed values leads to \blue{severely biased estimates}.

\vfill

\begin{center}
\includegraphics[width = 0.65\linewidth]{graphics/Section09/p_int3.pdf}
\end{center}




## Longitudinal outcome
\begin{columns}[onlytextwidth]
\begin{column}{0.54\linewidth}
\includegraphics[width = \linewidth]{graphics/Section09/plong1_1.pdf}
\end{column}
\begin{column}{0.45\linewidth}
\scriptsize
```{r ex_datatable, echo = FALSE, size = 'scriptsize', warning = F}
lsp <- rep("", nrow(longDF1_tab))
lsp[cumsum(table(longDF1$id[longDF1$id %in% longDF1$id]))] <- "\\hdashline"

longdatatab <- longDF1_tab %>%
  mutate(time = cell_spec(time, 'latex', escape = FALSE,
                          color = c("#000000", colvec)[as.numeric(factor(id))]),
         
         id = cell_spec(id, 'latex', escape = FALSE,
                        color = c("#000000", colvec)[as.numeric(factor(id))])
  ) %>%
  kable(format = 'latex', escape = FALSE, booktabs = TRUE,
        linesep = lsp,
        align = 'c',
        col.names = c("ID", "y", "$x_1$", "$x_2$", "$x_3$", "$x_4$", "time"),
        row.names = F)
 
longdatatab <- gsub("\\\\\n\\bottomrule", "", longdatatab, fixed = T)
longdatatab
```
\end{column}
\end{columns}
Here, $x_1, \ldots, x_4$ are baseline covariates, i.e., not measured repeatedly
(e.g. age at baseline, gender, education level, \ldots


```{r fake_imp, echo = F}

```



## Longitudinal outcome
```{r, results='asis', echo = FALSE}
cat(paste0("\\definecolor{colvec", seq_along(colvec), "}{HTML}{", gsub("#", '', colvec), "}"))
```

\begin{columns}[onlytextwidth]
\begin{column}{0.5\linewidth}
If we use MICE in the data in this (long) format,
each row would be regarded as independent,
which may cause bias and \blue{inconsistent imputations}.

\bigskip

Imputed values of baseline covariates are imputed with different values,
creating data that could not have been observed.
\end{column}
\begin{column}{0.45\linewidth}
\scriptsize
```{r imptablong, echo = FALSE}
colvec30 <- paste0("colvec", seq_along(colvec), "!30")

ltDFexlong2 <- longDF1_tab2 %>%
  mutate(time = cell_spec(time, 'latex', escape = FALSE,
                          color = c("#000000", colvec)[as.numeric(factor(id))]),
         x2 = cell_spec(x2, "latex", escape = FALSE,
                        background = ifelse(x2 %in% c("\\checkmark", "\\vdots"),
                                            '#FFFFFF',
                                            colvec30[as.numeric(factor(id)) - 1])),
         x3 = cell_spec(x3, "latex", escape = FALSE,
                        background = ifelse(x3 %in% c("\\checkmark", "\\vdots"),
                                            '#FFFFFF',
                                            colvec30[as.numeric(factor(id)) - 1])),
         x4 = cell_spec(x4, "latex", escape = FALSE,
                        background = ifelse(x4 %in% c("\\checkmark", "\\vdots"),
                                            '#FFFFFF',
                                            colvec30[as.numeric(factor(id)) - 1]))
  ) %>%
  mutate(id = cell_spec(id, 'latex', escape = FALSE,
                        color = c("#000000", colvec)[as.numeric(factor(id))])) %>%
  kable(format = 'latex', escape = FALSE, booktabs = T, linesep = lsp, align = 'c',
        col.names = c("ID", "y", "$x_1$", "$x_2$", "$x_3$", "$x_4$", "time"),
        row.names = FALSE, digits = 2)

gsub("\\\\\n\\bottomrule", "", ltDFexlong2, fixe = T)
```
\end{column}
\end{columns}



## Longitudinal outcome
\begin{columns}[onlytextwidth]
\begin{column}{0.7\linewidth}
\includegraphics[width = \linewidth]{graphics/Section09/gglongex1.pdf}
\end{column}
\begin{column}{0.25\linewidth}
Estimates can be severely biased.
\end{column}
\end{columns}
\vspace*{-2ex}




## Longitudinal outcome
In some settings \blue{imputation in wide format} may be possible.

\bigskip

\begin{columns}[onlytextwidth]
\begin{column}{0.54\linewidth}
\only<1>{
\includegraphics[width = \linewidth]{graphics/Section09/plong1_1.pdf}
}
\only<2>{
\includegraphics[width = \linewidth]{graphics/Section09/plong1_1b.pdf}
}
\end{column}
\begin{column}{0.45\linewidth}
\scriptsize
```{r impwidedat, echo = FALSE, size = 'scriptsize', warning = F}
longdatatab
```
\end{column}
\end{columns}



## Longitudinal outcome
\footnotesize
```{r impwidedat2, size = 'scriptsize', echo = F}
subset(wideDF1, subset = id < 5, select = c('id', grep('\\.', names(wideDF1), value = T))) %>%
  round(2) %>%
  rbind(., rep("\\vdots", 11)) %>%
  cbind(., '\\ldots' = c(rep("\\ldots", 4), "$\\ddots$")) %>%
  kable(format = 'latex', row.names = FALSE, booktabs = T, escape = F, digits = 1)
```
\normalsize

\bigskip

In this \blue{wide format data} frame, missing values in the outcome and measurement times
need to be imputed (to be able to use them as predictors to impute covariates),
even though we would not need to impute them for the analysis
(mixed model is valid when outcome measurements are M(C)AR).



## Longitudinal outcome
\begin{columns}
\begin{column}{0.7\linewidth}
\includegraphics[width = \linewidth]{graphics/Section09/gg_wideex1.pdf}
\end{column}
\begin{column}{0.3\linewidth}
Better, but large confidence intervals.
\end{column}
\end{columns}

\vspace*{-2ex}


## Longitudinal outcome
\begin{columns}
\begin{column}{0.7\linewidth}
\includegraphics[width = \linewidth]{graphics/Section09/plong2_1}
\end{column}
\begin{column}{0.3\linewidth}
When the data is very \blue{unbalanced}, transformation to wide format is not possible.

\bigskip

(Or at least transformation to wide format leads to variables with high proportions
of missing values.)
\end{column}
\end{columns}





## Longitudinal outcome
\begin{columns}
\begin{column}{0.5\linewidth}
\includegraphics[width = \linewidth]{graphics/Section09/plong2_2}
\onslide<2->{
\includegraphics[width = \linewidth]{graphics/Section09/plong2_3}
}\vspace*{-1ex}
\end{column}
\begin{column}{0.4\linewidth}
Naive approaches that are sometimes used are to
\begin{itemize}
\item \blue{ignore the outcome} in the imputation\onslide<2->{, or to}
\item<2-> use only the \blue{first/baseline outcome}
\end{itemize}

\bigskip

\onslide<3>{However, \blue{important information may be lost},
resulting in invalid imputations and biased results.}

\end{column}
\end{columns}




## Survival Data
In \blue{survival analysis}, the aim is to estimate the effect of covariates on
the \blue{time until an event} of interest happens.

\bigskip

\pause
Commonly used method: \blue{Cox proportional hazards model}

$$h(t) = h_0(t) \exp(x \beta_x + z\beta_z),$$

\begin{itemize}
\item $h(t)$: hazard = the instantaneous risk of an event at time $t$,
      given that the event has not occurred until time $t$
\item $h_0(t)$: unspecified baseline hazard
\item \blue{$x$} and \blue{$z$}: \blue{incomplete} and \blue{complete}
covariates, respectively
\end{itemize}

\bigskip

\pause
\blue{Survival outcomes} are usually represented by the \blue{observed event
time $T$} and the \blue{event indicator $D$} ($D = 1$: event, $D = 0$: censored).


## Survival Data
\blue{Naive use of MICE}
\begin{itemize}
\item $T$ and $D$ are treated just like any other variable.
\item The resulting imputation model for $X$ would have the form
$$p(x \mid T, D, \mathbf z) = \theta_0 + \theta_1 T + \theta_2 D + \theta_3 z + \ldots.$$
\end{itemize}

\bigskip

\pause
The \blue{correct conditional distribution} of $x$ given the other variables is, however,
$$\log p(x\mid T, D, z) = \log p(x\mid z) + D(\beta_x x + \beta_z z) - 
H_0(T)\exp(\beta_x x+\beta_z z) + const.,$$
where $H_0(T)$ is the cumulative baseline hazard. [@White2009]



## Survival Data
Using the naively assumed imputation model can lead to \blue{severe bias}:

\begin{center}
\includegraphics[width = 0.8\linewidth]{graphics/Section09/p_cox.pdf}
\end{center}

(Results from MICE imputation with two incomplete normal and one incomplete binary covariate.)


## References

