---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "10. Requirements for MICE to work (well)"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: [aspectratio=169]
bibliography: references.bib
---

```{r setup, include = FALSE}
projdir <- gsub("/Slides", "", getwd())

knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)


knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%")
```


## Joint and Conditional Distributions
\blue{Recall:} The MICE algorithm is based on the idea of Gibbs sampling.

\bigskip
\pause

Gibbs sampling exploits the fact that a joint distribution is fully determined
by its full conditional distributions.

\begin{center}
\tcbox[nobeforeafter, box align=base, colback = EMCdark, colframe=EMCdark, boxsep = 1ex]{
   \parbox{2.5cm}{\centering\color{white} joint\\distribution}
}
\quad
\parbox[c]{3cm}{
\tikzfancyarrow[3cm]{\scriptsize \textbf{Gibbs}}\\
\onslide<2>{\tikzfancyarrow[3cm, shape border rotate = 180]{\scriptsize \textbf{MICE}}}
}
\quad
\tcbox[nobeforeafter, box align=base, colback = EMCdark, colframe=EMCdark, boxsep = 1ex]{
   \parbox{2.5cm}{\centering\color{white}full\\conditionals}
}
\end{center}

\onslide<2>{In MICE, the full conditionals are not derived from the joint distribution:\\
we directly specify the full conditionals and hope a joint distribution exists.}



## Joint and Conditional Distributions
The \blue{uncertainty about whether a joint distribution exists} for the specified
set of imputation models is often considered to be mainly a theoretical problem.

\bigskip

In practice, violations only have little impact on results in many applications.

\bigskip

\pause
However, as we have seen in the examples on the previous slides,
there are \blue{settings where the direct specification} of the full
conditionals/imputation models \blue{may lead to problems}, causing biased results.



## Some Conditions and Definitions
Two important definitions:

\bigskip


\blue{Compatibility:}
\begin{quote}
A joint distribution exists, that has the full conditionals (imputation models)
as its conditional distributions.
\end{quote}

\blue{Congeniality:}
\begin{quote}
The imputation model is compatible with the analysis model.
\end{quote}


## Some Conditions and Definitions
\blue{Important requirements} for MICE to work well include:
\begin{itemize}[<+->]
\item Compatibility
\item Congeniality
\item MAR or MCAR (in the standard implementations)
\item \blue{All relevant variables} need to be included.\
      (Omission might result in MNAR.)
\item \textbf{\textcolor{red}{The outcome needs to be included}} as predictor variable\\
      (but we usually do not impute missing outcome values).
\item The imputation models (and analysis model) need to be \blue{correctly specified}
      (which is a requirement in any standard analysis).
\end{itemize}



## Why imputation with MICE can go wrong

\blue{What went wrong in our previous examples?}

\bigskip\pause

When incomplete variables have \blue{non-linear associations} with the outcome,
or with each other, the requirement(s) of \blue{\textit{compatibility} and/or
\textit{congeniality} are violated}.

\bigskip\pause

\blue{Omission, or inadequate inclusion, of the outcome} may result in \blue{MNAR} missing
mechanisms. The same is the case when other relevant predictor variables
are not used as predictor variables in the imputation.
\bigskip\pause

Furthermore, \blue{omission of variables} may lead to \blue{mis-specified models},
however, models may also be mis-specified when all relevant covariates are included,
but \blue{distributional assumptions} or the specified \blue{form of associations}
are incorrect.






## Alternatives to MICE
To \blue{avoid incompatible} and \blue{uncongenial} imputation models, we need to
\begin{itemize}
\item specify the joint distribution
\item and derive full conditionals / imputation models from this joint distribution
\end{itemize}
instead of specifying them directly.

\bigskip

\pause
\blue{Problem:}\
The joint distribution may not be of any known form:
\begin{eqnarray*}
\begin{array}{c}
x_1 \sim N(\mu_1, \sigma_1^2)\\
x_2 \sim N(\mu_2, \sigma_2^2)\\
\end{array}
& \Rightarrow &
\left(
  \begin{array}{c}
  x_1\\ x_2
  \end{array}
\right) \sim N\left(
                \left[
                  \begin{array}{c}
                  \mu_1\\ \mu_2
                  \end{array}
                \right], \left[
                            \begin{array}{cc}
                            \sigma_1^2 & \sigma_{12}\\
                            \sigma_{12} & \sigma_2^2
                            \end{array}
                          \right]
              \right)\\[2ex]
\text{\blue{but}\quad}
\begin{array}{l}
x_1 \sim N(\mu_1, \sigma_1^2)\\
x_2 \sim Bin(\mu_2)\\
\end{array}
& \Rightarrow &
\left(
  \begin{array}{c}
  x_1\\ x_2
  \end{array}
\right) \sim ???
\end{eqnarray*}



## Alternatives to MICE
Possible approaches:

\bigskip

Approach 1: \blue{Multivariate Normal Model}\
Approximate the joint distribution by a known multivariate distribution.

(usually the normal distribution; this is the approach mentioned in Section 01)

\bigskip

Approach 2: \blue{Sequential Factorization}\
Factorize the joint distribution into a (sequence of) conditional and a marginal
distributions.


## Multivariate Normal Model
\blue{Assumption:}\
The outcome and incomplete variables follow a \blue{joint multivariate normal
distribution}, conditional on the completely observed covariates $\mathbf X_c$,
parameters $\boldsymbol\theta$ and, possibly, random effects, $\mathbf b$:
$$ p(\mathbf y, \mathbf x_1,\ldots, \mathbf x_p \mid \mathbf X_c, \boldsymbol\theta,
     \mathbf b) \sim N(\mathbf \mu, \boldsymbol\Sigma)$$

\bigskip

\onslide<2>{
\textbf{How do we get that multivariate normal distribution?}
\begin{enumerate}
\item Assume \blue{all} incomplete variables and the outcome are \blue{(latent) normal}.
\item Specify linear (mixed) \blue{models based on observed covariates}.
\item \blue{Connect} using multivariate normal for \blue{random effects \& error terms}.
\end{enumerate}
}


## Multivariate Normal Model
\textbf{1. Latent normal assumption:}\vspace*{-3ex}
$$\text{e.g.: } \mathbf x_k \text{ binary }
  \rightarrow \text{ latent } \boldsymbol{\hat{\mathbf{x}}_k} \text{ is standard normal: }
  \left\{\begin{array}{c} \mathbf x_k = 1\\ \mathbf x_k = 0\end{array}\right.
  \text{ if } \begin{array}{c} \boldsymbol{\hat{\mathbf{x}}_k}\geq 0\\
  \boldsymbol{\hat{\mathbf{x}}_k} < 0\end{array}
$$

```{r, echo = F, fig.width = 6, fig.height = 3, out.width = "70%"}
par(mar = c(3.5, 4.4, 0.5, 0.5), mgp = c(2.5, 0.6, 0))
plot(seq(-4, 4, 0.1), dnorm(seq(-4, 4, 0.1)), type = "l", cex.lab = 1.5,
     xlab = expression(hat(x)[k]), ylab = expression(density~of~hat(x)[k]))
polygon(x = c(seq(-4, 0, 0.1), 0),
        y = c(dnorm(seq(-4, 0, 0.1)), 0), col = grey(0.9), border = 'transparent')
polygon(x = c(seq(4, 0, -0.1), 0),
        y = c(dnorm(seq(4, 0, -0.1)), 0), col = grey(0.7), border = 'transparent')
abline(v = 0, lty = 2)
lines(seq(-4, 4, 0.1), dnorm(seq(-4, 4, 0.1)))
text(x = -1, y = 0.05, label = bquote(x[k] == 0), cex = 1.5)
text(x = 1, y = 0.05, label = bquote(x[k] == 1), cex = 1.5)
```

## Multivariate Normal Model
\only<1>{\textbf{2. Specify models:}\\}
\only<2>{\textbf{2. Specify models / 3. Connect random effects \& error terms:}\\}
\begin{picture}(200, 120)(-20, 0)
\onslide<1->{
\put(3, 60){
    $\addtolength\arraycolsep{-.8ex}\def\arraystretch{1.5}
    \begin{array}{rclclcl}
    \mathbf y   &=& \mathbf X_c \boldsymbol{\beta}_y     &+& \mathbf{ Z_y} \;\mathbf{ b_y} &+& \boldsymbol{ \varepsilon_y}\\
    \textcolor{black!30}{\mathbf w} &=& \textcolor{black!30}{\mathbf X_c \boldsymbol{\beta}_w}
                                   &+& \textcolor{black!30}{\mathbf{ Z_w} \,\mathbf{ b_w}}
                                   &+& \textcolor{black!30}{\boldsymbol{ \varepsilon_w}}\\
    \boldsymbol{\hat{\mathbf{x}}}_1 &=& \mathbf X_c \boldsymbol{\beta}_{x_1} &+& \phantom{\mathbf Z_w\,} \boldsymbol{ \varepsilon_{x_1}}      &&\\[-1ex]
               & \vdots&                                & & \phantom{\mathbf Z_w\;} \vdots                                &&\\[-1ex]
    \boldsymbol{\hat{\mathbf{x}}}_p &=& \mathbf X_c \boldsymbol{\beta}_{x_p} &+& \phantom{\mathbf Z_w\,} \boldsymbol{ \varepsilon_{x_p}}      &&
    \end{array}
    $
    }
}
\thicklines
\onslide<2>{
\put(112, 87){\color{red} \oval(15, 35)}
\put(88, 62.5){\color{red} \oval(18, 93)}
\put(97, 15){\line(0,-1){12}}
\put(97, 3){\vector(1, 0){10}}
\put(110, 0.5){multivariate normal}
%
\put(120, 67){\line(0,-1){17}}
\put(120, 50){\vector(1, 0){10}}
\put(132.5, 47.5){multivariate normal (optional)}
}
\end{picture}


## Multivariate Normal Model
\begin{columns}[T,onlytextwidth]
\begin{column}{0.5\linewidth}
\blue{Advantages:}
\begin{itemize}
\item easy to specify
\item relatively easy to implement
\item relatively easy to sample from
\item works for longitudinal outcomes
\end{itemize}
\end{column}
\begin{column}{0.5\linewidth}
\blue{Disadvantages:}
\begin{itemize}
\item assumes linear associations
\end{itemize}
\end{column}
\end{columns}

\bigskip

\begin{center}
\parbox{0.6\linewidth}{Imputation with \blue{non-linear associations} or \blue{survival data} is
possible with \blue{extensions} of the multivariate normal approach.
}\end{center}


## Sequential Factorization
\onslide<1->{
The \blue{joint distribution} of two variables $y$ and $x$ can be written
as the product of conditional distributions:
$$p(y,x) = p(y\mid x)\;p(x)$$
(or alternatively $p(y,x) = p(x\mid y)\;p(y)$)
}

\vfill

\onslide<2->{
This can easily be \blue{extended for more variables}:
$$p(y,x_1,\ldots,x_p, X_c) = \underset{\text{analysis model}}{
                             \underbrace{p(y\mid x_1,\ldots,x_p, X_c)}}\;
p(x_1\mid x_2,\ldots,x_p, X_c)\;
\ldots\; p(x_p\mid X_c)$$%

where $x_1, \ldots, x_p$ denote incomplete covariates and $X_c$ contains all
completely observed covariates.
}


## Sequential Factorization
The \blue{analysis model} is part of the specification of the joint distribution.


\begin{itemize}\itemsep3ex
\item[\blue{\ding{225}}] The \blue{outcome}
        \begin{itemize}
        \item<1-> is \blue{automatically included in the imputation} procedure
        \item<1-> does not appear in any of the predictors of the imputation models:
        \end{itemize}
        \blue{\ding{225}} \blue{no need to approximate/summarize} complex outcomes!

\item[\blue{\ding{225}}]<2-> The \blue{parameters of interest} are obtained directly\\
      \blue{\ding{225}} Imputation and analysis in one step!

\item[\blue{\ding{225}}]<3-> \blue{Non-linear associations / interactions}
        \begin{itemize}
        \item specified in the analysis model
        \end{itemize}
        \blue{\ding{225} automatically taken into account!}
\end{itemize}


\bigskip

\onslide<4>{Since the joint distribution usually does not have a known form, Gibbs sampling
is used to estimate parameters and sample imputed values.}


## Sequential Factorization
\begin{columns}
\begin{column}{0.55\linewidth}
\blue{Advantages:}
\begin{itemize}
\item \blue{flexible:}
        \begin{itemize}
        \item any outcome type
        \item separate imputation models per variable
        \end{itemize}
\item can handle \blue{non-linear associations} and interactions
\item assures \blue{congeniality and compatibility}
\end{itemize}
\end{column}
\begin{column}{0.45\linewidth}
\blue{Disadvantages:}
\begin{itemize}
\item specification takes requires time and consideration
\item sampling may be more \blue{computationally intensive}
\end{itemize}
\end{column}
\end{columns}


## Some Relevant R Packages
For complex settings there are alternatives to \blue{mice}:

For example the \textsf{R} packages \blue{JointAI}, \blue{smcfcs} and \blue{jomo}.

\vfill

\begin{itemize}\itemsep2ex
\item<2-> they use \blue{Bayesian methodology} to impute values
\item<3-> \blue{jomo} and \blue{smcfcs} perform \blue{multiple imputation};\\
      the imputed datasets that can then be analysed the same way data imputed
      by \textbf{mice} would be analysed.
\item<4-> \blue{JointAI} works \blue{fully Bayesian}
      \begin{itemize}
      \item performs analysis and imputation simultaneously
      \item[\blue{\ding{225}}] results from the analysis model of interest are
obtained directly
      \end{itemize}
\end{itemize}


## R package smcfcs
\href{https://cran.r-project.org/package=smcfcs}{\blue{Substantive Model Compatible Fully Conditional Specification}},\
a hybrid approach between FCS and sequential factorization [@Bartlett2015]

\bigskip
\textbf{smcfcs} (version 1.4.0) can impute incomplete covariates in
\begin{columns}
\begin{column}{0.4\linewidth}
\begin{itemize}
\item linear regression
\item logistic regression
\item poisson regression
\item Weibull survival models
\end{itemize}
\end{column}
\begin{column}{0.5\linewidth}
\begin{itemize}
\item Cox proportional hazard models
\item competing risk survival models
\item nested case control studies
\item case cohort studies
\end{itemize}
\end{column}
\end{columns}
\vspace*{3ex}
while ensuring compatibility between analysis model and imputation models.

\bigskip

For more information see the help files and the
\href{https://cran.r-project.org/package=smcfcs/vignettes/smcfcs-vignette.html}%
{vignette}.



## R Package jomo
\href{https://cran.r-project.org/package=jomo}{\blue{JOint MOdel imputation}} using the multivariate normal approach,\
with \blue{extensions to assure compatibility} between analysis and imputation models. [@Carpenter2012]

\bigskip

\textbf{jomo} (version 2.6-10) can handle
\begin{itemize}
\item linear regression
\item generalized linear regression
\item proportional odds (ordinal) probit regression
\item linear mixed models
\item generalized linear mixed models
\item (ordinal) cumulative link mixed models
\item Cox proportional hazards models.
\end{itemize}

For more info see the \href{https://cran.r-project.org/package=jomo}{help file}.




## R Package JointAI

\href{https://cran.r-project.org/package=JointAI}{\blue{Joint Analysis and Imputation}}, \
uses the \blue{sequential factorization approach} to perform
simultaneous analysis and imputation. [@Erler2016; @Erler2017]

\bigskip

\textbf{JointAI} (version 0.6.1) can analyse incomplete data using

\begin{columns}
\begin{column}{0.4\linewidth}
\begin{itemize}
\item linear regression
\item generalized linear regression
\item linear mixed models
\item generalized linear mixed models
\end{itemize}
\end{column}
\begin{column}{0.6\linewidth}
\begin{itemize}
\item (ordinal) cumulative logit regression
\item (ordinal) cumulative logit mixed models
\item parametric (Weibull) survival models
\item Cox proportional hazards models
\end{itemize}
\end{column}
\end{columns}
\vspace*{3ex}
while assuring compatibility between analysis model and imputation models
when non-linear functions or interactions are included.


## R Package JointAI
The necessary \blue{Gibbs sampling} is performed using \blue{\textsf{JAGS}}
(an external program), which is free, but needs to be
installed from \url{https://sourceforge.net/projects/mcmc-jags/files/}.

\bigskip

\blue{JointAI} can be installed from CRAN or \href{https://github.com/nerler/JointAI}{GitHub} (development version containing bug fixes and other improvements)
\small
```{r eval = FALSE}
install.packages("devtools")
devtools::install_github("NErler/JointAI")
```


\blue{JointAI} has its own web page (\href{https://nerler.github.io/JointAI/}{https://nerler.github.io/JointAI/})
with several vignettes on \href{https://nerler.github.io/JointAI/articles/VisualizingIncompleteData.html}{Visualization of Incomplete Data},
a \href{https://nerler.github.io/JointAI/articles/MinimalExample.html}{Minimal Example},
details on
\href{https://nerler.github.io/JointAI/articles/ModelSpecification.html}{Model Specification},
\href{https://nerler.github.io/JointAI/articles/}{etc.}

## References {.allowframebreaks}
