---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "11. Imputation with Non-linear Functional Forms"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: [aspectratio=169]
bibliography: references.bib
---

```{r setup, include = FALSE}
projdir <- gsub("/Slides", "", getwd())

runimps <- FALSE
  
knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)


knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%")

options(width = 120)

suppressPackageStartupMessages(library("mice"))
suppressPackageStartupMessages(library("JointAI"))

library(kableExtra)
library(ggplot2)


# simulate data
set.seed(2018)
N <- 200
x <- rnorm(N)
z <- rbinom(N, size = 1, prob = plogis(x))
y <- x + x^2 + z + x*z + rnorm(N, 0, 0.5)

DF_nonlin <- data.frame(y = y, x = x, z = z)

# model on complete data
mod_nonlin <- lm(y ~ x + I(x^2) + z + x:z, data = DF_nonlin)

# create missing values
DF_nonlin$x[sample(1:length(x), size = N/2)] <- NA
```


## Imputation with mice
There is \blue{no strategy} for MICE that can \blue{guarantee valid imputations}
when non-linear functional forms and/or interactions are involved,
but \blue{some settings} in \textbf{mice} \blue{may help} to reduce bias in the
resulting estimates.

\bigskip\pause

For imputation of variables that have non-linear associations
\begin{itemize}
\item \blue{PMM often works better} than imputation with a normal model,
\item the \blue{Just Another Variable} approach can reduce bias in interactions,
\item passive imputation
\item \Rstring{quadratic} can help to impute variables with quadratic association.
\end{itemize}

## Imputation with mice
For demonstration, we use a simulated example dataset \Robj{DFnonlin}:

\begincols[onlytextwidth]
\begincol{0.54\linewidth}
\begin{itemize}
\item[\blue{y}] continuous outcome
\item[\blue{x}] continuous (normal) covariate\
      (50\% missing values MCAR)
\item[\blue{z}] binary covariate (complete)
\end{itemize}

\bigskip\vspace*{3ex}

We assume a

* \blue{quadratic effect} of \blue{$x$} on \blue{$y$}, and
* an \blue{interaction} between \blue{$x$} and \blue{$z$}

\endcol
\begincol{0.4\linewidth}
\small
```{r}
head(DF_nonlin)
dim(DF_nonlin)
```
\endcol
\endcols

## Imputation with mice: JAV
\blue{Just Another Variable (JAV)} approach:
\begin{itemize}
\item \blue{pre-calculate} the non-linear form (or interaction term) in the incomplete data,
\item add it as a column to the dataset, and
\item impute it as if it was just another variable.
\end{itemize}

\bigskip\pause

\small
```{r}
DF2 <- DF_nonlin         # copy of the data, only for this example
DF2$xx <- DF2$x^2        # pre-calculate the quadratic term
DF2$xz <- DF2$x * DF2$z  # pre-calculate the interaction
```

```{r}
# JAV imputation (using pmm and full predictor matrix)
impJAV <- mice(DF2, maxit = 20, printFlag = FALSE)
```
\normalsize


## Imputation with mice: JAV

To \blue{relax the assumption} of linear associations even more, we could introduce
\blue{additional interactions} with the outcome.

\bigskip
In this example, we can add an interaction between $z$ and $y$:

\small
```{r}
DF3 <- DF2               # make another copy of the data
DF3$yz <- DF3$y * DF3$z  # add interaction y and z
```

```{r}
# JAV imputation with additional interaction
impJAV2 <- mice(DF3, maxit = 20, printFlag = FALSE)
```

## Imputation with mice: Passive Imputation

Alternative: impute all non-linear terms and interactions passively:
\small
```{r}
# adapt the imputation method (we re-use the vector from impJAV2 here)
meth_passive <- impJAV2$method
meth_passive[c("xx", "xz", "yz")] <- c("~I(x^2)", "~I(x*z)", "~I(y*z)")
```
\pause
```{r}
# adapt the predictor matrix (we re-use the matrix from impJAV2 here)
pred_passive <- impJAV2$predictorMatrix
pred_passive['x', 'xx'] <- 0
pred_passive[c('x', 'z'), 'xz'] <- 0
pred_passive[c('y', 'z'), 'yz'] <- 0
```
\pause
```{r}
imp_passive <- mice(DF3, method = meth_passive,
                    predictorMatrix = pred_passive,
                    maxit = 20, printFlag = FALSE)
```


## Imputation with mice: Polynomial Combination

The imputation method \Rstring{quadratic} uses the \blue{``polynomial combination''
method} to impute covariates that have a \blue{quadratic association} with the
outcome [@Buuren2012 \ pp. 139--141; @Vink2013].

\bigskip

\begin{itemize}
\item[\ding{225}] ensure the \blue{imputed values} for $x$ and $x^2$ are \blue{consistent}
\item[\ding{225}] \blue{reduce bias} in the subsequent analysis that uses $x$ and $x^2$
\end{itemize}

\bigskip\pause

\small
```{r}
# adapt the imputation method (we re-use the vector from impJAV here)
methqdr <- impJAV$meth
methqdr[c("x", "xx", "xz")] <- c("quadratic", "~I(x^2)", "~I(x*z)")
```
\normalsize
\blue{\ding{225}} Here we use passive imputation for $x^2$ and the interaction.

## Imputation with mice: polynomial combination
\small
```{r, warning = FALSE}
# adapt the predictor matrix (we re-use the matrix from impJAV here)
predqdr <- impJAV$pred
predqdr['x', "xx"] <- 0          # prevent feedback
predqdr[c('x', 'z'), 'xz'] <- 0  # prevent feedback

impqdr <- mice(DF3, meth = methqdr, pred = predqdr,
               maxit = 20, printFlag = FALSE)
```
\normalsize

\bigskip\pause
For comparison, we also run a naive version (using defaults):
\small
```{r}
# naive imputation, using only y, x, z
impnaive <- mice(DF_nonlin, printFlag = FALSE)
``` 


## Imputation with mice
```{r, echo = FALSE}
res_impnaive <- with(impnaive, lm(y ~ x + I(x^2) + z + x:z)) %>% pool %>% summary(conf.int = TRUE)
res_JAV <- with(impJAV, lm(y ~ x + xx + z + xz)) %>% pool %>% summary(conf.int = TRUE)
res_JAV2 <- with(impJAV2, lm(y ~ x + xx + z + xz)) %>% pool %>% summary(conf.int = TRUE)
res_qdr <- with(impqdr, lm(y ~ x + xx + z + x:z)) %>% pool %>% summary(conf.int = TRUE)
res_passive <- with(imp_passive, lm(y ~ x + xx + z + x:z)) %>% pool %>% summary(conf.int = TRUE)

reslist <- list(naive = as.data.frame(res_impnaive[, c("term", "estimate", "2.5 %", "97.5 %")]),
                pas = as.data.frame(res_passive[, c("term", "estimate", "2.5 %", "97.5 %")]),
                JAV = as.data.frame(res_JAV[, c("term", "estimate", "2.5 %", "97.5 %")]),
                JAV2 = as.data.frame(res_JAV2[, c("term", "estimate", "2.5 %", "97.5 %")]),
                qdr = as.data.frame(res_qdr[, c("term", "estimate", "2.5 %", "97.5 %")])
)
```

```{r comp_plot, echo = FALSE, fig.width = 7.5, fig.height = 4}
plotfun <- function(reslist, models = c("naive", "JAV", "JAV2", "pas", "qdr")) {
  res_nonlin <- lapply(reslist, function(x) {
    colnames(x) <- gsub('5%', '5 %', colnames(x)) %>%
      gsub('Mean', 'estimate', .)
    
    x$term <- gsub("xx", "I(x^2)", x$term) %>%
      gsub("z1", "z", .) %>%
      gsub("xz", "x:z", .)
    x
  })
  
  plot_nonlin <- merge(reshape2::melt(res_nonlin, id.vars = colnames(res_nonlin[[1]])), 
                       data.frame(term = names(coef(mod_nonlin)),
                                  betas = coef(mod_nonlin))
  )
  
  plot_nonlin$type = 'true'
  
  
  polyDF_nonlin <- data.frame(x = rep(c(0, length(models) + 1, length(models) + 1, 0), 5),
                              y = c(apply(confint(mod_nonlin), 1, rep, each = 2)),
                              term = rep(c("(Intercept)", "x", "I(x^2)", "z", "x:z"),
                                         each = 4),
                              type = 'true'
  )

  
  ggplot(plot_nonlin, aes(x = L1, y = estimate)) +
    geom_polygon(data = polyDF_nonlin,
                 aes(x = x, y = y, fill = type), alpha = 0.2) +
    geom_point() +
    geom_errorbar(aes(ymin = `2.5 %`, ymax = `97.5 %`), width = 0.3) +
    facet_wrap("term", scales = 'free') +
    geom_hline(data = plot_nonlin, aes(yintercept = betas, lty = type)) +
    scale_x_discrete(limits = models) +
    ylab("estimate & 95% CI") +
    xlab("") +
    scale_fill_manual(name = '', limits = 'true', values = 'blue', 
                      labels = 'Result from\nthe analysis of\nthe full data') +
    scale_linetype_manual(name = '', limits = 'true', values = 2,
                          labels = 'Result from\nthe analysis of\nthe full data') +
    theme(panel.background = element_rect(fill = grey(0.98), color = grey(0.85)),
          panel.grid = element_blank(),
          legend.position = c(0.85, 0.2),
          legend.key.width = unit(10, "mm"),
          legend.key.height = unit(10, "mm")) +
    coord_cartesian(xlim = c(1.2, length(models) - 0.2))
}

plotfun(reslist)
```

\vspace*{-2ex}

## Imputation with JointAI
The syntax we use to analyse and impute the current example using \blue{JointAI} is
similar to the specification of a standard linear model using \Rfct{lm}.

\small
```{r JointAI_nonlin, eval = FALSE}
library("JointAI")
JointAI_nonlin <- lm_imp(y ~ x*z + I(x^2), data = DF_nonlin,
                         n.iter = 2500)
```

```{r runJointAI_nonlin, eval = runimps, echo = FALSE}
JointAI_nonlin <- lm_imp(y ~ x*z + I(x^2), data = DF_nonlin, n.iter = 2500,
                         seed = 1234)
save(JointAI_nonlin, file = file.path(projdir, "Slides/workspaces/JointAI_nonlin.RData"))
``` 

```{r getJointAI_nonlin, echo = FALSE, message = FALSE}
load(file.path(projdir, "Slides/workspaces/JointAI_nonlin.RData"))
```

\pause

Convergence of the Gibbs sampler can be checked using a `traceplot`.
```{r JointAI_nonlin_trace, eval = FALSE}
traceplot(JointAI_nonlin, ncol = 3, use_ggplot = TRUE)
```

\pause

Results (no separate analysis \& pooling is necessary) can be obtained with
the \Rfct{summary} function:

```{r eval = FALSE}
summary(JointAI_nonlin)
```


## Imputation with JointAI: Convergence
```{r fig.width = 7, fig.height = 3.8, echo = FALSE}
traceplot(JointAI_nonlin, ncol = 3, use_ggplot = TRUE) +
  theme(legend.position = 'none', panel.background = element_blank())
```


## Imputation with JointAI: Model Summary
\scriptsize
```{r, echo = FALSE}
res_JointAI_nonlin <- summary(JointAI_nonlin)
sumJointAI <- capture.output(summary(JointAI_nonlin))

cat(paste0(sumJointAI[1:20], collapse = "\n"),
    "[...]")
```

## Imputation with JointAI: Model Summary
\scriptsize
```{r, echo = FALSE}
cat("[...]","\n", 
    paste0(sumJointAI[20:27], collapse = "\n")
)
```



## Imputation with Non-linear Effects: Comparison

```{r, echo = FALSE}
reslist <- c(reslist,
             list(JointAI = cbind(term = names(coef(JointAI_nonlin)), 
                                   as.data.frame(res_JointAI_nonlin$stat[
                                     names(coef(JointAI_nonlin)), c(1,3,4)])
                   )))
```

```{r, echo = FALSE, fig.width = 7.5, fig.height = 4}
plotfun(reslist, models = c("naive", "JAV", "JAV2", "pas", "qdr", "JointAI"))
```
\vspace*{-2ex}

## Imputation with Non-linear Effects: Comparison

\blue{Just Another Variable:}\vspace*{-1ex}

 * Easy specification
 * usually less bias than naive approach
 * inconsistent imputed values
 
\pause
\blue{passive imputation}\vspace*{-1ex}

 * easy specification
 * consistent imputed values
 * less flexible than JAV

\pause
\blue{polynomial combination}\vspace*{-1ex}

* consistent imputed values
* only available for quadratic association
* often numeric instabilities (warning messages)


## Imputation with Non-linear Effects: Comparison

\blue{JointAI}\vspace*{-1ex}

 * theoretically valid approach (= unbiased)
 * similar specification to standard models
 * simultaneous analysis & imputation instead of three steps

\pause

To use JointAI appropriately and to interpret the results correctly 
requires more knowledge about the underlying method than can be 
covered in this course.

\pause   

\blue{Note:}\
The example used here only serves to demonstrate the different approaches.
We cannot use these results to conclude which approach works better in general.

## References
