---
title: "EP16: Missing Values in Clinical Research: Multiple Imputation"
subtitle: "14. Strategies for using MICE"
author: "Nicole Erler"
institute: "Department of Biostatistics, Erasmus Medical Center"
date: ""
email: "n.erler@erasmusmc.nl"
output:
  beamer_presentation:
    keep_tex: false
    template: mytemplate.latex
    includes:
      in_header: [SlideTemplate.tex, defs.tex]
    incremental: false
classoption: [aspectratio=169]
bibliography: references.bib
---


```{r setup, include = FALSE}
projdir <- gsub("/Slides", "", getwd())

library(kableExtra)

knitr::knit_hooks$set(
  nospace = function(before, options, envir) {
    if (before) {
      knitr::asis_output("\\vspace*{-1.5ex}")
    }
  }
)


knitr::opts_chunk$set(echo = TRUE, nospace = TRUE, nospaceafter = TRUE,
                      fig.align = 'center', out.width = "100%")
```


## Number of Imputations

In \blue{early publications} on multiple imputation:

* 3 -- 5 imputations are sufficient
* still is a common assumption in practice [@Rubin2004]

\bigskip

Reasoning behind using a small number of imputations:

* \blue{storage of imputed data was ``expensive''} (no longer the case)
* larger number of imputations would only have little advantage [@Schafer1997]


## Number of Imputations
More \blue{recent work} from various authors considers

* the \blue{efficiency} of the pooled estimates
* \blue{reproducibility} of the results,
* \blue{statistical} power of tests or 
* the \blue{width of the resulting confidence intervals} compared to the width of the true confidence intervals

[@White2011; @Buuren2012; @Graham2007]

## Number of Imputations
A \blue{rule of thumb} [@White2011]:

\blue{The number of imputed datasets} should be similar to the 
\blue{percentage of incomplete cases}.


\blue{Alternative:}\
The percentage of complete cases depends on the size of the dataset\
\blue{\ding{225} average percentage of missing values} per
variable [@Buuren2012]

\bigskip

Generally:\vspace*{-1ex}

* using \blue{more imputed datasets should be preferred}
* especially in settings with acceptable computational burden

Even though results are unlikely to change with a larger number of imputations,
it can increase the efficiency and reproducibility of the results.


## What to do with large datasets?
General advice:
\blue{Include as much information as possible} in the imputation.

\bigskip

Using a large number of predictor variables
\begin{itemize}
\item makes the \blue{MAR assumption more plausible} 
      (\blue{\ding{225}} reduces bias due to MNAR missingness)
\item can \blue{reduce uncertainty} about the missing values
\end{itemize}

\bigskip
\pause

This \blue{works in small or medium sized datasets}
(20 -- 30 separate variables, i.e. without interactions, variables derived
from others, \ldots).

\bigskip

\blue{In large datasets} (contain hundreds or thousands of
variables) this is \blue{not feasible}. [@Buuren2012]




## What to do with large datasets?
For large datasets a possible strategy is to


* Include all \blue{variables used in the analysis model(s)}\
      (including the outcome!)
      
\medskip
      
* Include auxiliary variables
    * if they are \blue{strong predictors of missingness}, or
    * if they have \blue{strong associations with the incomplete variables}
    * \blue{only if they do not have too many missing values} themselves
    * only in those imputation models for which they are \blue{relevant}
    
\medskip

* Use \blue{summary scores} when multiple items referring to the same concept.
      

## How much missing is too much?
There is \blue{no clear cut-off} for the proportion of missing values that can
be handled adequately using MICE (or any other imputation method).

\bigskip

The amount of missingness that can be handled \blue{depends on the information that
is available} to impute it.

\begin{itemize}
\item Are there \blue{strong predictor variabels} available \& observed?
\item Are there \blue{sufficient observed cases} to get reliable estimates for the
      predictive distribution?
\end{itemize}

\vfill

\pause
\textbf{Example:}
\begin{itemize}
\item In a set of $N = 50$ cases, 50\% missing values leaves 25 cases to
      estimate the parameters of the predictive distribution.
\item In a large set of $N = 5000$ subjects, 50\% missing cases leaves 2500
      observed cases to estimate parameters.
\end{itemize}




## Imputation of Outcomes
Usually, \blue{missing outcome values are not imputed}.

\bigskip

\textbf{Why?}\
When there are no auxiliary variables, imputation and analysis model are equal.
\begin{itemize}
\item Parameters of the imputation model are estimated on observed cases of the outcome.
\item Imputed values will fit the assumed model perfectly.
\item Including imputed cases in the analysis does not add any information.
\end{itemize}

## Imputation of Outcomes
\textbf{Exception:}
\begin{itemize}
\item When very strong auxiliary variables are available.
\item Outcomes may be imputed when one imputation is performed for
      several analysis models, because not imputing the outcome(s) would mean
      \begin{itemize}
      \item excluding cases with missing outcome(s) from the imputation, or
      \item excluding the outcome variable(s) as predictor(s).
      \end{itemize}
\end{itemize}



## Notes of Caution \& Things to Keep in Mind
Multiple imputation is \blue{not a quick and easy solution for missing data}.


It requires \blue{care and knowledge} about
\begin{itemize}
\item the \blue{data} to be imputed (and the context of the data),
\item the statistical \blue{method} used for imputation, and
\item the \blue{software} implementation used.
\end{itemize}

\vfill

Moreover
\begin{itemize}
\item \blue{Never accept default settings of software blindly.}
\item \blue{Question the plausibility of the MAR assumption.}\\
      If it is doubtful, use sensitivity analysis.
\end{itemize}


## Notes of Caution \& Things to Keep in Mind
\begin{itemize}\itemsep2ex
\item \blue{Use as much information as possible}
      \begin{itemize}
      \item include all covariates \textcolor{red}{and the outcome}
      \item use auxiliary information
      \item use the most detailed version of variables if possible
      \end{itemize}
\item \blue{Avoid feedback} from derived variables to their originals.
\item Think carefully how to handle variables that are derived from other variables.
\item Consider the impact the \blue{visit sequence} may have.
\item \blue{Imputation models must fit the data}\\
      (correct assumption of error distribution and functional forms and
      possible interactions of predictor variables).
\end{itemize}

## Notes of Caution \& Things to Keep in Mind
\begin{itemize}\itemsep2ex
\item Choose an appropriate \blue{number of imputations}.
\item Make sure the imputation algorithm has \blue{converged}.
\item Use \blue{common sense} when evaluating if the imputed values are plausible.
\item Be aware of the assumptions of your analysis model
      \begin{itemize}
      \item non-linear effects
      \item interactions
      \item complex outcomes
      \end{itemize}
\end{itemize}



## Tips & Tricks
In complex settings, variables may need to be \blue{re-calculated} or
\blue{re-coded} after imputation:

* Use `complete()`{.R} to convert the imputed data from a \Robj{mids} object
      to a \Robj{data.frame}.
* Perform the necessary calculations.
* Convert the changed \Robj{data.frame} back to a \Robj{mids} object
      using the functions such as `mice::as.mids()`{.R}, `miceadds::datalist2mids()`{.R},
      `mitools::imputationList()`{.R}, \ldots

\vfill
\pause

Not just in imputation: Set a \blue{seed value} to create reproducible results.


* in \includegraphics[width = 1em]{TemplateGraphics/Rlogo.png}: `set.seed()`{.R}
* in `mice()`{.R}: argument \Rarg{seed}



## Imputation Methods
We have focussed on a few imputation methods that cover the most common
types of data, but there are many more methods implemented.

\bigskip

Imputation methods implemented in the \textbf{mice} package:\vspace*{-1ex}
\footnotesize
```{r getmicemeths, echo = FALSE, warning = FALSE, results = 'asis'}

options(knitr.kable.NA = '')

used <- paste0("mice.impute.",
               c("norm", "pmm", "midastouch", "logreg", "polyreg", "polr",
                 "2lonly.norm", "2lonly.pmm", "quadratic")
)

mentioned <- paste0("mice.impute.",
                    c("2l.lmer", "2l.bin", "2l.norm", "2l.pan", "2lonly.mean",
                      "passive",
                      "logreg.boot", "norm.boot", "norm.predict")
)


micemeth <- sort(grep("mice.impute", getNamespaceExports("mice"), value = TRUE)) %>%
  cell_spec("latex", color = ifelse(. %in% used, "blue",
                                    ifelse(. %in% mentioned, "green!40!black", "black")))


methtab1 <- matrix(data = c(micemeth, rep(NA, 3 - length(micemeth) %% 3)),
                   ncol = 3)

methtab1 %>%
  kable(format = "latex", booktabs = TRUE, linesep = "", escape = F) %>%
  kable_styling(position = "center")
```
\normalsize

\blue{Note: }\parbox[t]{0.75\linewidth}{That a method is implemented does
not mean you need to / should use it.}


## Imputation Methods
Imputation methods implemented in the \textbf{miceadds} package:
\footnotesize
```{r getmiceaddsmeths, echo = FALSE, warning = FALSE}
miceaddsmeth <- sort(grep("mice.impute", getNamespaceExports("miceadds"),
                          value = TRUE))
methtab2 <- matrix(data = c(miceaddsmeth, rep(NA, length(miceaddsmeth) %% 3)),
                   ncol = 3)

methtab2 %>%
  kable(format = "latex", booktabs = TRUE, linesep = "", escape = T) %>%
  kable_styling(position = "center", font_size = 7.5)
```



## Imputation Methods
Imputation methods implemented in the \textbf{micemd} package:
```{r getmicemdmeths, echo = FALSE, warning = FALSE}
micemdmeth <- sort(grep("mice.impute", getNamespaceExports("micemd"), 
                        value = TRUE))
methtab2 <- matrix(data = c(micemdmeth, rep(NA, length(micemdmeth) %% 2)),
                   ncol = 2)


methtab2 %>%
  kable(format = "latex", booktabs = TRUE, linesep = "", escape = F) %>%
  kable_styling(position = "center")
```






## Other R Packages for Imputation
\blue{CRAN Task View on Missing Data:}

\href{https://cran.r-project.org/web/views/MissingData.html}{https://cran.r-project.org/web/views/MissingData.html}

\begin{itemize}
\item gives an overview on the available \includegraphics[width = 1em]{TemplateGraphics/Rlogo.png}
      packages for missing data / imputation
\item good point to start when searching for a package with a particular functionality
\end{itemize}

## Other R Packages for Imputation
```{r getallpkgs, echo = FALSE}
getPackagesWithTitle <- function() {
  chooseCRANmirror(graphics = FALSE, ind = 1)
  contrib.url(getOption("repos")["CRAN"], "source")
  description <- sprintf("%s/web/packages/packages.rds",
                         getOption("repos")["CRAN"])
  con <- if (substring(description, 1L, 7L) == "file://") {
    file(description, "rb")
  } else {
    url(description, "rb")
  }
  on.exit(close(con))
  db <- readRDS(gzcon(con))
  rownames(db) <- NULL
  
  db[, c("Package", "Title", "Description")]
}

allpkgs <- as.data.frame(getPackagesWithTitle(), stringsAsFactors = FALSE)

words <- c("impute", "imputation", "missing", "incomplete")
res <- sapply(words, function(x) {
  apply(allpkgs, 2, grep, pattern = x, ignore.case = TRUE)
}, simplify = FALSE)

# allpkgs[res$incomplete$Title, ]
#
res2 <- lapply(lapply(res, unlist), unique)

```


Currently, there are \blue{`r length(unique(unlist(res)))` packages} available
on CRAN that use the word \blue{``missing'', ``impute'', ``imputation'' or ``incomplete''}
in either the title or description.

\bigskip

Not all of these packages perform imputation or are useful for our purposes,
but even if we excluded those packages, the number of useful packages for
dealing with missing data would still be to large to mention them all.

\bigskip

\blue{\ding{225} The mice package is often a good option, but certainly not the
only option to perform imputation!}



## Imputation in other Software
In this second half of the course, we have focused on (multiple) imputation
using \includegraphics[width = 1em]{TemplateGraphics/Rlogo.png}.

\bigskip

Naturally, \includegraphics[width = 1em]{TemplateGraphics/Rlogo.png}
is not the only statistical software that can perform
multiple imputation.
\begin{itemize}
\item \blue{Stata, SAS and MPLUS} provide packages/functions to perform multiple
      imputation and pool the results.
\item There are macros and additional packages available, e.g.,
      \blue{smcfcs} is implemented for \blue{Stata} as well
\item \blue{SPSS} provides some functionality to perform MI
\end{itemize}


## Other Approaches to Handle Missing Values
Finally, we should not forget that
\blue{MICE is not the only method to handle missing values}.

\bigskip

Besides MICE, \blue{multiple imputation} can be performed in a
\blue{joint model approach} (as for instance implemented in the \textsf{R}
package \textbf{jomo}).

\bigskip

Furthermore,
\begin{itemize}
\item \blue{direct likelihood methods},
\item \blue{fully Bayesian methods} (as implemented in \textbf{JointAI}), or
\item \blue{weighted estimating equations}
\end{itemize}
are valid approaches and may \blue{in certain settings be superior}.


## References {.allowframebreaks}
