---
title: "Getting to Know Your Incomplete Data"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: style/footer.html
    css: style/style.css
---

```{r, include = FALSE}
projdir <- gsub("/Practicals", "", getwd())

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}

options(width = 100)

set.seed(2020)


library(knitr)
library(VIM)
library(visdat)
library(mice)
library(JointAI)
library(plyr)
library(corrplot)

load(file.path(projdir, "Practicals/data/EP16dat1.RData"))
```

## Preface {data-progressive=FALSE}
### R packages

In this practical, a number of R packages are used.
If any of them are not installed you may be able to follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:

* `r R.version.string`
* `mice` (version: `r packageVersion("mice")`)
* `visdat` (version: `r packageVersion("visdat")`) 
* `JointAI` (version: `r packageVersion("JointAI")`)
* `VIM` (version: `r packageVersion("VIM")`)
* `plyr` (version: `r packageVersion("plyr")`)
* `corrplot` (version: `r packageVersion("corrplot")`)
* `ggplot2` (version: `r packageVersion("ggplot2")`)


### Help files
You can find help files for any function by adding a `?` before the name of the 
function.

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


## Dataset 
### Overview {.tabset .tabset-fade .tabset-pills}

For this practical, we will use the **EP16dat1** dataset, which is a 
subset of the [NHANES (National Health and Nutrition Examination Survey)](https://www.cdc.gov/nchs/nhanes/index.htm) data.

To get the **EP16dat1** dataset, load the file `EP16dat1.RData`.
You can download it [here](https://nerler.github.io/EP16_Multiple_Imputation/practical/data/index.html).

To load this dataset into R, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the file on your computer.

If you know the path to the file, you can also use `load("<path>/EP16dat1.RData")`.

#### Task
::: {.task}
Take a first look at the data. Useful functions are
`dim()`, `head()`, `str()` and `summary()`.
:::

#### Solution
```{r load-mice-solution}
dim(EP16dat1)
head(EP16dat1)
str(EP16dat1)
summary(EP16dat1)
```


### Variable coding {.tabset .tabset-fade .tabset-pills}
It is important to check that all variables are coded correctly, i.e., have
the correct `class`. Imputation software (e.g., the **mice** package) uses the
`class` to automatically select imputation methods.
When importing data from other software, it can happen that factors become
continuous variables or that ordered factors lose their ordering.

`str()` (in the solution above) showed that in the **EP16dat1** data the 
variables `smoke` and `alc` are correctly specified as ordinal variables,
but `educ` is an unordered factor.

#### Task
:::{.task}
Using `levels(EP16dat1$educ)` we can print the names of the categories of `educ`.
Convert the unordered factor to an ordered factor, for example using `as.ordered()`.
Afterwards, check if the conversion was successful.
:::

#### Solution
```{r ordered_educ-solution}
levels(EP16dat1$educ)

EP16dat1$educ <- as.ordered(EP16dat1$educ)

str(EP16dat1$educ)
```

```{r educrecode, echo = FALSE}
EP16dat1$educ <- as.ordered(EP16dat1$educ)
```


## Distribution of missing values
### Missing data pattern {.tabset .tabset-fade .tabset-pills}
In the `summary()` we could already see that there are missing values in several
variables. The missing data pattern can be obtained and visualized by several
functions from different packages.
Examples are 

* `md.pattern()` from package **mice**
* `md_pattern()` from package **JointAI** (with argument `pattern = TRUE`)
* `aggr()` from package **VIM**
* `vis_dat()` and `vis_miss()` from package **visdat**


#### Task
:::{.task}
Explore the missing data pattern of the **EP16dat1** data.
:::

#### Solution (i)
```{r, fig.width = 10, fig.height = 15}
mdp <- mice::md.pattern(EP16dat1)
```
```{r, class.output = "scroll-400"}
mdp
```

The function `md.pattern()` from the package **mice** gives us a matrix where
each row represents one missing data pattern (1 = observed, 0 = missing).
The row names show how many rows of the dataset have the given pattern.
The last column shows the number of missing values in each
pattern, the last row the number of missing values per variable.

`md.pattern()` also plots the missing data pattern automatically.


#### Solution (ii)
```{r fig.width = 7, fig.height = 10, class.output = "scroll-400"}
MDP <- JointAI::md_pattern(EP16dat1, pattern = TRUE)
MDP
```

The function `md_pattern()` from package **JointAI** gives a matrix very
similar to the one obtained from `mice::md.pattern()`.
However, here, the number of rows in the data that have a particular missing
data pattern are given in the last column.

For more information on how to customize the visualization by `md_pattern()`
see the vignette [Visualizing Incomplete Data](https://nerler.github.io/JointAI/articles/VisualizingIncompleteData.html)
on the [webpage of **JointAI**](https://nerler.github.io/JointAI/index.html).


#### Solution (iii)
```{r md-pattern-solution-VIM, fig.width = 8, fig.height = 5}
VIM::aggr(EP16dat1)
```

`aggr()` from **VIM** plots a histogram of the proportion of missing values
per column as well as a visualization of the missing data pattern. Here,
a small histogram on the right edge of the plot shows the proportion of cases
in each pattern.

For more options how to customize the plot, see the 
[**VIM** documentation](https://CRAN.R-project.org/package=VIM).

### Proportion of missing values {.tabset .tabset-fade .tabset-pills}

#### Task
:::{.task}
Now, get an overview of

* how much missingness there is in each variable, and
* the proportion of (in)complete cases.

```{r, echo = FALSE, results = 'asis'}
hint("The function `complete.cases()` can be applied to a `data.frame` and 
will give you a vector of values `TRUE` 
(if the the row does not have missing values) and `FALSE` 
if there are missing values.

`is.na()` returns `TRUE` if the supplied object is missing, `FALSE` if it is 
observed.

`colSums()` calculates the sum of values in each column of a data.frame or matrix

`colMeans()` calculates the mean of values in each column of a data.frame or matrix")
```
:::

#### Solution
```{r misprop-hint-2, solution = TRUE}
# number and proportion of complete cases
cbind(
  "#" = table(ifelse(complete.cases(EP16dat1), 'complete', 'incomplete')),
  "%" = round(100 * table(!complete.cases(EP16dat1))/nrow(EP16dat1), 2)
)

# number and proportion of missing values per variable
cbind("# NA" = sort(colSums(is.na(EP16dat1))),
      "% NA" = round(sort(colMeans(is.na(EP16dat1))) * 100, 2))

```

### Relationship between variables {.tabset .tabset-fade .tabset-pills}
In the missing data pattern we could already see that some variables tend to 
be missing together. But there may be different types of relationships between
variables that are of interest.

#### Task 1
:::{.task}
Our data contains `hgt` (height), `wgt` (weight) and `BMI`. 
Check the missing data pattern specifically for those three variables.
:::

#### Solution 1
There are multiple solutions how you could approach this.

We could obtain the missing data pattern for only the three variables of interest:
```{r heightweight-solution}
JointAI::md_pattern(EP16dat1[, c("hgt", "wgt", "BMI")], pattern = TRUE, plot = FALSE)
```

Alternatively, we could look at a 3-dimensional table of the missing data indicator
for the three variables:
```{r}
with(EP16dat1, 
     table(ifelse(is.na(hgt), 'height mis.', 'height obs.'),
           ifelse(is.na(wgt), 'weight mis.', 'weight obs.'),
           ifelse(is.na(BMI), 'BMI mis.', 'BMI obs.'))
)
```

We could also use a table in long format to describe the missing data pattern:
```{r}
plyr::ddply(EP16dat1, c(height = "ifelse(is.na(hgt), 'missing', 'observed')",
                        weight = "ifelse(is.na(wgt), 'missing', 'observed')",
                        BMI = "ifelse(is.na(BMI), 'missing', 'observed')"),
            plyr::summarize,
            N = length(hgt)
)
```

As we have also seen in the lecture, there are some cases where
only either `hgt` or `wgt` is missing. `BMI` is only observed when both
components are observed. To use all available information, we want to 
impute `hgt` and `wgt` separately and calculate `BMI` from the imputed values.


#### Task 2
:::{.task}
The data contains variables on hypertension (`hypten`) and the use of medication
to treat hypertension (`HyperMed`).
We might expect that medication is only prescribed for patients with hypertension, hence, we need to investigate the relationship between `HyperMed` and `hypten`.

* Make a table of the two variables to confirm our expectation.
* Make sure that missing values are also displayed in that table!

```{r, echo = FALSE, results = 'asis'}
hint("If you use the function `table()`, you need to specify the argument 
     'exclude = NULL'.")
```
:::

#### Solution 2
```{r hyptentab-solution, solution = TRUE}
with(EP16dat1, table(HyperMed, hypten, exclude = NULL))
```

#### Task 3
:::{.task}
Furthermore, we can expect a systematic relationship between the variables
`hypchol` (hypercholesterolemia) and `chol` (serum cholesterol).

* Find out how these two variables are related.
:::

#### Solution 3
```{r hypchol-solution}
with(EP16dat1, plot(chol ~ hypchol))
```

Based on this plot it seems that `hypchol` is `yes` if `chol` is above some
threshold. To be sure, and to find out what this threshold may be, we 
need to look at the numbers:
```{r}
with(EP16dat1, summary(chol[hypchol == "no"]))
with(EP16dat1, summary(chol[hypchol == "yes"]))
```

Indeed, there is no overlap between values of `chol` for participants in the
two groups defined by `hypchol`.
The cut-off for `hypchol` is probably 6.2.


#### Task 4
:::{.task}
Visualize the correlations between variables to detect associations that you
may have missed.
:::

#### Solution 4
```{r corrplot-solution, solution = TRUE, fig.width = 9, fig.height = 9, warning=FALSE}
# convert all variables to numeric
dat_num <- sapply(EP16dat1, as.numeric)

# calculate the Spearman correlation matrix
cormat <- cor(dat_num, use = 'pair', method = 'spearman')

# plot the correlation matrix
corrplot::corrplot(cormat, method = 'square', type = 'lower')
```

:::{.notebox}
[Note:]{.Note}
Correlations involving categorical variables are not usually done!
We only use this as a quick-and-dirty method for visualization of relationships.
:::

The question marks in the plot indicate that no correlation
could be calculated for `cohort` (because `cohort` is the same for all subjects)
and neither between `HyperMed` and `hypten` (but we already knew that from the 
lecture).

Besides the relations we were already aware of, the plot shows us some additional
potentially relevant correlations, for example, that `wgt` is
strongly correlated with `WC`.

Comparing the number of cases where only either `wgt` or `WC` is missing
shows that there are 14 cases where `wgt` is missing but `WC` is observed
and 58 cases where `WC` is missing and `wgt` is observed.

```{r}
with(EP16dat1, table(ifelse(is.na(WC), 'WC mis.', 'WC obs.'),
                     ifelse(is.na(wgt), 'wgt mis.', 'wgt obs.')
))
```


## Distribution of the data {.tabset .tabset-fade .tabset-pills}
Before imputing missing values it is important to take a look at how the 
observed parts of incomplete variables are distributed, so that we can choose
appropriate imputation models.

### Task
:::{.task}
Visualize the distributions of the incomplete continuous and categorical 
variables. The package **JointAI** provides the convenient function `plot_all()`,
but you can of course use functions from other packages as well.


Pay attention to

* whether continuous distributions deviate considerably from the normal distribution,
* if variables have values close to the limits of the range they are defined in,
* whether categorical variables are very unbalanced (i.e., some category very small).
:::

<br>

:::{.notebox}
[Note:]{.Note}
When you use RStudio and the "Plots" panel is too small, you may not be able to
show the distributions of all variables.
You can try to extend the panel, or tell R to plot the graph in a new window, 
by first calling `windows()`.

It can also help to reduce the white space around the separate plots by changing
the margins in the graphical parameters: `par(mar = c(3, 3, 2, 1))`
:::


### Solution
```{r distr-solution, solution = TRUE, fig.width = 9, fig.height = 10.5, dpi=600}
par(mar = c(3, 3, 2, 1), mgp = c(2, 0.6, 0))
JointAI::plot_all(EP16dat1, breaks = 30, ncol = 4)
```

To learn more about additional options of `plot_all()` check out the vignette [Visualizing Incomplete Data](https://nerler.github.io/JointAI/articles/VisualizingIncompleteData.html)
on the [webpage of **JointAI**](https://nerler.github.io/JointAI/index.html).
