---
title: 'Multiple Imputation with the mice Package'
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: style/footer.html
    css: style/style.css
---

```{r, include = FALSE}
projdir <- gsub("/Practicals", "", getwd())

runimp <- FALSE

hint <- function(text) {
  id <- paste0('hint_', sample.int(1e10, size = 1))
    cat(
    paste0('\n',
      '<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" ',
      'data-target="#', id, '">Hint</button>',
      '<div id = "', id, '" class="collapse" style="border:1px; ',
      'border-style:solid; padding: 1em; border-color:#1F78B4">',
      text, '</div>')
  )
}

options(width = 100)

set.seed(2020)

library(kableExtra)
library(knitr)
library(mice)

load(file.path(projdir, "Practicals/data/EP16dat1.RData"))
load(file.path(projdir, "Practicals/workspaces/imps.RData"))
```

## Preface {data-progressive=FALSE}
### R packages

In this practical, a number of R packages are used.
If any of them are not installed you may be able to follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:

* `r R.version.string`
* `mice` (version: `r packageVersion("mice")`)


### Help files
You can find help files for any function by adding a `?` before the name of the 
function.

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


## Dataset 
### Overview {.tabset .tabset-fade .tabset-pills}

For this practical, we will use the **EP16dat1** dataset, which is a 
subset of the [NHANES (National Health and Nutrition Examination Survey)](https://www.cdc.gov/nchs/nhanes/index.htm) data.

To get the **EP16dat1** dataset, load the file `EP16dat1.RData`.
You can download it [here](https://nerler.github.io/EP16_Multiple_Imputation/practical/data/index.html).

To load this dataset into R, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the file on your computer.

If you know the path to the file, you can also use `load("<path>/EP16dat1.RData")`.


If you have not followed the first practical or if you re-loaded the **EP16dat1**
data, you need to re-code the variable `educ` again:

```{r}
EP16dat1$educ <- as.ordered(EP16dat1$educ)
```


## Preparing for imputation
### Set-up run {.tabset .tabset-fade .tabset-pills}
Imputation needs to be tailored to the dataset at hand and, hence, using
the function `mice()` well requires several arguments to be specified.
To make the specification easier it is useful to do a dry-run which will
create the default versions of everything that needs to be specified. 

These default settings can then be adapted to our data.

#### Task
:::{.task}
Do the set-up run of `mice()` with the **EP16dat1** data without any 
iterations (`maxit = 0`).
:::

#### Solution
```{r imp0-solution}
# Note: This command may not produce any output.
library("mice")
imp0 <- mice(EP16dat1, maxit = 0)
```


### Imputation method {.tabset .tabset-fade .tabset-pills}
There are many imputation methods available in **mice**. You can find the list in
the help page of the 
[`mice()` function](https://www.rdocumentation.org/packages/mice/versions/3.14.0/topics/mice).
We will focus here on the following ones:

```{r, echo = FALSE}
tab <- rbind(
  c("pmm", "any", "Predictive mean matching"),
  c("norm", "numeric", "Bayesian linear regression"),
  c("logreg", "binary", "Logistic regression"),
  c("polr", "ordered", "Proportional odds model"),
  c("polyreg", "unordered", "Polytomous logistic regression")
)

colnames(tab) <- c("name", "variable type", "description")

knitr::kable(tab, row.names = FALSE) %>%
  kable_styling(full_width = FALSE)
  
```

The default imputation methods that `mice()` selects can be specified in the
argument `defaultMethod`. 

If unspecified, `mice` will use 

* `pmm` for numerical columns,
* `logreg` for factor columns with two categories, 
* `polyreg` for columns with unordered and 
* `polr` for columns with ordered factors with more than two categories.

#### Task 1
:::{.task}
When a normal imputation model seems to be appropriate for most of the 
continuous covariates, you may want to specify `norm` as
the default method in the set-up run. Let's do that!

```{r results = 'asis', echo = FALSE}
hint('The order of the types of variable is: continuous, binary, factor, ordered factor.')
```
:::

#### Solution 1
```{r imp0v2-hint-2}
imp0 <- mice(EP16dat1, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
```

#### Task 2
:::{.task}
In the histograms we made for the continuous variables during the previous practical,
we could see that the variable `creat` had a skewed distribution, hence, 
using a normal imputation model may not work well. 

* Extract the default settings of the imputation `method` from `imp0`.
* Change the imputation method for `creat` so that this variable will be imputed
  using predictive mean matching.
* Check that all specified imputation methods are correct. When no imputation 
  method is specified (`""`) the variable will not be imputed.
:::

#### Solution 2
```{r changeimpmod-solution}
meth <- imp0$method
meth["creat"] <- "pmm"
meth
```


### Predictor matrix {.tabset .tabset-fade .tabset-pills}
The predictor matrix specifies which variables are used in the linear predictors
of each of the imputation models. 

A value of `1` specifies that the variable given in the column name is used in the model
to impute the variable given in the row name (and `0` specifies that this variable
is not used in that model).

#### Task 1
:::{.task}
Get the `predictorMatrix` from `imp0`. Notice that **mice** has already set
some of the values to `0`. Do you understand why?
:::

#### Solution 1
```{r predmat-solution}
pred <- imp0$predictorMatrix
pred
```

The column corresponding to the variable `cohort` is set to `0` which means 
that this variable is not used in any of the imputation models.
`cohort` has the same value for all observations, so it would not be useful as a covariate.

#### Task 2
:::{.task}
Because BMI is calculated from height (`hgt`) and weight (`wgt`), and there are
cases where only one of these two variables is missing, we want to impute 
`hgt` and `wgt` separately. `BMI` should be imputed using "passive imputation".

To avoid multicollinearity (which may lead to problems during imputation),
imputation models should not include all three variables as predictor variables.
In this example, we will use `BMI` to impute the other incomplete variables in
the data.

Moreover, we need to exclude `WC` from the imputation model for `wgt` because
the high correlation between `WC`, `BMI` and `wgt` would otherwise lead to problems
during imputation.

And since `HyperMed` does not give us a lot more information than `hypten`, but
has a lot more missing values, we do not want to use it as a predictor variable.

Apply the necessary changes to  `pred` and `meth`.
```{r results = 'asis', echo = FALSE}
hint('For passive imputation, you need to specify the formula used to calculate
BMI in meth using `"~I(...)"`.')
```
:::


#### Solution 2
```{r changepred-solution}
# BMI will not be used as predictor of height and weight
pred[c("hgt", "wgt"), "BMI"] <- 0

# height and weight will not be used as predictor in any model
pred[, c("hgt", "wgt")] <- 0

# height and weight will be used as predictors for each other
pred["hgt", "wgt"] <- 1
pred["wgt", "hgt"] <- 1

# WC is not used as predictor for weight
pred["wgt", "WC"] <- 0

# HyperMed will not be used as predictor in any model
pred[, "HyperMed"] <- 0

# hypchol will not be used as predictor in the imputation model for chol
pred["chol", "hypchol"] <- 0

# BMI will be imputed passively
meth["BMI"] <- "~I(wgt/hgt^2)"

# HyperMed will not be imputed
meth["HyperMed"] <- ""
```

### Visit sequence {.tabset .tabset-fade .tabset-pills}
The visit sequence specifies the order in which the variables are imputed.

#### Task
:::{.task}
To be sure that the imputed values of `BMI` match the imputed values of `hgt`
and `wgt`, `BMI` needs to be imputed after `hgt` and `wgt`.

* Get the `visitSequence` from `imp0`, and
* change it if necessary.

:::


#### Solution
```{r changevisseq-solution, solution = TRUE}
visSeq <- imp0$visitSequence

which_BMI <- match("BMI", visSeq)
visSeq <- c(visSeq[-which_BMI], visSeq[which_BMI])
```



## Imputation
### Running the imputation  {.tabset .tabset-fade .tabset-pills}
#### Task
:::{.task}
With the changes that we have made to the `predictorMatrix` and `method`, we
can now perform the imputation. Use `m = 5` and `maxit = 10`.
:::

#### Solution
```{r eval = runimp}
imp <- mice(EP16dat1, method = meth, predictorMatrix = pred, visitSequence = visSeq,
            maxit = 10, m = 5, seed = 2020)
```


`mice()` prints the name of the variable being imputed for each
iteration and imputation. If you run `mice()` on your own computer the output
will show up continuously. There, you may notice that imputation is slowest for 
categorical variables, especially when they have many categories.

You can hide the lengthy output by specifying
`printFlag = FALSE`.

### What does mice return? {.tabset .tabset-fade .tabset-pills}
#### Task
:::{.task}
`mice()` does not return a `data.frame`. 
Find out the class of the object returned by `mice()` function using the 
function `class()`, and take a look at the [help file](https://www.rdocumentation.org/packages/mice/versions/3.14.0/topics/mids-class) for this class.
:::


#### Solution
```{r findclass-solution}
class(imp)
```

We see that `imp` is an object of class `mids`.

The [help file](https://www.rdocumentation.org/packages/mice/versions/3.14.0/topics/mids-class) tells us that a `mids` object is a list with several elements:

```{r, echo = FALSE}
midsdf <- rbind(
  c("`data`:", "Original (incomplete) data set."),
  c("`imp`:",
    paste0("The imputed values: A list of `ncol(data)` components, ",
           "each list component is a matrix with `nmis[j]` rows and `m` columns.")),
  c("`m`:",
    "The number of imputations."),
  c("`where`:", "The missingness indicator matrix."),
 c("`blocks`", "The blocks argument of the `mice()` function."),  
  c("`call`:", "The call that created the `mids` object."),
  c("`nmis`:",
    "The number of missing observations per variable."),
  c("`method`:",
    "The vector imputation methods."),
  c("`predictorMatrix`:",
    "The predictor matrix."),
  c("`visitSequence`:", 
    "The sequence in which columns are visited during imputation."),
 c("`formulas`", "A named list of formulas corresponding the the imputed variables (blocks)."),
  c("`post`:",
    "A vector of strings of length `length(blocks)` with commands for post-processing."),
 # c("`blots`", ""),
  c("`seed`:",
    "The seed value of the solution."),
  c("`iteration`:", 
    "The number of iterations."),
  c("`lastSeedValue`:",
    "The most recent seed value."),
  c("`chainMean:`", 
    paste0("The mean of imputed values per variable and iteration: ",
           "a list of `m` components. Each component is a matrix with `maxit`",
           "columns and `length(visitSequence)` rows.")),
  c("`chainVar`:",
    paste0("The variances of imputed values per variable and iteration" ,
           "(same structure as `chainMean`).")),
  c("`loggedEvents`:",
    paste0("A `data.frame` with the record of automatic corrective actions and warnings; ",
           "(`NULL` if no action was made).")),
  # c("`pad`:",
  #   paste0("A list containing the internally used version of the `data`,",
  #          "`method`, `predictorMatrix`, `visitSequence`, `post` and dummy coding.")
 c("`version`", "Version number of the **mice** package that created the object."),
 c("`date`", "Date at which the object was created.")
) %>% as.data.frame

names(midsdf) <- c(" ", " ")

midsdf %>% kable(format = 'html') %>%
  kable_styling()
  # row_spec(c(4:10, 12:16), background = "#f2f2f2")
```


**Details of the `loggedEvents`:**

`mice()` does some pre-processing of the data:

  * variables containing missing values, that are not imputed but used as
    predictor are removed
  * constant variables are removed
  * collinear variables are removed
  
  
Furthermore, during each iteration

  * variables (or dummy variables) that are linearly dependent are removed
  * `polr` imputation that does not converge is replaced by `polyreg`. 
  
The `data.frame` in `loggedEvents` has the following columns:
```{r, echo = FALSE}
rbind(
  c("`it`", "iteration number"), 
  c("`im`", "imputation number"),
  c("`dep`", "name of the name of the variable being imputed"),
  c("`meth`", "imputation method used"),
  c("`out`", "character vector with names of altered/removed predictors")
) %>%
  kable(format = "html") %>%
  kable_styling(full_width = FALSE)
```


## A Note on Imputation Methods
In Section 4 of the lectures, we have seen different methods to obtain imputed 
values: Bayesian imputation, bootstrap imputation and predictive mean matching.

The imputation methods implemented in the **mice** that we talked about in 
Section 6 of the lectures and here of course fall into these categories.

`norm` and `logreg` perform Bayesian linear and logistic regression, respectively.
There are bootstrap alternatives available: `norm.boot` and `logreg.boot`.

The default method `pmm` performs predictive mean matching using, by default,
5 donors and type-1 matching.

The number of donors can be changed using the argument `donors`. A selection
between type-0, type-1 and type-2 matching is possible via the argument `matchtype`.

For example, we could adjust the syntax from above:
```{r, eval = FALSE}
mice(data = EP16dat1, m = 5, method = meth, predictorMatrix = pred, 
     visitSequence = visSeq, maxit = 10, seed = 2020,
     donors = 5, matchtype = 2L)
```

Both arguments can be passed to `mice()` and will apply to all imputation models
that use `pmm`.


An alternative implementation of predictive mean matching is the imputation method
`midastouch`. It uses type-2 matching with a leave-one-out approach,
where parameter estimates are obtained using bootstrap.
Donors are selected from all cases for which the variable that is to be imputed
is observed, with probability depending on the distance of the predicted values.



```{r saveimp, include = FALSE, eval = runimp}
imp30 <- update(imp, maxit = 30)
impnaive <- mice(EP16dat1, maxit = 30, m = 5, seed = 2020)

save(imp, imp30, impnaive, 
     file = file.path(projdir, "Practicals/workspaces/imps.RData"))
```
