---
title: "Multiple imputation with the mice package"
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    pandoc_args: [
      "--output=Practical_MImice.html"
      ]
#   learnr::tutorial:
#     progressive: true
#     allow_skip: true
# runtime: shiny_prerendered
---

```{r, include = F}
if (names(rmarkdown::metadata$output) == "learnr::tutorial") {
  library(learnr)
  learnr::initialize_tutorial()
  static <- FALSE
}

if (names(rmarkdown::metadata$output) == "html_document") {
  knitr::opts_hooks$set(eval = function(opt) {
    if (any(opt$exercise))
      opt$eval <- opt$include <- FALSE
    
    opt
  })
  
  static <- TRUE
  
  options(width = 100)
}
```


```{r packages, include = FALSE}
# library(ggplot2)
# library(JointAI)
library(kableExtra)
library(knitr)
# library(miceadds)
# library(mitools)
# library(magrittr)
# library(nlme)
# library(plyr)
# library(RColorBrewer)
# library(reshape2)
# library(VIM)
# library(visdat)
```


```{r load_data, context="data", include=FALSE}
load("www/NHANES_for_practicals.RData")
NHANES$educ <- as.ordered(NHANES$educ)
```


## Preface {data-progressive=FALSE}
### R packages

```{r, eval = static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
If any of them are not installed you can still follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:')
```
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
    The names and version numbers of the packages available on this platform are:')
```

* `mice` (version: `r packageVersion("mice")`)
* `JointAI` (version: `r packageVersion("JointAI")`)



### Help files
You can find help files for any function by adding a `?` before the name of the 
function.

```{r, eval = !static, results = 'asis', echo = FALSE}
cat("The files might look a bit funny when they are displayed as R output.")
```

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


## Dataset 

For this practical, we will again use the **NHANES** dataset that we have seen in
the previous practical.
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat("In the interactive version of this practical, the data is already loaded
    and the variable `educ` correctly coded as an ordered factor.")
```
```{r, eval = static, echo = FALSE, results = 'asis'}
cat(
'To load this dataset, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the file
`NHANES_for_practicals.RData` on your computer.
If you know the path to the file, you can also use `load("<path>/NHANES_for_practicals.RData")`.

If you have not followed the first practical or if you re-loaded the NHANES data
you need to re-code the variable `educ` again:'
)
```
```{r, eval = FALSE, echo = static}
NHANES$educ <- as.ordered(NHANES$educ)
```


## Preparing for imputation
### Set-up run {.tabset .tabset-fade .tabset-pills}
Imputation needs to be tailored to the dataset at hand and, hence, using
the function `mice()` well requires several arguments to be specified.
To make the specification easier it is useful to do a dry-run which will
create the default versions of everything that needs to be specified. 

These default settings can then be modified to what is necessary for our data.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Do the set-up run of `mice()` with the NHANES data without any iterations (`maxit = 0`).

```{r imp0, exercise = TRUE, exercise.setup = "educrecode"}
imp0 <- 
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r imp0-solution, solution = TRUE}
# Note: This command will not produce any output.
library(mice)
imp0 <- mice(NHANES, maxit = 0)
```

### Imputation method {.tabset .tabset-fade .tabset-pills}
There are many imputation methods available in **mice**. You can find the list in
the help page of the 
[`mice()` function](https://www.rdocumentation.org/packages/mice/versions/3.3.0/topics/mice).
We will focus here on the following ones:

```{r, echo = F}
tab <- rbind(
  c("pmm", "any", "Predictive mean matching"),
  # c("midastouch", "any", "Weighted predictive mean matching"),
  # c("sample", "any", "Random sample from observed values"),
  # c("cart", "any", "Classification and regression trees"),
  # c("rf", "any", "Random forest imputations"),
  # c("mean", "numeric", "Unconditional mean imputation"),
  c("norm", "numeric", "Bayesian linear regression"),
  # c("norm.nob", "numeric", "Linear regression ignoring model error"),
  # c("norm.boot", "numeric", "Linear regression using bootstrap"),
  # c("norm.predict", "numeric", "Linear regression, predicted values"),
  # c("quadratic", "numeric", "Imputation of quadratic terms"),
  # c("ri", "numeric", "Random indicator for nonignorable data"),
  c("logreg", "binary", "Logistic regression"),
  # c("logreg.boot", "binary", "Logistic regression with bootstrap"),
  c("polr", "ordered", "Proportional odds model"),
  c("polyreg", "unordered", "Polytomous logistic regression")
  # c("lda", "unordered", "Linear discriminant analysis"),
  # c("2l.norm", "numeric", "Level-1 normal heteroscedastic"),
  # c("2l.lmer", "numeric", "Level-1 normal homoscedastic, lmer"),
  # c("2l.pan", "numeric", "Level-1 normal homoscedastic, pan"),
  # c("2l.bin", "binary", "Level-1 logistic, glmer"),
  # c("2lonly.mean", "numeric", "Level-2 class mean"),
  # c("2lonly.norm", "numeric", "Level-2 class normal"),
  # c("2lonly.pmm", "any", "Level-2 class predictive mean matching")
)

colnames(tab) <- c("name", "variable type", "description")

knitr::kable(tab, row.names = FALSE) %>%
  kable_styling(full_width = FALSE)
  
```

The default imputation methods that `mice()` selects can be specified in the
argument `defaultMethod`. 

If unspecified, `mice` will use 

* `pmm` for numerical columns,
* `logreg` for factor columns with two categories, 
* `polyreg` for columns with unordered and 
* `polr` for columns with ordered factors with more than two categories.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

When a normal imputation model seems to be appropriate for most of the 
continuous covariates, you may want to specify `norm` as
the default method in the setup run. Let's do that:

```{r imp0v2, exercise = TRUE, exercise.setup = "educrecode"}
imp0 <- ...(..., defaultMethod = ...)
```

```{r imp0v2-hint-1, eval = !static, echo = FALSE}
# The order for the types of variables is:
# continuous, binary, factor, ordered factor
```

```{r imp0v2_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#imp0v2">Hint</button>
<div id = "imp0v2" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
The order of the types of variable is: continuous, binary, factor, ordered factor.
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```


```{r imp0v2-hint-2, solution = TRUE}
imp0 <- mice(NHANES, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
```

```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

In the histograms we made for the continuous variables during the previous practical,
we could see that the variable `creat` had a skewed distribution, hence, 
using a normal imputation model may not work well. 

* Extract the default settings of `meth` from `imp0`.
* Change the imputation method for `creat` so that this variable will be imputed
  using predictive mean matching.
* Check that all specified imputation methods are correct. When no imputation 
  method is specified (`""`) the variable will not be imputed.

```{r changeimpmod-setup, exercise.setup = "educrecode", eval = !static, echo = FALSE}
imp0 <- mice(NHANES, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
```

```{r changeimpmod, exercise = TRUE}
meth <- ...

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r changeimpmod-solution, solution = TRUE}
meth <- imp0$meth
meth["creat"] <- "pmm"
meth
```


### Predictor matrix {.tabset .tabset-fade .tabset-pills}
The predictor matrix specifies which variables are used in the linear predictors
of each of the imputation models. 

A value of `1` specifies that the variable given in the column name is used in the model
to impute the variable given in the row name (and `0` specifies that this variable
is not used in that model).

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Get the `predictorMatrix` from `imp0`. Notice that **mice** has already set
some of the values to `0`. Do you understand why?

```{r predmat-setup, eval = !static, echo = FALSE}
imp0 <- mice(NHANES, maxit = 0,
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
meth <- imp0$meth
meth["creat"] <- "pmm"
```

```{r predmat, exercise = TRUE}
pred <-
pred
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

```{r predmat-solution, solution = TRUE}
pred <- imp0$predictorMatrix
pred
```

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

The column corresponding to the variable `cohort` is set to `0` which means 
that this variable is not used in any of the imputation models.
`cohort` has the same value for all observations, so it would not be useful as a covariate.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Because BMI is calculated from height (`hgt`) and weight (`wgt`), and there are
cases where only one of these two variables is missing, we want to impute 
`hgt` and `wgt` separately. `BMI` should be imputed using "passive imputation".

To avoid multicollinearity (which may lead to problems during imputation),
imputation models should not include all three variables as predictor variables.
In this example, we will use `BMI` to impute the other variables.

And since `HyperMed` does not give us a lot more information than `hypten`, but
has a lot more missing values, we do not want to use it as a predictor variable.

Apply the necessary changes to  `pred` and `meth`.
```{r changepred-setup, include = FALSE}
NHANES$educ <- as.ordered(NHANES$educ)
imp0 <- mice(NHANES, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
meth <- imp0$meth
meth["creat"] <- "pmm"

pred <- imp0$pred
```

```{r changepred, exercise = TRUE}

```

```{r changepred-hint-1, eval = !static, echo = FALSE}
# For passive imputation, you need to specify the formula used to calculate
# BMI in meth using "~I(...)".
```

```{r changepred_hint1, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#changepred">Hint</button>
<div id = "changepred" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
For passive imputation, you need to specify the formula used to calculate BMI in meth using `"~I(...)"`.
</div>')
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r changepred-solution, solution = TRUE}
# BMI will not be used as predictor of height and weight
pred[c("hgt", "wgt"), "BMI"] <- 0
# height and weight will not be used as predictor in any model
pred[, c("hgt", "wgt")] <- 0
# height and weight will be used as predictors for each other
pred["hgt", "wgt"] <- 1
pred["wgt", "hgt"] <- 1

# HyperMed will not be used as predictor in any model
pred[, "HyperMed"] <- 0

# hypchol will not be used as predictor in the imputation model for chol
pred["chol", "hypchol"] <- 0

# BMI will be imputed passively
meth["BMI"] <- "~I(wgt/hgt^2)"
# HyperMed will not be imputed
meth["HyperMed"] <- ""
```

### Visit sequence {.tabset .tabset-fade .tabset-pills}
The visit sequence specifies the order in which the variables are imputed.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```


To be sure that the imputed values of `BMI` match the imputed values of `hgt`
and `wgt`, `BMI` needs to be imputed after `hgt` and `wgt`.

* Get the `visitSequence` from `imp0`, and
* change it if necessary.

```{r changevisseq-setup, include = FALSE}
NHANES$educ <- as.ordered(NHANES$educ)
imp0 <- mice(NHANES, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
meth <- imp0$meth
meth["creat"] <- "pmm"

pred <- imp0$pred
pred[c("hgt", "wgt"), "BMI"] <- 0
pred[, c("hgt", "wgt")] <- 0
pred["hgt", "wgt"] <- 1
pred["wgt", "hgt"] <- 1

pred[, "HyperMed"] <- 0

pred["chol", "hypchol"] <- 0


meth["BMI"] <- "~I(wgt/hgt^2)"
meth["HyperMed"] <- ""
```

```{r changevisseq, exercise = TRUE}
visSeq <- ...
...
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r changevisseq-solution, solution = TRUE}
visSeq <- imp0$visitSequence

# the lazy version:
visSeq <- c(visSeq[-2], visSeq[2])

# the save version:
# To make sure you still do the correct thing if the order of the columns in 
# NHANES would change
which_BMI <- match("BMI", visSeq)
visSeq <- c(visSeq[-which_BMI], visSeq[which_BMI])
```


## Running the imputation  {.tabset .tabset-fade .tabset-pills}
```{r allprep, include = FALSE}
NHANES$educ <- as.ordered(NHANES$educ)
imp0 <- mice(NHANES, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
meth <- imp0$meth
meth["creat"] <- "pmm"
meth["HyperMed"] <- ""

pred <- imp0$pred

pred[c("hgt", "wgt"), "BMI"] <- 0
pred[, c("hgt", "wgt")] <- 0
pred["hgt", "wgt"] <- 1
pred["wgt", "hgt"] <- 1

pred[, "HyperMed"] <- 0

pred["chol", "hypchol"] <- 0

meth["BMI"] <- "~I(wgt/hgt^2)"
meth["HyperMed"] <- ""

visSeq <- imp0$visitSequence
which_BMI <- match("BMI", visSeq)
visSeq <- c(visSeq[-which_BMI], visSeq[which_BMI])
```

```{r, eval = static, echo = FALSE}
asis_output("### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```



With the changes that we have made to the `predictorMatrix` and `method`, we
can now perform the imputation. Use `m = 5` and `maxit = 10`.

```{r runimp, exercise = TRUE, exercise.timelimit = 100, exercise.setup = "allprep"}
imp <- 
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution\\n")
```

```{r runimp-solution, solution = TRUE, eval = FALSE}
imp <- mice(NHANES, method = meth, predictorMatrix = pred, visitSequence = visSeq,
            maxit = 10, m = 5,
            seed = 2019)
```

```{r saveimp, include = FALSE, eval = static, cache = TRUE}
savedimps_imp <- mice(NHANES, method = meth, predictorMatrix = pred,
                      visitSequence = visSeq,
                      maxit = 10, m = 5,
                      seed = 2019)
savedimps_impnaive <- mice(NHANES, maxit = 10, m = 5, seed = 2019)

save(savedimps_imp,
     savedimps_impnaive, file = "www/imps.RData")
```

`mice()` prints the name of the variable being imputed for each
iteration and imputation. If you run `mice()` on your own computer the output
will show up continuously. There, you may notice that imputation is slowest for 
categorical variables, especially when they have many categories.

You can hide the lengthy output by specifying
`printFlag = FALSE`.



## Investigate the imputation
### Type of class {.tabset .tabset-fade .tabset-pills}

```{r, eval = static, echo = FALSE}
asis_output("#### Task \\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

`mice()` does not return a `data.frame`. 
Find out the class of the object returned by `mice()` function using the 
function `class()`, and take a look at the [help file](https://www.rdocumentation.org/packages/mice/versions/2.46.0/topics/mids-class) for this class:

```{r allimp, echo = FALSE}
NHANES$educ <- as.ordered(NHANES$educ)
imp0 <- mice(NHANES, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
meth <- imp0$meth
meth["creat"] <- "pmm"
meth["HyperMed"] <- ""

pred <- imp0$pred

pred[c("hgt", "wgt"), "BMI"] <- 0
pred[, c("hgt", "wgt")] <- 0
pred["hgt", "wgt"] <- 1
pred["wgt", "hgt"] <- 1

pred[, "HyperMed"] <- 0

pred["chol", "hypchol"] <- 0

meth["BMI"] <- "~I(wgt/hgt^2)"
meth["HyperMed"] <- ""

visSeq <- imp0$visitSequence
which_BMI <- match("BMI", visSeq)
visSeq <- c(visSeq[-which_BMI], visSeq[which_BMI])

impnaive <- savedimps_impnaive
imp <- savedimps_imp
```


```{r findclass, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r findclass-solution, solution = TRUE}
class(imp)
```



## Elements of the `mids` object {.tabset .tabset-fade .tabset-pills}
We see that `imp` is an object of class `mids`.

The [help file](https://www.rdocumentation.org/packages/mice/versions/2.46.0/topics/mids-class) tells us that a `mids` object is a list with several elements:

```{r, echo = FALSE}
midsdf <- rbind(
  c("`call`:", "The call that created the `mids` object."),
  c("`data`:", "A copy of the incomplete data set."),
  c("`where`:", "The missingness indicator matrix."),
  c("`m`:",
    "The number of imputations."),
  c("`nmis`:",
    "The number of missing observations per variable."),
  c("`imp`:",
    paste0("The imputed values: A list of `ncol(data)` components,",
           "each list component is a matrix with `nmis[j]` rows and `m` columns.")),
  c("`method`:",
    "The vector imputation methods."),
  c("`predictorMatrix`:",
    "The predictor matrix."),
  c("`visitSequence`:", 
    "The sequence in which columns are visited during imputation."),
  c("`post`:",
    "A vector of strings of length `ncol(data)` with commands for post-processing."),
  c("`seed`:",
    "The seed value of the solution."),
  c("`iteration`:", 
    "The number of iterations."),
  c("`lastSeedValue`:",
    "The most recent seed value."),
  c("`chainMean:`", 
    paste0("The mean of imputed values per variable and iteration: ",
           "a list of `m` components. Each component is a matrix with `maxit`",
           "columns and `length(visitSequence)` rows.")),
  c("`chainVar`:",
    paste0("The variances of imputed values per variable and iteration" ,
           "(same structure as `chainMean`).")),
  c("`loggedEvents`:",
    paste0("A matrix with the record of automatic removal actions (details below); ",
           "(`NULL` if no action was made).")),
  c("`pad`:",
    paste0("A list containing the internally used version of the `data`,",
           "`method`, `predictorMatrix`, `visitSequence`, `post` and dummy coding.")
  )
) %>% as.data.frame

names(midsdf) <- c(" ", " ")

midsdf %>% kable(format = 'html') %>%
  kable_styling()
  # row_spec(c(4:10, 12:16), background = "#f2f2f2")
```


**Details of the `loggedEvents`:**

`mice()` does some pre-processing of the data:

  * variables containing missing values, that are not imputed but used as
    predictor are removed
  * constant variables are removed
  * collinear variables are removed
  
  
Furthermore, during each iteration

  * variables that are linearly dependent are removed
  * `polr` imputation that does not converge is replaced by `polyreg`. 
  
The matrix in `loggedEvents` has the following columns:
```{r, echo = FALSE}
rbind(
  c("`it`", "iteration number"), 
  c("`im`", "imputation number"),
  c("`co`", "column number in the data"),
  c("`dep`", "name of the name of the variable being imputed"),
  c("`meth`", "imputation method used"),
  c("`out`", "character vector with names of altered/removed predictors")
) %>%
  kable(format = "html") %>%
  kable_styling()

```


```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

It is good practice to make sure that `mice()` has not done 
any processing of the data that was not planned or that you are not aware of.
This means checking the `loggedEvents`, but also ensuring that the correct
`method`, `predictorMatrix` and `visitSequence` were used.

Do these checks for `imp`.

```{r checkmids, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r checkmids-solution, solution = TRUE}
imp$method
imp$predictorMatrix
imp$visitSequence
imp$loggedEvents
```

### Logged events {.tabset .tabset-fade .tabset-pills}
Checking the `loggedEvents` we see that for `alc`, `educ` and `smoke` the method
`multinom` was used. That is a bit confusing: we specified `polr` in our
imputation method and `imp$method` is indeed `polr` for the three ordinal variables.

The `mice` help page told us that in case `polr` does not converge `polyreg` will
be used. 

```{r, eval = static, echo = FALSE}
asis_output("#### Task 1\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Find out if there is more information in the
[help](https://www.rdocumentation.org/packages/mice/versions/2.46.0/topics/mice.impute.polr)
[pages](https://www.rdocumentation.org/packages/mice/versions/2.46.0/topics/mice.impute.polyreg)
for those two functions.

```{r checkpolr, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, echo = F, results = 'asis'}
if (static) {
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#checkpolr">Hint</button>
<div id = "checkpolr" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">')
} else {
  cat('<div id = "checkpolr-hint">')
}
```

Using `polr` and `polyreg` when specifying the `method` argument in `mice`
actually calls the functions `mice.impute.polr` and  `mice.impute.polyreg`
when running the imputation.

</div>

```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 1\\n")
```

The help for `mice.impute.polr()` tells us:

> The call to `polr` might fail, usually because the data are very sparse. 
> In that case, `multinom` is tried as a fallback, and a record is written to 
> the `loggedEvents` component of the `mids` object.

and the help for `mice.impute.polyreg()` says:

> The algorithm of `mice.impute.polyreg` uses the function `multinom()`  from 
> the **nnet** package.

So that explains it, everything seems to be in order, but unfortunately the
ordinal models for `alc`, `educ` and `smoke` did not converge and multinomial models
had to be used instead.

```{r, eval = static, echo = FALSE}
asis_output("#### Task 2\\n")
```

Let's see what else would have come up if we had not prepared the
`predictorMatrix`, `method` and `visitSequence` before imputation.
The object `impnaive` contains the result of 
```{r, eval = FALSE, echo = TRUE, include = TRUE}
impnaive <- mice(NHANES, m = 5, maxit = 10)
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Take a look at the `loggedEvents` of `impnaive`. 

```{r impnaiveloggedEvents, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 2\\n")
```

```{r impnaiveloggedEvents-solution, solution = TRUE}
impnaive$loggedEvents
```


```{r, eval = !static, echo = FALSE}
asis_output("### \\n")
```


The `loggedEvents` of the "naive" imputation show that the constant variable
`cohort` was excluded before the imputation (as it should be).

Furthermore, in the imputation model for `HyperMed`, the variable `hypten.1`
was excluded. `hypten.1` is the first dummy variable belonging to `hypten`.
We can check this by looking at `impnaive$pad$categories`:

```{r, eval = FALSE}
impnaive$pad$categories
```

```{r, echo = FALSE}
impnaive <- savedimps_impnaive
impnaive$pad$categories
```

Here, we see in row `r match("hypten", rownames(impnaive$pad$categories))`
that `hypten` has 1 dummy variable and that `hypten.1` 
(row `r match("hypten.1", rownames(impnaive$pad$categories))`) is a dummy variable, belonging to column
`r impnaive$pad$categories$father[match("hypten.1", rownames(impnaive$pad$categories))]`
in the dataset.
Hence, `hypten` was not used in the imputation of `HyperMed`.

<br>

```{r, eval = static, echo = FALSE}
asis_output("#### Task 3\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

We also did not change the `visitSequence` in `impnaive`. Find out what
the effects of that are.

```{r impnaiveVisSeq, exercise = TRUE, exercise.setup = "allimp"}

```

```{r impnaiveVisSeq-hint-1, include = !static, echo = FALSE}
# You can get an imputed datasets from a mids object using the function complete().
# Use ?complete for more information.
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution 3\\n")
```

```{r, impnaiveVisSeq-solution, solution = TRUE}
naiveDF1 <- complete(impnaive, 1)
naivecalcBMI <- with(naiveDF1, wgt/hgt^2)

impDF1 <- complete(imp, 1)
impcalcBMI <- with(impDF1, wgt/hgt^2)

cbind(naiveBMI = naiveDF1$BMI, naivecalcBMI,
      impBMI = impDF1$BMI, impcalcBMI)[which(is.na(NHANES$BMI)), ]

```

When we compare the imputed and calculated values of `BMI` from `impnaive` we
can see that the imputed `hgt` and `wgt` give a different `BMI` than is
imputed. This is because `BMI` is imputed before `wgt`, which means that the
most recent imputed value of `wgt` is from the previous iteration.

Changing the `visitSequence` in `imp` prevented this inconsistency.





### Convergence {.tabset .tabset-fade .tabset-pills}
The mean and variance of the imputed values per iteration and variable is
stored in the elements `chainMean` and `chainVar` of the `mids` object.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Let's plot them to see if our imputation `imp` has converged:

```{r convergence, exercise = TRUE, exercise.setup = "allimp", fig.width = 9, fig.height = 7}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r convergence-solution, solution = TRUE, fig.width = 9, fig.height = 7}
# implemented plotting function (use layout to change the number of rows and columns)
plot(imp, layout = c(6, 6))

# plot for a single variable
matplot(imp$chainMean["wgt", , ], type = "l", ylab = "imputed value",
        xlab = "iteration", main = "wgt")

# or the ggplot version
library(ggplot2)
library(reshape2)
ggplot(melt(imp$chainMean), aes(x = Var2, y = value, color = Var3)) +
  geom_line() +
  facet_wrap("Var1", scales = 'free') +
  theme(legend.position = 'none') +
  xlab("iteration") +
  ylab("imputed value")

```

```{r, eval = !static, echo = FALSE}
asis_output("###\\n")
```

The chains in `imp` seem to have converged, but in practice, more iterations 
should be done to confirm this.

```{r, eval = static, echo = FALSE}
asis_output("#### Continue\\n")
```

In comparison, `impnaive` had some convergence problems:
```{r, fig.width = 9, fig.height = 7}
plot(impnaive, layout = c(6, 6))
```

`hgt`, and `BMI` show a clear trend and the chains do not mix well, i.e., there
is more variation between the chains than within each chain (the same is the 
case for `wgt`). Moreover, `hyperchol` shows some undesired behaviour, where
four chains move upwards and the 5th chain separates from them after half of the
iterations. These are clear signs that there is correlation or identification
problems between these variables and some other variables (which is why we made
adjustments to the `predictorMatrix` for `imp`).

### Imputed values {.tabset .tabset-fade .tabset-pills}
Now that we know that `imp` has converged, we can compare the distribution of
the imputed values against the distribution of the observed values.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Plot the distributions of the imputed variables (continuous and categorical),
and make sure the imputed values are realistic (e.g., height of 2.50m or weight of 10kg for adults).
If there are differences in the distributions of observed and imputed values see if
those differences can be explained by other variables.

```{r distrplot, exercise = TRUE, exercise.setup = "allimp", fig.width = 9, fig.height = 7}

```

```{r impdistrhint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#impdistrhint">Hint</button>
<div id = "impdistrhint" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
You can use `densityplot()` and `probplot()` to get plots for all continuous and
categorical variables.

You can obtain `probplot()` from [https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab](https://gist.github.com/NErler/0d00375da460dd33839b98faeee2fdab).
 
To check all imputed values you can either get a summary of the `imp` element
of the `mids` object or create a complete dataset containing all imputations
using the function `complete()` and get the summary of that.
</div>')
```

```{r distrplot-hint-1, include = !static, echo = FALSE}
# You can use "densityplot()" and "probplot()" to get plots for all continuous and
# categorical variables.
# 
# To check all imputed values you can either get a summary of the "imp" element
# of the mids object or create a complete dataset containing all imputations
# using the function "complete()" and get the summary of that.
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r distrplot-solution, solution = TRUE, fig.width = 9, fig.height = 7, cache = TRUE, eval = F}
densityplot(imp)

# create a long dataset with all imputed data sets stacked onto each other
longDF <- complete(imp, "long")
summary(longDF)

# get the summary of the "raw" imputed values
lapply(imp$imp, function(x) 
  summary(unlist(x))
)

# look into the distribution of SBP and height
densityplot(imp, ~SBP|hypten)
densityplot(imp, ~hgt|gender)

# plot for all categorical variables
probplot(imp)

# look into the categorical variables where proportions of imputed and observed differ
probplot(imp, smoke ~ alc)
probplot(imp, alc ~ gender)
```

```{r, eval = !static, echo = FALSE}
asis_output("###\\n")
```

Unfortunately, we have some negative imputed values for `bili`. Often, this
would not result in bias in the analysis, but may be difficult to explain
when providing a summary of the imputed data in a publication.
In the present example we can see that the observed values have a slightly
right-skewed distribution, compared to the imputed values. Re-doing the
imputation with `pmm` instead of `norm` for `bili` should fix this.
(However, since the imputations seem fine overall, and there is little knowledge
gain in re-doing the previous steps, we will skip this repetition 
in this practical.)

The distributions of the imputed values for `hgt` and `SBP` differ a bit
from the distributions of the observed data, but these differences can be explained
by adjusting for `gender` and `hypten`.

We also imputed a larger proportion than might have been expected in the 
highest category of `alc`, and the distribution of values for `educ` and `smoke`
looks a bit weird.

The difference in distribution of `alc` could be explained by `gender`.
`smoke` and `educ` each have only one missing value, which makes it difficult
to judge the distribution of imputed values. Adjusting the distributions
of observed `smoke` for `alc` shows that the subject with missing
smoking status was a drinker, and drinkers were more often in the `current`
group of smoking.

