---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo = F, message = F}
library(ggplot2)
library(DT)
library(shiny)
library(shinyWidgets)
library(ggiraph)
```

## Topic 1

### Exercise 


```{r, data}
p <- 2
N <- 100
set.seed(2018)
s <- rnorm(p*(p - 1)/2, 0, 0.2)
S <- diag(rgamma(p, 1, 2))
S[lower.tri(S)] <- s
S[upper.tri(S)] <- t(S)[upper.tri(S)]
S <- Matrix::nearPD(S)$mat

normDF <- data.frame(MASS::mvrnorm(N, rep(0, p), S))
names(normDF) <- c("z", "x")

# delete 5 values
normDF$x[order(normDF$z)[seq(from = 2, to = N - 2, length = 5)]] <- NA

lm1 <- lm(x~z, data = normDF)
normDF$pred <- predict(lm1, newdata = normDF)

dist <- sapply(1:sum(is.na(normDF$x)), function(i) {
               normDF$pred - normDF$pred[which(is.na(normDF$x))[i]]
})

dist[which(is.na(normDF$x)), ] <- NA

distorder <- sapply(1:sum(is.na(normDF$x)), function(i) {
                   order(order(abs(dist[, i])))
})

colnames(distorder) <- paste0("nn", 1:sum(is.na(normDF$x)))

normDF <- cbind(normDF, distorder)
```


```{r}
fluidRow(
  column(8, plotOutput("normDFplot")),
  column(4, dataTableOutput("normDFtab"),
         br(),
         checkboxInput("addloc", "show z-location of missing x values", value = 0)
  )
)
```

```{r, context = "server", warning = F}
output$normDFplot <- renderPlot({
  p <- ggplot(normDF, aes(x = z, y = x)) +
    geom_point(na.rm = TRUE) +
    theme(axis.text = element_text(size = 10),
          axis.title = element_text(size = 12, face = 'bold'))
  if (input$addloc == 1) {
    p <- p + geom_vline(xintercept = normDF$z[is.na(normDF$x)], lty = 2,
                        col = 'firebrick2')
  }
  if (input$addpred) {
    p <- p + geom_point(aes(y = pred, col = is.na(x)), size = 1.5) +
      theme(legend.position = 'none') +
      scale_color_manual(name = "x missing",
                         limits = c(F, T),
                         values = c("chartreuse4", 'firebrick2'))
  }
  p
})

output$normDFtab <- renderDataTable({
  datatable(normDF[, c("z", "x")], extensions = "Scroller",
          options = list(dom = 't',
                         deferRender = TRUE,
                         scrollY = 200,
                         scroller = TRUE)) %>%
  formatRound(1:2, 2)
})
```

In predictive mean mathcin, the predicted value for the observed values
$x_{obs}$ as  well as for the missing values $x_{mis}$ is calculated:
\begin{eqnarray*}
\hat x_{obs} = \hat\beta_0 + \hat\beta_1 z_{obs}\\
\hat x_{mis} = \hat\beta_0 + \hat\beta_1 z_{mis}\\
\end{eqnarray*}

```{r}
# checkboxInput("addpred", "show predicted values", value = 0)
materialSwitch(inputId = "addpred", 
               label = "show the predicted values in the plot above",
               status = "primary", right = TRUE)
```


Since we have only included one covariate in our simple example, the predicted 
values all lie on the regression line.


For each of the predicted values of a missing $x$, a pool of candidate values
to impute from is needed. This pool of candidates is chosen based on the 
distance of the predicted values of the observed cases from the predicted
values of the incomplete case.

Let's look at an examle

```{r, context = "server"}
output$plotneighbors <- renderggiraph({
  gg <- ggplot(normDF, aes(x = pred, y = x)) +
    # geom_point(na.rm = T) +
    geom_point_interactive(aes(tooltip = paste("Neighbor number", nn1)),
                           na.rm = T) +
    xlab("predicted value of x") +
    ylab('observed value of x') +
    geom_vline(xintercept = normDF$pred[which(is.na(normDF$x))[1]], lty = 2,
               color = "firebrick2")
  
  ggiraph(code = print(gg), zoom_max = 5)
})
```

```{r}
ggiraphOutput("plotneighbors")
```

