---
title: "How to Analyse Data after Multiple Imputation"
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footer.html
    css: ../style.css
#   learnr::tutorial:
#     progressive: true
#     allow_skip: true
# runtime: shiny_prerendered
---

```{r, include = F}
if (names(rmarkdown::metadata$output) == "learnr::tutorial") {
  library(learnr)
  learnr::initialize_tutorial()
  static <- FALSE
}

if (names(rmarkdown::metadata$output) == "html_document") {
  knitr::opts_hooks$set(eval = function(opt) {
    if (any(opt$exercise))
      opt$eval <- opt$include <- FALSE
    
    opt
  })
  
  static <- TRUE
  
  options(width = 100)
}
```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
# library(JointAI)
# library(reshape2)
# library(plyr)
# library(RColorBrewer)
# library(ggplot2)
```


```{r load_data, context="data", include=FALSE}
load("www/NHANES_for_practicals.RData")
NHANES$educ <- as.ordered(NHANES$educ)

# source("www/propplot.R")
load("www/imps.RData")

imp <- savedimps_imp

library(mice)
```


## Preface {data-progressive=FALSE}
### R packages

```{r, eval = static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
If any of them are not installed you can still follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:')
```
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
    The names of the packages used (and their version numbers available on this platform) are:')
```

* `mice` (version: `r packageVersion("mice")`)

### Help files
You can find help files for any function by adding a `?` before the name of the 
function.
```{r, eval = !static, results = 'asis', echo = FALSE}
cat("The files might look a bit funny when they are displayed as R output.")
```

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


### Dataset 
For this practical, we will again use the **NHANES** dataset that we have seen in
the previous practicals.
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat("In this interactive version of this practical, the data is already loaded
    and the variable `educ` correctly coded as an ordered factor.")
```
```{r, eval = static, echo = FALSE, results = 'asis'}
cat(
'To load this dataset, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the file
`NHANES_for_practicals.RData` on your computer.
If you know the path to the file, you can also use `load("<path>/NHANES_for_practicals.RData")`.

If you have not followed the first practical or if you re-loaded the NHANES data
you need to re-code the variable `educ` again:'
)
```
```{r, eval = FALSE, echo = static}
NHANES$educ <- as.ordered(NHANES$educ)
```

### Imputed data
The imputed data are stored in a `mids` object called `imp` that we created in
the previous practical.
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat("`imp` is already loaded in this interactive version of this practical.")
```
```{r, eval = static, echo = FALSE, results = 'asis'}
cat(
'You can load it into your workspace by clicking the object `imps.RData` 
if you are using RStudio.
Alternatively, you can load this workspace using `load("<path>/imps.RData")`.
You then need to run:')
```
```{r, eval = FALSE, echo = static}
imp <- savedimps_imp
```


## Analysing mulitple imputed datasets
### Fitting models on `mids` objects {.tabset .tabset-fade .tabset-pills}
When we are confident that the imputation was successfull, the imputed data can
be analysed with any standard complete data method. 

Depending on the research question, for the NHANES data you might use

* linear regression using `lm()`
* logistic regression using `glm()` with `family = binomial()`
* other generalized linear regression models using `glm()` with an appropriate `family`

Since our imputed data `imp` is not a `data.frame` but a `mids` object, we cannot 
use the standard specification
`lm(formula = ..., data = imp, ...)`
but need to use the function `with()` from the **mice** package:

`with(imp, lm(formula, ...))`


```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

* Choose any analysis model of interest and fit it on the imputed data. 
* Find out the `class` and structure of the resulting object.

```{r analysisstep_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#analysisstep">Hint</button>
<div id = "analysisstep" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
For example:\n
`mod1 <- with(imp, lm(SBP ~ age + gender + bili + chol + BMI))`\n
`mod2 <- with(imp, glm(hypchol ~ age + gender + bili + race + BMI, family = "binomial"))`\n
`mod3 <- with(imp, glm(creat ~ age + gender + uricacid + WC + BMI, family = Gamma(link = "log")))`\n
</div>')
```

```{r analysisstep, exercise = TRUE, exercise.setup = "allimp"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r analysisstep-hint-1}
# You may, for example, choose one of these models:
mod1 <- with(imp, lm(SBP ~ age + gender + bili + chol + BMI))
mod2 <- with(imp, glm(hypchol ~ age + gender + bili + race + BMI,
                       family = "binomial"))
mod3 <- with(imp, glm(creat ~ age + gender + uricacid + WC + BMI,
                       family = Gamma(link = 'log')))
```

```{r analysisstep-hint-2}
# to determine the class
class(mod1)

# to determine the structure
names(mod1)
lapply(mod1, class)

# explore further
mod1$analyses
summary(mod1$analyses[[1]])
```


### Structure of `mira` objects
Analyses performed on `mids` objects will result in a `mira` object
(multiply imputed repeated analyses).


A `mira` object is a list with the following elements:
```{r, echo = FALSE}
miradf <- rbind(
  c("`call`:", "The call that created the `mira` object."),
  c("`call1`:", "The call that created the `mids` object."),
  c("`nmis`:", "The number of missing observations per column."),
  c("`analyses`:",
    paste0("The models fitted on each of the imputed datasets: ",
           "A list with `imp$m` components, where each component contains
           a fitted model object."))
) %>% as.data.frame

names(miradf) <- c(" ", " ")

miradf %>% kable(format = 'html') %>%
  kable_styling()
```

The class of each of the components of `imp$analyses` will depend on the
type of model that was fitted.

If you want to check the results from one particular imputed dataset, e.g., the
3rd, you can access the fitted model via `mod1$analyses[[3]]`. 
Alternatively you can use the function `getfit()`.

## Pooling 
```{r, eval = static, echo = FALSE}
asis_output('### Combining results {.tabset .tabset-fade .tabset-pills}')
```

To pool the results from the repeated analyses contained in a `mira` object,
the function `pool()` can be used.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Pool one of the provided `mira` objects (`mod1`, `mod2` or `mod3`)
and investigate its structure.

```{r impmods, echo = FALSE}
impnaive <- savedimps_impnaive
imp <- savedimps_imp

mod1 <- with(imp, lm(SBP ~ age + gender + bili + chol + BMI))
mod2 <- with(imp, glm(hypchol ~ age + gender + bili + race + BMI, 
                       family = "binomial"))
mod3 <- with(imp, glm(creat ~ age + gender + uricacid + WC + BMI,
                       family = Gamma(link = 'log')))
```

```{r pooling, exercise = TRUE, exercise.setup = "impmods"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```


```{r pooling-solution, solution = TRUE}
# pool the model
pooled1 <- pool(mod1)

# investigate the structure
class(pooled1)
names(pooled1)
str(pooled1)
```


### Structure of `mipo` objects {.tabset .tabset-fade .tabset-pills}
Pooling of a `mira` object results in a `mipo` object (multiply imputed pooled analysis).

The `mipo` object is a list that has the following elements:
```{r, echo = FALSE}
pooldf <- rbind(
  c("`call`:", "The call that created the `mipo` object."),
  # c("`call1`:", "The call that created the `mids` object."),
  # c("`call2`:", "The call that created the `mira` object."),
  # c("`data`", "The original, incomplete data."),
  # c("`nmis`:", "The number of missing values per variable."),
  c("`m`:", "The number of imputed datasets."),
  c("`pooled`:", "A `data.frame` containing pooled estimates and variances.")
  # c("`qhat`:", "The coefficients from all analyses: a matrix with `m` rows and `npar`columns."),
  # c("`u`:", paste0("The variance-covariance matrix of the coefficients from each analysis: ",
  #                  "an array of dimension `c(m, npar, npar)`.")),
  # c("`qbar`:", "The pooled parameter estimates: a vector of length `npar`."),
  # c("`ubar`:", paste0("The average within imputation variance: ",
  #                     "a matrix with `npar` rows and `npar` columns containing ",
  #                     "the averaged variance-covariance matrices.")),
  # c("`b`:", "The between imputation variance-covariance matrix ($u + b + b/m$)."),
  # c("`t`:", "The total variance-covariance matrix."),
  # c("`r`:", "The relative increase in variance due to the missing values."),
  # c("`dfcom`:", "Degrees of freedom in the hypothetically complete data (`N - # free parameters`)."),
  # c("`df`:", "Degrees of freedom associated with the $t$-statistics."),
  # c("`fmi`:", "Fraction of missing information."),
  # c("`lambda`:", "Proportion of the variation due to the missing values: $(b+b/m)/t$.")
) %>% as.data.frame

names(pooldf) <- c(" ", " ")

pooldf %>% kable(format = 'html') %>%
  kable_styling(full_width = FALSE)
```


`pooled` has the following columns:
```{r, echo = FALSE}
tab <- rbind(
  c("`estimate`:", "Pooled parameter estimates"),
  c("`ubar`:", "The within-imputation variance of the `estimate`"),
  c("`b`:", "The between-imputation variance of the `estimate`"),
  c("`t`:", "The total variance of the `estimate`"),
  c("`dfcom`:", "Degrees of freedom in the complete data"),
  c("`df`:", "Degrees of freedom associated with the $t$-statistics."),
  c("`riv`:", "The relative increase in variance due to the missing values: $(b+b/m)/ubar$"),
  c("`lambda`:", "Proportion of the variation due to the missing values: $(b+b/m)/t$."),
  c("`fmi`:", "Fraction of missing information.")
)

names(tab) <- c(" ", " ")

tab %>% kable(format = 'html') %>%
  kable_styling(full_width = FALSE)
```

From the elements contained in the `mipo` object, we could calculate the pooled
confidence intervals and p-values by hand, using the function provided in 
the course notes.

However, the function `summary()` applied to the `mipo` object will do this for us.
This function has four more arguments that may be relevant:

* `conf.int`: needs to be set to `TRUE` in order to to get confidence intervals
* `conf.level`: the confidence level, by default this is 0.95
* `exponentiate`: whether to exponentiate the pooled estimates and confidence
  intervals (e.g., for logistic regression)

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

```{r, eval = static, echo = FALSE}
asis_output("Get the summary of the pooled results.")
```

```{r summarypooled, exercise = TRUE, exercise.setup = "impmods"}
pooled1 <- pool(mod1)
...
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```

```{r, summarypooled-solution, solution = TRUE}
pooled1 <- pool(mod1)
summary(pooled1, conf.int = TRUE)
```


<div style="border:1px; border-style:solid; padding: 1em; border-color:#D11450">
**Note:**<br>
`pool()` extracts the model coefficients and variance-covariance matrices using
`tidy` from the package **broom**. 
Hence, pooling using the `pool()` function from **mice** only works for models
of classes for which a method `tidy` exists.
</div>
