---
title: "Analysing imputed data"
output: 
  html_document:
    code_folding: show
    df_print: paged
    toc: true
    toc_float:
      collapsed: false
    number_sections: false
    theme: spacelab
    highlight: tango
    includes:
      after_body: ../footer.html
    css: ../style.css
    pandoc_args: [
      "--output=Practical_AnalyseMIData.html"
      ]
#   learnr::tutorial:
#     progressive: true
#     allow_skip: true
# runtime: shiny_prerendered
---

```{r, include = F}
if (names(rmarkdown::metadata$output) == "learnr::tutorial") {
  library(learnr)
  learnr::initialize_tutorial()
  static <- FALSE
}

if (names(rmarkdown::metadata$output) == "html_document") {
  knitr::opts_hooks$set(eval = function(opt) {
    if (any(opt$exercise))
      opt$eval <- opt$include <- FALSE
    
    opt
  })
  
  static <- TRUE
  
  options(width = 100)
}
```


```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
library(JointAI)
library(reshape2)
library(plyr)
library(RColorBrewer)
library(ggplot2)
```


```{r load_data, context="data", include=FALSE}
load("www/NHANES_for_practicals.RData")
NHANES$educ <- as.ordered(NHANES$educ)

source("www/propplot.R")
load("www/imps.RData")

imp <- savedimps_imp

library(mice)
```


## Preface {data-progressive=FALSE}
### R packages

```{r, eval = static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
If any of them are not installed you can still follow the practical
but will not be able to run all of the code. The packages used (with versions
that were used to generate the solutions) are:')
```
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat('In this practical, a number of R packages are used.
    The names of the packages used (and their version numbers available on this platform) are:')
```

* `mice` (version: `r packageVersion("mice")`)
* `JointAI` (version: `r packageVersion("JointAI")`)
* `reshape2` (version: `r packageVersion("reshape2")`)
* `RColorBrewer` (version: `r packageVersion("RColorBrewer")`)
* `ggplot2` (version: `r packageVersion("RColorBrewer")`)


### Help files
You can find help files for any function by adding a `?` before the name of the 
function.

```{r, eval = !static, results = 'asis', echo = FALSE}
cat("The files might look a bit funny when they are displayed as R output.")
```

Alternatively, you can look up the help pages online at 
[https://www.rdocumentation.org/](https://www.rdocumentation.org/)
or find the whole manual for a package at
[https://cran.r-project.org/web/packages/available_packages_by_name.html](https://cran.r-project.org/web/packages/available_packages_by_name.html)


### Dataset 

For this practical, we will again use the **NHANES** dataset that we have seen in
the previous practical.
```{r, eval = !static, echo = FALSE, results = 'asis'}
cat("In the interactive version of this practical, the data is already loaded
    and the variable `educ` correctly coded as an ordered factor.")
```
```{r, eval = static, echo = FALSE, results = 'asis'}
cat(
'To load this dataset, you can use the command `file.choose()` which opens the
explorer and allows you to navigate to the location of the file
`NHANES_for_practicals.RData` on your computer.
If you know the path to the file, you can also use `load("<path>/NHANES_for_practicals.RData")`.

If you have not followed the first practical or if you re-loaded the NHANES data
you need to re-code the variable `educ` again:'
)
```
```{r, eval = FALSE, echo = static}
NHANES$educ <- as.ordered(NHANES$educ)
```



## Fitting models on `mids` objects {.tabset .tabset-fade .tabset-pills}
When we are confident that the imputation was successfull, the imputed data can
be analysed with any complete data method. 

```{r, eval = static, echo = FALSE}
asis_output("### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Choose any analysis model of interest and fit it on the imputed data. 
The model could for instance be a linear regression model (using `lm`) or a 
generalized linear regression model, e.g., logistic regression
(using `glm()`).
Find out the `class` and structure of the resulting object.

```{r analysisstep_hint, eval = static, results = 'asis', echo = FALSE}
cat('<button type="button" class="btn btn-info btn-sm" data-toggle="collapse" data-target="#analysisstep">Hint</button>
<div id = "analysisstep" class="collapse" style="border:1px; border-style:solid; padding: 1em; border-color:#1F78B4">
`imp` is not a dataframe but an object of class `mids`, so we cannot use the
normal specification of a model in which we specify the argument `data`
to be a `data.frame`. Instead, use `with(imp, <model function without data>)`.
</div>')
```

```{r analysisstep, exercise = TRUE, exercise.setup = "allimp"}

```

```{r analysisstep-hint-1, include = !static}
# "imp" is not a dataframe but an object of class "mids", so we cannot use the
# normal specification of a model in which we specify the argument "data"
# to be a "data.frame". Instead, use "with(imp, <model function without data>)".

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution\\n")
```

```{r analysisstep-hint-2}
# for example:
models1 <- with(imp, lm(SBP ~ age + gender + bili + chol + BMI))
models2 <- with(imp, glm(hypchol ~ age + gender + bili + race + BMI,
                       family = "binomial"))
models3 <- with(imp, glm(creat ~ age + gender + uricacid + WC + BMI,
                       family = Gamma(link = 'log')))
```

```{r analysisstep-hint-3}
# to determine the class
class(models1)

# to determine the structure
names(models1)
lapply(models1, class)

# explore further
models1$analyses
summary(models1$analyses[[1]])
```



## Structure of `mira` objects {.tabset .tabset-fade .tabset-pills}
Analyses performed on `mids` objects will result in a `mira` object
(multiply imputed repeated analyses).


A `mira` object is a list with the following elements:
```{r, echo = FALSE}
miradf <- rbind(
  c("`call`:", "The call that created the `mira` object."),
  c("`call1`:", "The call that created the `mids` object."),
  c("`nmis`:", "The number of missing observations per column."),
  c("`analyses`:",
    paste0("The models fitted on each of the imputed datasets: ",
           "A list with `imp$m` components, where each component contains
           a fitted model object."))
) %>% as.data.frame

names(miradf) <- c(" ", " ")

miradf %>% kable(format = 'html') %>%
  kable_styling()
```

The class of each of the components of `imp$analyses` will depend on the
type of model that was fitted.

## Pooling 
```{r, eval = static, echo = FALSE}
asis_output('### Combining results {.tabset .tabset-fade .tabset-pills}')
```

To pool the results from the repeated analyses contained in a `mira` object,
the function `pool()` can be used.

```{r, eval = static, echo = FALSE}
asis_output("#### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

Pool one of the provided `mira` objects (`models1`, `models2` or `models3`)
and investigate its structure.

```{r impmods, echo = FALSE}
NHANES$educ <- as.ordered(NHANES$educ)
imp0 <- mice(NHANES, maxit = 0, 
             defaultMethod = c("norm", 'logreg', 'polyreg', 'polr'))
meth <- imp0$meth
meth["creat"] <- "pmm"
meth["HyperMed"] <- ""

pred <- imp0$pred

pred[c("hgt", "wgt"), "BMI"] <- 0
pred[, c("hgt", "wgt")] <- 0
pred["hgt", "wgt"] <- 1
pred["wgt", "hgt"] <- 1

pred[, "HyperMed"] <- 0

pred["chol", "hypchol"] <- 0

meth["BMI"] <- "~I(wgt/hgt^2)"
meth["HyperMed"] <- ""

visSeq <- imp0$visitSequence
which_BMI <- match("BMI", visSeq)
visSeq <- c(visSeq[-which_BMI], visSeq[which_BMI])

impnaive <- savedimps_impnaive
imp <- savedimps_imp

models1 <- with(imp, lm(SBP ~ age + gender + bili + chol + BMI))
models2 <- with(imp, glm(hypchol ~ age + gender + bili + race + BMI, 
                       family = "binomial"))
models3 <- with(imp, glm(creat ~ age + gender + uricacid + WC + BMI,
                       family = Gamma(link = 'log')))
```

```{r pooling, exercise = TRUE, exercise.setup = "impmods"}

```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("#### Solution\\n")
```


```{r pooling-solution, solution = TRUE}
# pool the model
pooled1 <- pool(models1)

# investigate the structure
class(pooled1)
names(pooled1)
str(pooled1)
```


## Structure of `mipo` objects {.tabset .tabset-fade .tabset-pills}
Pooling of a `mira` object results in a `mipo` object (multiply imputed pooled analysis).

The `mipo` object is a list that has the following elements:
```{r, echo = FALSE}
pooldf <- rbind(
  c("`call`:", "The call that created the `mipo` object."),
  c("`call1`:", "The call that created the `mids` object."),
  c("`call2`:", "The call that created the `mira` object."),
  c("`data`", "The original, incomplete data."),
  c("`nmis`:", "The number of missing values per variable."),
  c("`m`:", "The number of imputations."),
  c("`qhat`:", "The coefficients from all analyses: a matrix with `m` rows and `npar`columns."),
  c("`u`:", paste0("The variance-covariance matrix of the coefficients from each analysis: ",
                   "an array of dimension `c(m, npar, npar)`.")),
  c("`qbar`:", "The pooled parameter estimates: a vector of length `npar`."),
  c("`ubar`:", paste0("The average within imputation variance: ",
                      "a matrix with `npar` rows and `npar` columns containing ",
                      "the averaged variance-covariance matrices.")),
  c("`b`:", "The between imputation variance-covariance matrix ($u + b + b/m$)."),
  c("`t`:", "The total variance-covariance matrix."),
  c("`r`:", "The relative increase in variance due to the missing values."),
  c("`dfcom`:", "Degrees of freedom in the hypothetically complete data (`N - # free parameters`)."),
  c("`df`:", "Degrees of freedom associated with the $t$-statistics."),
  c("`fmi`:", "Fraction of missing information."),
  c("`lambda`:", "Proportion of the variation due to the missing values: $(b+b/m)/t$.")
) %>% as.data.frame

names(pooldf) <- c(" ", " ")

pooldf %>% kable(format = 'html') %>%
  kable_styling()
```

From the elements contained in the `mipo` object, we could obtain the pooled
coefficients (`qbar`) and variances (`t`),
and calculate the pooled confidence intervals and p-values by hand, using the
function provided in the course notes.

However, the function `summary()` applied to the `mipo` object will do this for us.

<div style="border:1px; border-style:solid; padding: 1em; border-color:#D11450">
**Note:**<br>
`pool()` extracts the model coefficients and variance-covariance matrices using
the functions `coef()` and `vcov()`. Hence, pooling using the `pool()` 
function from **mice** only works for those types of models for which these functions
will return the regression coefficients and corresponding variance-covariance matrix.

Moreover, `pool()` only works for objects of class `mira`. When data has been
imputed in a different package/software, or analyses have not been performed
in the way described above, a list of fitted models can be converted to a `mira`
object using `as.mira()`.
</div>


```{r, eval = static, echo = FALSE}
asis_output("### Task\\n")
```

```{r, eval = static, echo = FALSE}
asis_output('<div style="border:2px; border-style:solid; padding: 1em; border-color:#446E9B">')
```

```{r, eval = static, echo = FALSE}
asis_output("Get the summary of the pooled results.")
```

```{r summarypooled, exercise = TRUE, exercise.setup = "impmods"}
pooled1 <- pool(models1)
...
```

```{r, eval = static, echo = FALSE}
asis_output("</div>")

asis_output("### Solution\\n")
```

```{r, summarypooled-solution, solution = TRUE}
pooled1 <- pool(models1)
summary(pooled1)
```
