---
title: "\\textbf{Assignment II for\\\\[1ex]EP16: Missing Values in Clinical Research}"
subtitle: "\\LARGE\\textbf{Multiple Imputation}"
date: "\\textbf{13 -- 17 May, 2019}"
output:
  pdf_document:
    includes:
      in_header: header.tex
geometry: margin = 2.2cm
params:
  year: NA
  theseed: NA
---

```{r, echo = FALSE}
datatype <- 3
year <- params$year
theseed <- params$theseed

load(paste0('../data/MIdat', datatype, theseed, '_', year, '.RData'))
datname <- paste0('MIdat', datatype, theseed)
dat <- get(datname)

```

# Data {-}
The **`r datname`** data comprise `r nrow(dat)` observations of `r length(unique(dat$id))` mothers measured repeatedly
before and during pregnancy. All women were scheduled to have their weight measured
once each trimester and were asked for their pre-pregnancy weight and BMI.

The dataset contains the following variables:

```{r, echo = FALSE, results = 'asis'}

vars <- c(
  "id" = "subject identifier",
  "gage" = "gestational age at measurement (\\texttt{gage = 0} refers to a measurement before pregnancy)",
  "weight" = "maternal weight",
  "gestbir" = "gestational age at birth",
  "kcal" = "average daily kcal intake (calculated from food frequency questionnaire)",
  "BMI" = "self reported maternal BMI before pregnancy",
  "bmi" = "self reported maternal BMI before pregnancy",
  "date_incl" = "date of inclusion in the study",
  "bd_mom" = "birth date of the mother",
  "preterm" = "was the baby born before 37 weeks of gestation (preterm) or later?",
  "smoke" = "smoking behaviour of the mother during pregnancy",
  "stress" = "self reported stress score (0 -- 5)",
  "sex" = "child sex",
  "gender" = "child gender",
  "parity" = "number of pregnancies of more than 20 weeks the mother had (nulliparity: this was the first pregnancy, >=1 child: mother had previous pregnancies)",
  "educ" = "educational level of the mother",
  "visit_center" = "was intake performed at the study center? (0: no, 1: yes)",
  "income" = "household income",
  "trimester" = "trimester of measurement",
  "inclusion" = "date of inclusion in the study",
  "alc" = "alcohol consumption behaviour of the mother during pregnancy"
)


# 
# if (theseed == 1) {
#   vars['birthwgt'] <- 'birthweight (SD-score)'
#   vars['length'] <- "child's length at time of DXA scan in centimeters"
# }
# 
# if (theseed == 3) {
#   vars['sun'] <- 'average sun light duration in hours/day in the month before blood sampling'
#   vars['length'] <- "child's length at time of DXA scan in centimeters"
# }

if (any(!names(dat) %in% names(vars))) {
  n <- names(dat)[which(!names(dat) %in% names(vars))]
  print(paste0("\\textcolor{red}{", n, "not described!!!}"))
}

vars <- vars[match(names(dat), names(vars))]

names(vars) <- gsub('_', '\\_', names(vars), fixed = TRUE)

cat(
  paste0('\\begin{table}[!ht]
          \\centering
          \\begin{tabular}{lp{13cm}}\\toprule
         variable & explanation\\\\\\midrule',
  paste0("\\texttt{", names(vars), "} & ", vars, collapse = '\\\\\n'),
  '\\\\\\bottomrule
\\end{tabular}
\\end{table}'
))
```

```{r, echo = FALSE}
out <- 'weight'
timevar <- 'gage'
covars <- c("smoke", "kcal", "stress", "preterm", "parity", "educ", "income")

# if (theseed == 1) {
#   covars <- gsub('ancestry', 'ethn', covars)
#   covars <- gsub('gender', 'sex', covars)
# }
# 
if (theseed == 3) {
  covars <- gsub('smoke', 'alc', covars)
}

covstring <- paste(paste0(paste0('`', covars[-length(covars)], '`'), collapse = ', '),
                   'and', paste0('`', covars[length(covars)], '`'))

if (any(!covars %in% names(dat))) {
  n <- covars[!covars %in% names(dat)]
  print(paste0("\\textcolor{red}{", n, "not in data!!!}"))
}

```


# Analysis model of interest {-}
The analysis model of interest is a linear mixed model for `r out` 
with random intercept and slope for `r timevar`.
Covariates are `r covstring`.

```{r, eval = FALSE, echo = FALSE}
for(theseed in 1:3) {
  rmarkdown::render('Assignment/instructions/template03.Rmd',
                    output_file = paste0('MIdat3', theseed, '_2019_documentation.pdf'), 
                    params = list(year = 2019, theseed = theseed))
  
  file.remove(paste0("Assignment/instructions/MIdat3", theseed, "_2019_documentation.tex"))
}

file.show("Assignment/instructions/MIdat31_2019_documentation.pdf")
file.show("Assignment/instructions/MIdat32_2019_documentation.pdf")
file.show("Assignment/instructions/MIdat33_2019_documentation.pdf")


load('Assignment/data/MIdat31_2019.RData')
load('Assignment/data/MIdat32_2019.RData')
load('Assignment/data/MIdat33_2019.RData')


summary(MIdat31[, intersect(names(MIdat31), names(MIdat32))])
summary(MIdat32[, intersect(names(MIdat31), names(MIdat32))])

summary(MIdat31[, intersect(names(MIdat31), names(MIdat33))])
summary(MIdat33[, intersect(names(MIdat31), names(MIdat33))])

summary(MIdat32[, intersect(names(MIdat32), names(MIdat33))])
summary(MIdat33[, intersect(names(MIdat32), names(MIdat33))])
```

