---
title: "\\textbf{Assignment II for\\\\[1ex]EP16: Missing Values in Clinical Research}"
subtitle: "\\LARGE\\textbf{Multiple Imputation}"
date: "\\textbf{13 -- 17 May, 2019}"
output:
  pdf_document:
    includes:
      in_header: header.tex
geometry: margin = 2.2cm
params:
  year: NA
  theseed: NA
---

```{r, echo = FALSE}
datatype <- 4
year <- params$year
theseed <- params$theseed

load(paste0('../data/MIdat', datatype, theseed, '_', year, '.RData'))
datname <- paste0('MIdat', datatype, theseed)
dat <- get(datname)

```

# Data {-}
The **`r datname`** data comprise information of `r nrow(dat)` participants of an observational
study on the effects of testosterone on heart rate.

The dataset contains the following variables:

```{r, echo = FALSE, results = 'asis'}

vars <- c(
"ID" = "subject identifier",
"Gender" = "participant's gender",
"Income" = "Household income",
"HR" = "heart rate",
"sleep" = "hours of sleep per night",
"PA" = "number of days per week on which physical activity is performed",
"alc" = "number of alcoholic drinks consumed per week",
"smoke" = "smoking status",
"PC" = "number of hours spent using a computer per day",
"TV" = "number of hours spent watching television per day",
"height" = "participant's height in cm",
"weight" = "participant's weight in kg",
"BMI" = "participant's BMI",
"Age" = "participant's Age",
"Testosterone" = "participant's testosterone level in ng/dL",
"PhysActive" = "Does the participant perform physical activity regularily?",
"Pregnant" = 'Was the participant pregnant when the measurements were taken?',
"diab" = 'Does the participant have diabetes?'
)


# 
# if (theseed == 1) {
#   vars['birthwgt'] <- 'birthweight (SD-score)'
#   vars['length'] <- "child's length at time of DXA scan in centimeters"
# }
# 
# if (theseed == 3) {
#   vars['sun'] <- 'average sun light duration in hours/day in the month before blood sampling'
#   vars['length'] <- "child's length at time of DXA scan in centimeters"
# }

if (any(!names(dat) %in% names(vars))) {
  n <- names(dat)[which(!names(dat) %in% names(vars))]
  print(paste0("\\textcolor{red}{", n, "not described!!!}"))
}

vars <- vars[match(names(dat), names(vars))]

names(vars) <- gsub('_', '\\_', names(vars), fixed = TRUE)

cat(
  paste0('\\begin{table}[!ht]
          \\centering
          \\begin{tabular}{lp{13cm}}\\toprule
         variable & explanation\\\\\\midrule',
  paste0("\\texttt{", names(vars), "} & ", vars, collapse = '\\\\\n'),
  '\\\\\\bottomrule
\\end{tabular}
\\end{table}'
))
```

```{r, echo = FALSE}
out <- 'HR'
covars <- c("Testosterone", "Gender", "Age", "smoke", "sleep", "PA", "PC", "BMI", "alc")

# if (theseed == 1) {
#   covars <- gsub('ancestry', 'ethn', covars)
#   covars <- gsub('gender', 'sex', covars)
# }
# 
# if (theseed == 3) {
#   covars <- gsub('smoke', 'alc', covars)
# }

covstring <- paste(paste0(paste0('`', covars[-length(covars)], '`'), collapse = ', '),
                   'and', paste0('`', covars[length(covars)], '`'))

if (any(!covars %in% names(dat))) {
  n <- covars[!covars %in% names(dat)]
  print(paste0("\\textcolor{red}{", n, "not in data!!!}"))
}

```


# Analysis model of interest {-}
The analysis model of interest is a linear regression with outcome `r out` 
and covariates `r covstring`.

We assume that there are interaction effects between `Testosterone`
and `Gender` as well as between `Age` and `smoke`.


```{r, eval = FALSE, echo = FALSE}
for(theseed in 1:3) {
  rmarkdown::render('Assignment/instructions/template04.Rmd',
                    output_file = paste0('MIdat4', theseed, '_2019_documentation.pdf'), 
                    params = list(year = 2019, theseed = theseed))
  
  file.remove(paste0("Assignment/instructions/MIdat4", theseed, "_2019_documentation.tex"))
}

file.show("Assignment/instructions/MIdat41_2019_documentation.pdf")
file.show("Assignment/instructions/MIdat42_2019_documentation.pdf")
file.show("Assignment/instructions/MIdat43_2019_documentation.pdf")


load('Assignment/data/MIdat41_2019.RData')
load('Assignment/data/MIdat42_2019.RData')
load('Assignment/data/MIdat43_2019.RData')


summary(MIdat41[, intersect(names(MIdat41), names(MIdat42))])
summary(MIdat42[, intersect(names(MIdat41), names(MIdat42))])

summary(MIdat41[, intersect(names(MIdat41), names(MIdat43))])
summary(MIdat43[, intersect(names(MIdat41), names(MIdat43))])


summary(MIdat42[, intersect(names(MIdat42), names(MIdat43))])
summary(MIdat43[, intersect(names(MIdat42), names(MIdat43))])
```

