---
title: "\\textbf{Assignment II for\\\\[1ex]EP16: Missing Values in Clinical Research}"
subtitle: "\\LARGE\\textbf{Multiple Imputation}"
date: "\\textbf{13 -- 17 May, 2019}"
output:
  pdf_document:
    includes:
      in_header: header.tex
params:
  year: NA
  theseed: NA
---

```{r, echo = FALSE}
datatype <- 1
year <- params$year
theseed <- params$theseed

load(paste0('../data/MIdat', datatype, theseed, '_', year, '.RData'))
datname <- paste0('MIdat', datatype, theseed)
dat <- get(datname)

```

# Data {-}
The **`r datname`** data comprise data of `r nrow(dat)` children and
their mothers on vitamin D exposure of the mother during pregnancy and child
bone health, measured by DXA scan, at 6 years of age. Maternal
serum samples were taken in the third trimester of pregnancy.

The dataset contains the following variables:

```{r, echo = FALSE, results = 'asis'}
vars <- c("ID" = "subject identifier",
          "gender" = "gender of the child",
          "sex" = "sex of the child",
          "sun_birth" = "average sun light duration in the month before birth in hours/day",
          "bdate" = "child'd birth date",
          "birth" = "date of birth",
          "lean" = "lean mass in kg",
          "length" = "child's length at time of DXA scan in meters",
          "ancestry" = "child's ancestry",
          "ethn" = "child's ethnicity",
          "singleton" = "is the child a singleton birth?",
          "sun" = "average sun light duration in minutes/day in the month before blood sampling",
          "birthwgt" = "birthweight in kg",
          "BMC" = "bone mineral content of the child in grams, determined by DXA scan",
          "weight" = "child's total weight in kg at DXA scan",
          "gravidity" = "number of times the mother has been pregnant
         (primigravida: this was the first pregnancy,
          multigravida: the mother had previous pregnancies)",
         "leanfrac" = "proportion of child's lean mass (lean mass/total mass; lean mass = total mass - fat mass)",
         "vitD" = "mother's serum vitamin D concentrations in 10 nmol/L",
         "season" = "season of blood sampling",
         "sports" = "does the child do sports regularily?",
         "parity" = "number of pregnancies of more than 20 weeks the mother had (nulliparity: this was
the first pregnancy, multiparity: mother had previous pregnancies)"
)


if (theseed == 1) {
  vars['birthwgt'] <- 'birthweight (SD-score)'
  vars['length'] <- "child's length at time of DXA scan in centimeters"
}

if (theseed == 3) {
  vars['sun'] <- 'average sun light duration in hours/day in the month before blood sampling'
  vars['length'] <- "child's length at time of DXA scan in centimeters"
}

if (any(!names(dat) %in% names(vars))) {
  n <- names(dat)[which(!names(dat) %in% names(vars))]
  print(paste0("\\textcolor{red}{", n, "not described!!!}"))
}

vars <- vars[match(names(dat), names(vars))]

names(vars) <- gsub('_', '\\_', names(vars), fixed = TRUE)

cat(
  paste0('\\begin{table}[!ht]
          \\centering
          \\begin{tabular}{lp{13cm}}\\toprule
         variable & explanation\\\\\\midrule',
  paste0("\\texttt{", names(vars), "} & ", vars, collapse = '\\\\\n'),
  '\\\\\\bottomrule
\\end{tabular}
\\end{table}'
))
```

```{r, echo = FALSE}
out <- 'BMC'
covars <- c('vitD', 'ancestry', 'gender', 'leanfrac', 'sports', 'sun', 'season', 'length', 'weight')

if (theseed == 1) {
  covars <- gsub('ancestry', 'ethn', covars)
  covars <- gsub('gender', 'sex', covars)
}

if (theseed == 3) {
  covars <- gsub('ancestry', 'ethn', covars)
}

covstring <- paste(paste0(paste0('`', covars[-length(covars)], '`'), collapse = ', '),
                   'and', paste0('`', covars[length(covars)], '`'))

if (any(!covars %in% names(dat))) {
  n <- covars[!covars %in% names(dat)]
  print(paste0("\\textcolor{red}{", n, "not in data!!!}"))
}

```


# Analysis model of interest {-}
The analysis model of interest is a linear regression with outcome `r out`,
and covariates `r covstring`.


We assume that `vitD` has a non-linear (quadratic) effect.

```{r, eval = FALSE, echo = FALSE}
for(theseed in 1:3) {
  rmarkdown::render('Assignment/instructions/template01.Rmd',
                    output_file = paste0('MIdat1', theseed, '_2019_documentation.pdf'), 
                    params = list(year = 2019, theseed = theseed))
  
  file.remove(paste0("Assignment/instructions/MIdat1", theseed, "_2019_documentation.tex"))
}

file.show("Assignment/instructions/MIdat11_2019_documentation.pdf")
file.show("Assignment/instructions/MIdat12_2019_documentation.pdf")
file.show("Assignment/instructions/MIdat13_2019_documentation.pdf")


load('Assignment/data/MIdat11_2019.RData')
load('Assignment/data/MIdat12_2019.RData')
load('Assignment/data/MIdat13_2019.RData')


summary(MIdat11[, intersect(names(MIdat11), names(MIdat12))])
summary(MIdat12[, intersect(names(MIdat11), names(MIdat12))])

summary(MIdat11[, intersect(names(MIdat11), names(MIdat13))])
summary(MIdat13[, intersect(names(MIdat11), names(MIdat13))])


summary(MIdat12[, intersect(names(MIdat12), names(MIdat13))])
summary(MIdat13[, intersect(names(MIdat12), names(MIdat13))])
```

